\documentclass[output=paper]{langsci/langscibook} 
\author{Bruno Herin\affiliation{Inalco, IUF}}
\title{Northern Domari}
\abstract{This chapter provides an overview of the linguistic outcomes of contact between Arabic and Northern Domari. Northern Domari is a group of dialects spoken in Syria, Lebanon, Jordan and Turkey. It remained until very recently largely unexplored. This article presents unpublished first-hand linguistic data collected in Lebanon, Syria, Jordan and Turkey. It focuses on the Beirut/Damascus variety, with references to the dialects spoken in northern Syria and southern Turkey.}
\IfFileExists{../localcommands.tex}{
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle
 
\section{Current state and historical development}

Domari is an Indic language spoken by the Doms in various countries of the Middle East. The Doms are historically itinerant communities who specialize in service economies. This occupational profile led the lay public to call them the Middle Eastern Gypsies. Common occupations are informal dentistry, metalwork, instrument crafting, entertainment and begging. Most claim Sunni Islam as their religion, with various degrees of syncretic practices. Although most have given up their semi-nomadic lifestyle and settled in the periphery of urban centres, mobility is still a salient element in the daily lives of many Doms.

The ethnonym Dom is mostly unknown to non-Doms, who refer to them with various appellations such as \textit{nawar,} \textit{qurbāṭ} or \textit{qarač}. The Standard Arabic word \textit{ɣa\v{g}ar} for `Gypsy' is variably accepted by the Doms, who mostly understand with this term European Gypsies. All these appellations are exonyms and the only endonym found across all communities is \textit{dōm}. Only the Gypsies of Egypt, it seems, use a reflex of \textit{ɣa\v{g}ar} to refer to themselves. 

From the nineteenth century onwards, European travellers reported the existence of Domari in the shape of word lists collected in the Caucasus, Iran, Iraq and the Levant (see \citealt{Herin2012} for a discussion of these sources). The first full-length grammatical description of a dialect of Domari is by \citet{Macalister1914}, who described the dialect spoken in Palestine in the first years of the twentieth century. At present, the language is known to be spoken in Palestine, Jordan, Lebanon, Syria and Turkey. No recent account can confirm that it is still spoken in Iraq and Iran. There are roughly two dialectal areas:  Southern Domari, spoken in Palestine and Jordan, and Northern Domari, spoken in Lebanon, Syria and southern Turkey. This geographical division is not clear cut, as I have recorded speakers of Southern varieties in Lebanon and speakers of Northern dialects in Jordan. The main isogloss separating these two groups is the maintenance of a two-way gender system. Southern dialects have maintained the gender distinction, whereas it has mostly disappeared in the north. Compare Northern \textit{gara} ‘(s)he went’ vs. Southern \textit{gara} ‘he went’ and \textit{garī} ‘she went’. These are sufficiently different to allow us to posit an early split. Mutual intelligibility appears to be very limited. A case in point is kinship terminology, which is largely divergent in both groups. Within Northern Domari, the Beirut/Damascus dialect stands out because of the glottal realization [ʔ] of etymological /q/ and the loss of the differential subject marker \textit{-ən}. 

No general statement can be made about language endangerment. Jerusalem Domari is reported to have only one fluent speaker left (Matras, this volume),\ia{Matras, Yaron@Matras, Yaron} but the presence of speakers of Palestinian Domari in other places may not be excluded. Young fluent speakers of Southern dialects are easy to find in Jordan. As far as Northern Domari is concerned, the language is no longer transmitted to the young generation in Beirut but it is in Damascus. In northern Syria, intergenerational transmission is quite solid. The situation in southern Turkey is, according to some consultants, more precarious, but I have personally witnessed quite a few children fully conversant with the language. In any case, bilingual Doms acquire both Domari and Arabic in early childhood, making both languages equally “dominant” in Van Coestem’s (\citeyear{VanCoetsem1988,VanCoetsem2000}) terms.

Many Dom groups are also found in Eastern Anatolia. These groups have shifted to Kurdish but maintained an in-group lexicon based on Domari, locally called Domani. According to what I could personally observe on the ground and what well-informed local actors reported to me, full-fledged Domari is not spoken beyond Urfa. East of Urfa, the shift to Kurdish is complete and even the in-group lexicon is only remembered by elderly individuals. 

There are no reliable figures on the number of speakers of Domari. The language has often been mistaken for a variety of Romani but this claim has no linguistic grounds, except that they are both classified as Central Indo-Aryan Languages with a possible Dardic adstrate.
\largerpage
\section{Contact languages}

Besides a Central Indic core and a Dardic adstrate, the language exhibits various layers of influence. Easily identifiable sources of contact are Persian, Kurdish, Turkish and finally Arabic. This suggests, quite logically, that the ancestors of the Doms left the Indian subcontinent, and then travelled into Persian-speaking lands, before reaching Kurdish- and Turkish-speaking areas (most probably in eastern Anatolia), before venturing into Arab lands. It is striking to see that the Iranian and Turkic elements in Domari are not uniform across Northern and Southern varieties, which suggests an early split in eastern Anatolia between speakers of both groups. The impact of Arabic is also not uniform both across Southern and Northern Domari, nor even within Northern Domari. What this means is that the validity of any discussion of the Arabic component in Domari is limited to the varieties considered.

The Beirut/Damascus dialect is undoubtedly the most Arabized one within the Northern group, pointing to an earlier settlement of the community in an Arabic-speaking environment. Bilingualism (Domari--Arabic) is general in Lebanon and Syria. Except perhaps for very young children who have not yet acquired any other language, monolinguals in Domari are not to be found.

As far as Turkey is concerned, trilingualism in Domari, Turkish and Kurdish is not uncommon, especially in southern Turkey around Gaziantep. In the Hatay province, many speakers above the age of forty are trilingual Domari--Arabic--Turkish. The generations born here in the eighties onwards did not acquire Arabic. 

According to personal recollection from various consultants, the community of Beirut/Damascus used to spend the winter in Lebanon, and would go back to Damascus in the summer. This semi-nomadic way of life seems to have stopped when the civil war in Lebanon began. Although movements between Beirut and Damascus remained frequent, this phenomenon ceased to be seasonal. In Damascus, they settled in the area of Sayyida Zaynab, in the suburbs of the city, and in Beirut many of them settled in Sabra. Since the civil war started in Syria, virtually all the Damascus community have moved to Lebanon and settled in refugee camps in the Bekaa Valley close to the Syrian border. 

\section{Contact-induced changes in Northern Domari}

As noted above, Domari speakers in Lebanon and Syria are also fully proficient in Arabic, to the point that I have never encountered or heard of any monolingual adult. The Dom community, although largely endogamous and socially isolated, cannot afford monolingualism, primarily because of their peripatetic profile. As far as one can judge, their proficiency in Arabic is that of any monolingual native speaker of Arabic. Their pronunciation, however, is often not fully congruent with the local dialect spoken in the immediate vicinity of their settlements. This is, as usual, due to the variety of inputs and migration after acquisition. The Doms of Beirut for instance, do not speak Beirut Arabic and their speech is immediately perceived as Syrian by Lebanese because they do not raise /ā/. Raising of /ā/ towards [eː] is the hallmark of Lebanese Arabic in perceptual dialectology. Proficient speakers of Domari all exhibit Arabic--Domari bilingualism. On the whole, there is a general license to integrate any Arabic lexeme in Domari speech, even when a non-Arabic morpheme exists. Code-switching is also very common and there seems to be no conservative ideology about linguistic practices, leading to a very permissive environment for language mixing.


 
 \subsection{Phonology}


All the segmental phonology of Arabic has made its way into Domari. Arabic stands out cross-linguistically because of its series of back consonants such as the pharyngeals /ḥ/ and /ʕ/, the post-velars /q/, /ḫ/ and /ɣ/, and a set of velarized consonants whose number varies from dialect to dialect. Typically, sedentary varieties in the Levant minimally exhibit contrast between /ḍ/, /ẓ/, /ṭ/ and /ṣ/. In Domari, the pharyngeals /ḥ/ and /ʕ/ are commonly found in loans from Arabic: \textit{ḥḍər} \textit{h-} ‘watch’ (from Levantine Arabic \textit{ḥiḍir} ‘he watched’). The same goes for /ʕ/: \textit{ʕammər} \textit{kar-} `build' (from Arabic \textit{ʕammar} `he built'). An oddity surfaces in the word for coffee, realized \textit{ʔaḥwa} from Arabic \textit{ʔahwe}. These pharyngeals are also common in Kurdish-derived items such as \textit{ḥazār} ‘thousand’, \textit{moʕōri} ‘ant’ and also in the inherited (Indic) stock in \textit{ʕaqqōr} ‘nut’. Post-velar /q/, /ḫ/ and /ɣ/ are found in all the layers of the language: \textit{qāla} ‘black’ (inherited), \textit{qāpī} ‘door’ (Turkish), \textit{sāɣ} ‘alive’ (Kurdish), \textit{ɣarīb} ‘strange’ (Arabic). The most striking innovation of the Beirut/Damascus dialect is the glottal realization [ʔ] of /q/: \textit{ʔər} ‘son’ (< \textit{qər}), \textit{ʔāyīš} ‘food’ (< \textit{qāyīš}). This innovation is very likely contact-induced because it is commonly found in the Arabic dialects of both Damascus and Beirut and beyond. 

Velarized consonants mostly surface in the Arabic-derived stock as in \textit{naḍḍəf} \textit{kar-} ‘clean’ (< Arabic \textit{naḍḍaf} ‘he cleaned’), but also in pre-Arabic items: \textit{ḍāwaṭ} ‘wedding’ (borrowed from Kurdish but ultimately from Arabic \textit{daʕwa} ‘invitation’), \textit{pạ̄ṣ} ‘at him’ (< Old Indo-Aryan \textit{pārśvá} ‘side’). It is still unclear to what extent velarization in Domari continues Indo-Aryan retroflexion \citep[64]{Matras2012}. Domari also kept a contrast between /p/ and /b/, not found in Arabic: \textit{bīrōm} ‘I feared’ vs. \textit{pīrōm} ‘I drank’. 

As far as vowels are concerned, Levantine Arabic exhibits either a two-way distinction in the short vowel system (/a/ and /ə/) or a three-way distinction (/a/, /i/ and /u/). In Northern Domari, only the two short vowels /a/ and /ə/ are contrastive: \textit{kərī} ‘house’ vs. \textit{karī} ‘pot’. Such a paucity of contrastive short vowels is probably due to contact with Arabic varieties which exhibit a two-way system (/a/ vs. /ə/), such as many Lebanese and Syrian dialects. Most Arabic dialects in the area have a five-way system of long vowels because of the monophthongization of /ay/ and /aw/: /ā/, /ī/, /ū/, /ē/ and /ō/. In addition to these long vowels, Domari displays another contrast between /ā/ and a back /ạ̄/ (IPA [ɑː]): \textit{māsī} [maːsiː] ‘meat’ (< Old Indo-Aryan \textit{māṁsá})  vs. \textit{mạ̄s-ī} [mɑːsiː] ‘month-\textsc{pl}’ (< Old Indo-Aryan \textit{mā́sa}).

Domari has also preserved distinct suprasegmental features, such as final syllable stress assignment. Arabic-derived items are fully integrated into this pattern and bear final primary stress, whether common nouns or proper nouns: Domari [faːˈdja] vs. Arabic [ˈfaːdja] (personal name \textit{Fādya}). An interesting phenomenon is that Arabic epenthetic vowels in final-syllable position are reinterpreted as plain vowels and bear primary stress.  Compare Domari [sˤaˈʕab] and Arabic [ˈsˤaʕəb]) ‘difficult’; Domari [waˈdˤaʕ] and Arabic [ˈwadˤəʕ] ‘situation’.


 
 \subsection{Morphology}


Northern Domari has not borrowed any derivational or inflectional morphemes from Arabic. This is of course due to the fact that Arabic morphology is mostly non-concatenative. Borrowed morphology mostly comes from Kurdish and Turkish, whose morpheme segmentation is much more transparent. These borrowed morphemes must have entered Domari when Kurdish and Turkish were contact languages of Domari. A case in point is the Kurdish diminutive \textit{-ək}, which has made its way into all layers of the lexicon: \textit{panč-ək} ‘tail’, \textit{ḫar-ək} ‘bone’ (both Indic), \textit{taḫt-ək} ‘wood’, \textit{qannīn-ək} ‘bottle’ (both derived from Arabic: \textit{taḫt} ‘bed’ and \textit{qannīne} ‘bottle’). The dialects of northern Syria and southern Turkey have also borrowed from Kurdish the comparative suffix \textit{{}-tar}, the Turkish conditional marker \textit{{}-sa} and the Turkish superlative marker \textit{ān}. These constructions are not available in the Beirut/Damascus dialect, which relies entirely on Arabic-derived material. Compare the translation of the Arabic sentence \textit{inte} \textit{aḥsan} \textit{minni} ‘you are better than me’ into Sarāqib Domari (\ref{better}) and Beirut/Damascus Domari (\ref{better1}):

\protectedex{
\ea
{Sarāqib Domari}\\ \label{better}
\gll tō dēšōm bḫēz-tar ištōre\\
     \textsc{2sg} \textsc{1sg.abl} good\textsc{-cmpr} \textsc{cop.2sg}\\
\glt ‘You are better than me.’
\z
}
\ea 
{Beirut/Damascus Domari}\\ \label{better1}
\gll tō aḥsan wēšōm ištōr\\
 \textsc{2sg} better \textsc{1sg.abl} \textsc{cop.2sg} \\
\glt ‘You are better than me.’\\
\z

Sarāqib is located in northern Syria and the dialect spoken by the Doms of Sarāqib is a good representative of the Domari of northern Syria and southern Turkey. Three differences are immediately apparent. The first is morphological, whereby there are different forms for the ablative of the first-person pronoun. The second difference is syntactic: in (\ref{better}) the standard of comparison precedes the comparative adjective (\textit{dēšōm} \textit{bḫēz-tar}) and in (\ref{better1}) it follows it (\textit{aḥsan} \textit{wēšōm}).\footnote{Comparative constructions typically involve two noun phrases (NPs). \citet{Stassen2013} labels the object of comparison the “compare NP” and the other the “standard NP”.} The Beirut/Damascus Domari syntax exhibits full congruence with the Arabic syntax. The third difference is lexical. Because Beirut/Damascus Domari does not have at its disposal the morpheme \textit{{}-tar}, speakers are obliged to draw on Arabic for the comparative. This phenomenon, labelled ``bilingual suppletion'' by Matras, is described at length for Jerusalem Domari (\citealt[379--382]{Matras2012}; see also Matras, this volume: §3.5).

Beirut/Damascus Domari also relies entirely on Arabic material for the expression of time and date, as shown in (\ref{manane}). In northern Syria, speakers favour the use of inherited numerals, as exemplified in (\ref{hatta}).

\ea
{Beirut/Damascus Domari}\\ \label{manane}
\gll mānane mi-s-sāʕa ʕašra la s-sāʕa sabʕa tmāne ōtanta sa čāɣ-an-sa\\
     stay.\textsc{ipfv.1pl} from-\textsc{det}{}-hour ten to \textsc{def}{}-hour seven eight there all children-\textsc{obl.pl-com}\\
   \glt ‘We stay there with the all the kids from ten o'clock to seven or eight o'clock.’\\
\z
\protectedex{
\ea \label{hatta}
{Sarāqib Domari}\\
\gll ḥatta saʕat štār ēwar mānde ē čōrt-ə{}-ma\\
     until hour four evening stay.\textsc{pfv.3sg} \textsc{dem.obl} wasteland-\textsc{obl-in}\\
\glt ‘He stayed until 4pm in this wasteland.’
\z
}
Some speakers of Beirut/Damascus Domari also extend the use of Arabic to higher numerals because, according to their own judgment, they have difficulties retrieving the pre-Arabic options. A look at their distribution reveals that the main parameter that triggers the use of Arabic items is not so much high numerals, but rather the complexity of the numeral. Compare in this regard (\ref{parda}) and (\ref{nacis}). In (\ref{parda}), the speaker uses Arabic for the more complex numeral ‘95000’ but uses Domari items for simpler `2000', `3000' and `4000'.

\ea
{Beirut/Damascus Domari}\\ \label{parda}
\gll pārda abōs šaʔʔ-āka ši ḫamse u tisʕīn alf dolar\\
     buy.\textsc{pfv.3sg} \textsc{3sg.ben} flat\textsc{{}-indf} about five and ninety thousand dollar\\
\glt ‘He bought a flat for her, about ninety-five thousand dollars.’
\z

\ea \label{nacis}
{Beirut/Damascus Domari}\\
\gll načīš-a-ki dī ḥazār trən ḥazār štār ḥazār dfaʕ kaštand dādōs kē\\
     dancing-\textsc{obl-abl} two thousand three thousand four thousand pay do.\textsc{prog.3pl} her.mother \textsc{ben}\\
\glt ‘They give two, three, four thousand (dollars) to her mother from dancing.’
\z

As noted above, it appears that the use of Arabic numerals is closely linked to language dominance. Speakers themselves are aware of it and when asked why they do not use Domari numerals, they justify it claiming a lack of proficiency. Looking at the distribution of inherited and Arabic numerals is therefore a good way to assess whether language attrition is incipient or not.

The impact of Arabic is also apparent in some morphological differences between the Beirut/Damascus variety and the dialects of northern Syria. For instance, the verb \textit{sək-} means ‘to learn’. The Beirut/Damascus dialect adds the passive suffix \textit{-yā}/\textit{-ī}. The corresponding verb in Arabic \textit{tʕallam} is marked with the valency-decreasing prefix \textit{t-}. What the speakers of the Beirut/Damascus dialect have done is to replicate the valency-decreasing prefix \textit{t-} of \textit{tʕallam} by means of the Domari passive suffix  \textit{yā}/\textit{{}-ī}: \textit{skə-rd-ōm} (learn-\textsc{pfv-1sg}; northern Syria) vs. \textit{sk-ī-r-ōm} (learn-\textsc{pass-pfv-1sg}; Beirut/Damascus) ‘I learnt’.

Unlike Southern Domari, Northern Domari does not normally transfer Arabic plurals. Speakers simply use the singular form and add the Domari plural marker \textit{-ī(n)}:  \textit{azʕar-īn} ‘thugs’ instead of the Arabic plural \textit{zuʕrān}. Arabic plurals do surface at times, but only when they exhibit a high degree of independence within the lexicon. Examples are \textit{ʔarāyb-ē-mā} (relatives-\textsc{pl-1pl}) ‘our relatives’, \textit{\v{g}īrān-ē-mā} (neighbors-\textsc{pl-1pl}) ‘our neighbors’, from Arabic \textit{qarāyib} and \textit{\v{g}īrān}. Although these items have singular forms (respectively \textit{qarīb} and \textit{\v{g}ār}), they are arguably lexicalized plurals and independent entries in the Arabic lexicon.


 
 \subsection{Syntax}
 \subsubsection{Constituent order}

The impact of Arabic in the realm of syntax is not uniform across Domari dialects. Dialects of northern Syria and southern Turkey show a strong tendency towards a head-final constituent-order typology, both within the NP and the clause. This feature is areal, so its presence in Domari may well be contact-induced. The canonical syntax of the NP is (demonstrative) (numeral) (adjective) (noun) noun. Complex NPs could only be retrieved through elicitation (examples \REF{star} to \REF{dom}) and hardly occur in spontaneous speech. Example (\ref{star}) illustrates the canonical syntax, where all the modifiers appear to the left of the head. Speakers of Beirut/Damascus Domari, however, tend to dislocate some modifiers to the right of the head, converging towards the Arabic syntax, as in (\ref{dad}), (\ref{nam}) and (\ref{dom}).

\ea
{Sarāqib Domari}\\ \label{star}
\gll ē štār lāfty-ən-ki dād-ō-sā\\
     \textsc{dem} four girl-\textsc{obl.pl-abl} mother-\textsc{sg-3pl}\\
\glt ‘the mother of these four girls’
\z

\ea \label{dad}
{Beirut/Damascus Domari}\\
\gll dād-ō-sā štār lāfty-an-ki\\
     mother-\textsc{sg-3pl} four girl-\textsc{obl.pl-abl}\\
\glt ‘the mother of the four girls’
\z
\protectedex{
\ea \label{nam}
{Beirut/Damascus Domari}\\
\gll nām-ē-sā \v{g}əwr-an-ki tərn-an-ki\\
     name-\textsc{pl-3pl} woman-\textsc{obl.pl-abl} three-\textsc{obl.pl-obl}\\
\glt ‘the names of the three girls’
\z
}
\ea \label{dom}
{Beirut/Damascus Domari}\\
\gll dōm-an-sa ēr-an-sa štār-an-sa\\
     dom-\textsc{obl.pl-com} \textsc{dem-obl.pl-com} four-\textsc{obl.pl-com}\\
\glt ‘with these four Doms’
\z

In (\ref{nam}), the speaker also dislocates to the right the numeral \textit{trən} ‘three’ which normally appears to the left giving the expected order \textit{trən} \textit{\v{g}əwr-an-ki nām-ē-sā} (three woman-\textsc{obl}.\textsc{pl}{}-\textsc{abl} name-\textsc{pl}{}-\textsc{3pl}). The numeral remains unmarked for case when it appears to the left of the head. When it is placed to the right, it agrees in case with the head. This is also the case with the demonstrative in (\ref{dom}). Here the normal order would be \textit{ē} \textit{štār} \textit{dōm-an-sa} (\textsc{dem} four Dom-\textsc{obl.pl}{}-\textsc{com}). The fact that speakers replicate case marking on right-dislocated modifiers suggests that they feel the need to strengthen constituency in case of non-canonical syntax.

The influence of Arabic also surfaces in the Beirut/Damascus dialect in the syntax of the quantifier \textit{sa} ‘all’. This is normally located to the right of the head: \textit{ammat} \textit{sa} ‘all the people’ (`people all'). In Beirut/Damascus, \textit{sa} consistently surfaces to the left, like the Arabic quantifier \textit{kull}: \textit{sa} \textit{ammat} (Arabic \textit{kull} \textit{in-nās}).\footnote{Arabic \textit{kull} can also appear to the right as in \textit{in-nās} \textit{kull-ha} \textit{{\textasciitilde} kull-hum} ‘all the people’ but this is a marked syntax.}


 \subsubsection{Internal object}

Domari speakers regularly replicate Arabic constructions and idioms, but tend to do so by recruiting inherited or pre-Arabic material -- they do not borrow Arabic material. For instance, all dialects have replicated the so-called internal object construction, commonly used in Arabic as a predicate-modifying construction. Consider for instance (\ref{jordar}) in Jordanian Arabic, where the speaker narrows the scope of the predication using the verbal noun \textit{ʕirəf} ‘knowledge’, derived from the verb \textit{ʕirif} ‘he knew’, and modifies it with the adjective \textit{ṭayyib} ‘good’. In (\ref{dados}), the speaker has used the deverbal derivation \textit{kūš} from the root \textit{kū-} ‘throw’ and coded it as an object, as evident from the accusative marker \textit{-əs}. This replicates the Arabic internal object construction as illustrated in (\ref{jordar}).

\ea
{Jordanian Arabic}\\ \label{jordar}
\gll baʕrif-hum ʕirəf ṭayyib\\
     know.\textsc{impf.1sg-3pl} knowledge good\\
\glt ‘I know them well.’
\z

\ea \label{dados}
{Sarāqib Domari}\\
\gll dād-ōs ibnḥarām e ē kūš-əs ktōs-s-e\\
     mother-\textsc{3sg} son.of.illicit \textsc{cop} \textsc{dem} throwing-\textsc{acc} throw.\textsc{pfv.3sg-obj.3sg-prs}\\
\glt ‘His mother is heartless for having thrown (her baby) in such a way.’
\z


 \subsubsection{Impersonal construction}

Speakers also replicate the Arabic impersonal construction with the indefinite pronoun \textit{il-wāḥad} by way of the inherited noun \textit{mānəs} ‘individual, people’. Example (\ref{jordar1}) illustrates the use of \textit{il-wāḥad} in (Jordanian) Arabic. In (\ref{asti}), the sequence \textit{gzare} \textit{māns-as} corresponds to Arabic \textit{biʕiḍḍ} \textit{il-wāḥad,} literally ‘it bites one’. The fact that \textit{māns-as} replicates \textit{il-wāḥad} is also apparent from the accusative marking in Domari, which normally surfaces only with definite objects. The referent here is by nature indefinite and non-referential, so accusative marking in Domari can only be explained by the presence of the definite article \textit{il-} in Arabic \textit{il-wāḥad}.

\ea
{Jordanian Arabic}\\ \label{jordar1}
\gll kān ʕēb il-wāḥad yrūḥ ʕala ʔutēl\\
     be.\textsc{prf.3sg.m} shameful \textsc{def-}one go.\textsc{impf.sbjv.2sg.m} to hotel\\
\glt ‘One was ashamed to spend the night in a hotel.’
\z

\ea \label{asti}
{Beirut Domari}\\
\gll ašti ši hana lli baḥr-a-ma e gzare māns-as\\
     \textsc{exs} too \textsc{dem} \textsc{rel} sea-\textsc{obl-in} \textsc{cop} bite.\textsc{ipfv.3sg} man-\textsc{acc}\\
\glt ‘There is this thing in the sea, it bites you.’
\z


 \subsubsection{Auxiliaries}

Probably the most striking difference between Southern and Northern Domari as far as the Arabic component is concerned is the absence of Arabic inflected material in the latter. Only the dialect of Beirut/Damascus has borrowed the auxiliaries \textit{kān} (with its imperfect form \textit{bikūn}), \textit{ṣār} and \textit{ḫalli}.

\ea
{Beirut Domari}\\ \label{sar}
\gll ṣār \v{g}ahhəz lakand lāfty-a kē bḫēr\\
     become.\textsc{prf.3sg} prepare do\textsc{.sbjv.3pl} girl\textsc{{}-obl} \textsc{ben} well\\
\glt ‘They prepare the girl well now (for the wedding).’
\z

\ea \label{xadra}
{Beirut Domari}\\
\gll ḫaḍra kān mə\v{g}nār-a\\
     Khadra be.\textsc{prf.3sg.m} breastfeed.\textsc{ipfv.3sg-pst}\\
\glt ‘Khadra was breastfeeding.’
\z

\ea \label{awande}
{Beirut Domari}\\
\gll āwande bikūn krēnde mā kē kyāmōr\\
     come.\textsc{ipfv.3pl} be.\textsc{impf.3sg} do.\textsc{prf.3pl} \textsc{1sg} \textsc{ben} something\\
\glt ‘(My kids) would come and they would have done something (naughty).’ 
\z

In (\ref{sar}), the subject is in the 3\textsc{pl} but \textit{ṣār} remains invariable, as the \textsc{3pl} is \textit{ṣāru}. In (\ref{xadra}), the subject is feminine so if there was agreement one would expect \textit{kānat}, not masculine \textit{kān}. A further intriguing feature in (\ref{xadra}) is the redundancy in past marking, first with \textit{kān} and second with the past suffix \textit{{}-a}, which in northern Syria and southern Turkey Domari suffices to mark past tense. The same invariability is apparent in (\ref{awande}) where the \textsc{3pl} of \textit{bikūn} should be \textit{bikūnu}. These auxiliaries have the same semantic load as in Arabic. The morpheme \textit{ṣār} puts emphasis on the inception of the event, \textit{kān} followed by the imperfect places the event in the past and gives it an iterative/habitual aspect and \textit{bikūn} describes a possible state of affairs not attested at the time of utterance. Arabic \textit{ṣār}, \textit{kān} and \textit{bikūn} are absent in the dialects of northern Syria and southern Turkey. The only auxiliary that has been replicated here is \textit{ṣār}. These dialects, however, have only replicated the structure, not the substance, that is they rely on inherited morphemes, as exemplified in (\ref{har}). The speaker simply translates Arabic \textit{ṣār} with the Domari equivalent \textit{hra}, replicating the Arabic structure \textit{ṣār} + subjunctive (see Manfredi, this volume).\ia{Manfredi, Stefano@Manfredi, Stefano} A further difference is word order, with the verb placed clause-finally in the subordinate clause. 

\ea \label{har}
{Sarāqib Domari}\\
\gll hər wārsīndạ lwār\\
     become.\textsc{pfv.3sg} rain hit.\textsc{sbjv.3sg}\\
\glt ‘It started raining.’
\z

As noted above, in these dialects the functions of Arabic \textit{kān} are expressed by the inherited past suffix \textit{-a}. The functions covered by Arabic, \textit{bikūn}, however do not seem to be encoded in the grammar of these dialects.

In Levantine Arabic, the imperative form \textit{ḫalli} ‘let’ of \textit{ḫalla} ‘he let’ is often used to soften an order and allows the speaker to avoid using an imperative, flagging a suggestion or an invitation, as shown in (\ref{xalli}):

\ea \label{xalli}
{Jordanian Arabic}\\
\gll ḫalli ibn-ak yrūḥ la \v{g}-\v{g}ēš\\
     let son-\textsc{2sg.m} go.\textsc{impf.3sg.m} to \textsc{def}{}-army\\
\glt ‘Let your son serve in the army.’
\z

This auxiliary has been borrowed into Beirut/Damascus Domari with the exact same function, as illustrated in (\ref{xalli1}). In this case too, \textit{ḫalli} remains invariable and does not surface as \textit{ḫallī-(h)un} (let.\textsc{imp.2sg-3pl}) as it would in Beirut/Damascus Arabic. Here again, the dialects of northern Syria and southern Turkey have borrowed the structure, but not the substance, and use the inherited root \textit{mək} ‘let’, as exemplified in (\ref{mek}).

\ea \label{xalli1}
{Beirut/Damascus Domari}\\
\gll ḫalli \v{g}ānd dfən lakrand-əs\\
     let go.\textsc{sbjv.3pl} bury do.\textsc{sbjv.3pl-3sg}\\
\glt ‘Let them go and bury him.’
\z

\ea \label{mek}
{Aleppo Domari}\\
\gll mək pāʋər pạ̄sōr\\
     let come.\textsc{sbjv.3sg} \textsc{2sg.ad}\\
\glt ‘Let him come to your place.’
\z


 \subsubsection{Negation}

Only two Arabic negators have made their way into the grammar of Northern Domari: Damascus Arabic \textit{mū} and the contrastive negative coordination markers \textit{lā...walā} ‘neither…nor’. Arabic \textit{mū} is only available in the dialect of Beirut/Damascus. Its distribution and functions, however, only partially match those of Damascus Arabic. The primary function of \textit{mū} in Damascus Arabic is to negate non-verbal predicates. This is not attested in Domari, which relies for this purpose only on inherited \textit{nye}. \textit{mū} surfaces first when negation has scope over non-clausal constituents, as shown in (\ref{seff}), and second when the predicate is in a non-indicative mood (subjunctive, jussive and imperative) as in (\ref{guz}):

\ea
{Beirut/Damascus Domari}\\ \label{seff}
\gll səff (h)ra wāšya mū wāšōm\\
     side become\textsc{.pfv.3sg} \textsc{3pl.com} \textsc{neg} \textsc{1sg.com}\\
\glt ‘He took sides with them, not with me.’
\z

\ea \label{guz}
{Beirut/Damascus Domari}\\
\gll bi\v{g}ūz mū māntyar wāš məṣrī\\
     possible \textsc{neg} stay\textsc{.sbjv.3sg} \textsc{3sg.com} money\\
\glt ‘He might not have any money left.’
\z

The Arabic structure \textit{lā…walā} is readily available in all varieties, but whereas it is the only option in Beirut/Damascus, it competes with the inherited structure \textit{nə…nə} in northern Syria and southern Turkey. Interestingly, this clash has led to a mixed form \textit{nə…walā}, as shown in (\ref{mix}). The Domari syntax is also reminiscent of the Turkish possessive predication syntax with possessive marking on the noun and an existential morpheme.

\ea \label{mix}
{Antioch Domari (Southern Turkey)}\\
\gll nə lawr-ōs ašti wala šarš-ōs ašti\\
     \textsc{neg} tree\textsc{-3sg} \textsc{exs} \textsc{neg} root\textsc{-3sg} \textsc{exs}\\
\glt ‘It doesn’t grow on a tree nor has it roots.’
\z


 \subsubsection{Complex sentences}

Complex sentences minimally include coordinated and subordinate clauses. The Arabic coordinators \textit{w} ‘and’, \textit{aw} ‘or’, \textit{walla} ‘or’, \textit{bass} `but' and others have all made their way into Domari. Originally, Domari seems to have distinguished clausal coordination from phrasal coordination, a not so frequent feature from a typological point of view. Nominal categories are coordinated with the Turkish-derived morpheme \textit{la} and clauses are coordinated with the Kurdish-derived enclitic -\textit{ši}. The intrusion of Arabic \textit{w}, which in Arabic is used indiscriminately for both kinds of coordination, has led to the marginalization of the original system in Beirut/Damascus Domari, which now tends to favour the use of Arabic \textit{w}.

\ea
{Beirut/Damascus Domari}\\
\gll illi mangar tōre māṣṭ-a-ma w illi mangar ʔār-s-e nāšif\\
     \textsc{rel} want.\textsc{ipfv.3sg} put.\textsc{ipfv.3sg} yoghurt-\textsc{obl-in} and \textsc{rel} want.\textsc{ipfv.3sg} eat.\textsc{ipfv.3sg-obj.3sg-prs} dry\\
\glt ‘Some eat it in yoghurt, some eat it dry.’
\z

As far as phrasal coordination is concerned, some alternation between Arabic \textit{w} and Turkish-derived \textit{la} is still observed: \textit{dōmwārī} \textit{w} \textit{ṭāṭwārī} ‘Domari and Arabic’ {\textasciitilde} \textit{dōm} \textit{la} \textit{ʕarabi} ‘Domari and Arabic’.

Virtually all the conjunctions of subordination found in Domari are borrowed from Arabic. This includes the relativizer \textit{illi}, the complementizer \textit{inno} and potentially all the adverbial conjunctions found in Levantine Arabic: \textit{lamma} ‘when’, \textit{qabəl-mā} ‘before’, \textit{baʕəd-mā} ‘after’, \textit{ʕa-bēn-mā} ‘by the time’, and many more. Pre-Arabic constructions are attested for relativization and conditional clauses, but these only survive in the dialects of northern Syria and southern Turkey, and tend to be replaced by Arabic material (except in the varieties spoken in Turkey). A case in point is conditional clauses. Arabic \textit{iza} and \textit{law} are available everywhere, even in Turkey, as shown in (\ref{aza}), recorded in Antioch. In this example, the speaker uses the Arabic morpheme \textit{aza} (< \textit{iza}) in the first sentence of the utterance, and no overt marking in the protasis, making parataxis a possible means to express condition. As far as counterfactual conditions are concerned, it appears that the dialect of Beirut/Damascus is fully congruent with Arabic in having borrowed also the morpheme \textit{kān} in both the protasis and the apodosis, as shown in (\ref{law}). The dialects of northern Syria and southern Turkey exhibit a native strategy using subjunctive mood and past marking in the protasis and perfective and past marking in the apodosis. The two clauses are coordinated with the Kurdish derived enclitic \textit{ši} (\ref{all}).

\ea \label{aza}
{Antioch Domari}\\
\gll aza kām karne qāne kām nə-karne nə-qāne\\
     if work do.\textsc{ipfv.1pl} eat.\textsc{ipfv.1pl} work \textsc{neg}{}-do.\textsc{ipfv.1pl} \textsc{neg}{}-eat.\textsc{ipfv.1pl}\\
\glt ‘If we work, we eat, (if) we don’t work, we don’t eat.’
\z

\ea \label{law}
{Beirut/Damascus Domari}\\
\gll law kān nəčnār-sā bāb-ōm kān abṣar kaki (h)re\\
     if be.\textsc{prf.3sg} make.dance\textsc{.ipfv.3sg-obj.3pl} father-\textsc{1sg} be.\textsc{prf.3sg} not.know what become.\textsc{pfv.3sg}\\
\glt ‘If my father had put them to dance, I don’t know what would have happened.’
\z
\protectedex{
\ea \label{all}
{Sarāqib Domari}\\
\gll aḷḷ-əs byātyənd-a nə-ktēnd-s-a ši\\
     God-\textsc{acc} fear.\textsc{sbjv.3pl}{}-\textsc{pst} \textsc{neg}{}-throw.\textsc{pfv.3pl-obj.3sg-pst} and\\
\glt ‘Had they feared God, they would not have thrown him.’
\z
}
 \subsection{Lexicon}
 \subsubsection{Function words}

Arabic prepositions do occur in Domari, but these are mostly non-core prepositions such as \textit{qabəl} ‘before’, \textit{baʕad} ‘after’, \textit{minšān} ‘for’, \textit{ɣēr} ‘other’. Some have made their way into Domari only recently, and still alternate with pre-Arabic options, such as the Iranian equative morpheme \textit{war,} which tends to be replaced by Arabic \textit{mitəl} ‘as, like’ especially in the dialect of Beirut/Damascus. Currently, \textit{war} and \textit{mitəl} are in a quasi-complementary distribution, with \textit{war} being used with full NPs and \textit{mitəl} with pronouns, as shown below in (\ref{rom}) and (\ref{tani}):

\ea
{Beirut/Damascus Domari}\\ \label{rom}
\gll tō ʔr-ōm war ištōr\\
     you son-\textsc{1sg} like \textsc{cop.2sg}\\
\glt ‘You are like my son.’
\z

\ea \label{tani}
{Beirut/Damascus Domari}\\
\gll tāni ʔər gēna mitl-ōs kām karre\\
     second son also like-\textsc{3sg} work do.\textsc{ipfv.3sg}\\
\glt ‘My second son has the same job.’
\z

The Arabic core preposition \textit{b-} ‘in, with’ occurs in Domari, but it appears to be restricted to certain constructions and idioms such as \textit{gāl} \textit{b-gāl} ‘discussion’ (word in-word), \textit{ārāt} \textit{əb-dīs} ‘night and day’ (night in-day), \textit{b-rəbʕ-āk} ‘for a quarter of a pound’ (with-quarter-\textsc{indf}). The preposition \textit{min} ‘from’ also sporadically occurs in Beirut/Damascus Domari:

\ea
{Beirut/Damascus Domari}\\
\gll min ši šēš mạ̄s ǧərsa krōm ḍāwaṭ-ōs\\
     from about six month wedding do.\textsc{pfv.1sg} wedding-\textsc{3sg}\\
\glt ‘Some six months ago I married him off.’
\z
 
Domari also borrows high-frequency adverbs, fillers, connectors and all kinds of discourse-structuring devices, such as \textit{masalan} ‘for instance’, \textit{abadan} ‘at all, never’, \textit{yaʕni}  ‘I mean’, \textit{aywa} ‘yes, so’, \textit{waḷḷa} ‘I swear’, \textit{inno} (complementizer and discourse marker) and many more. One finds also common adverbial phrases such as \textit{ṭūl} \textit{in-nhār} ‘all day long’, \textit{ṭūl} \textit{il-waʔət} ‘all the time’, and \textit{ʕala} \textit{ṭūl} ‘immediately’.  The very common Domari phrase \textit{tīka} \textit{tīka} ‘slowly’ replicates Arabic \textit{šwayy} \textit{əšwayy}.


 \subsubsection{Content words}

In Syria and Lebanon, Arabic is the \textit{de facto} lexical reservoir of Domari, so there is a general licence to integrate any element from Arabic if no pre-Arabic option exists. The issue is the replacement of pre-Arabic options with Arabic material. There is of course a certain amount of variation in lexical knowledge across speakers, but it seems possible to differentiate several levels of replaceability. Some items have long been replaced by Arabic words, and only a handful of speakers are able to retrieve them, such as \textit{lōrga} ‘tomato’ or \textit{pīsənga} ‘bulgur’, replaced respectively by Arabic \textit{bandōra} and \textit{bərɣəl}. Other items tend to be replaced by Arabic equivalents but may still surface in the speech of some speakers, such as \textit{čatīn} ‘hard’, \textit{čirkī} ‘bird’, \textit{alčāḫ} ‘low’ replaced by Arabic \textit{ṣaʕab}, \textit{ṭēr}/\textit{ʕaṣfūr} and \textit{wāṭi}. Some items seem stable but are sporadically replaced with Arabic-derived items such as \textit{drəs} \textit{kar-} ‘study’ instead of inherited \textit{sək}{}-. Finally, other items  such as \textit{\v{g}awwəz h-} ‘get married’ and \textit{\v{g}irsāwī h-} freely alternate. It appears therefore that every pre-Arabic item is somewhere on a continuum of replaceability from ``very unlikely'' to ``completely disappeared''. To illustrate the variablility in replaceability judgment, I remember an elicitation session in Aleppo with a father and his son. One of the sentences contained the Arabic word \textit{baṣal} ‘onion’. The son simply translated the sentence with the Arabic word \textit{baṣal} but the father strongly objected to this answer, stating that the proper Domari word is \textit{pīwāz}.

As noted above, Arabic nouns are integrated in their singular form, except in the case of lexicalized plurals. Adjectives are borrowed in their masculine form and never agree in gender, as shown in (\ref{hala}). Other than the past copula \textit{a}, all the words in this example are Arabic. Two features, however, allow its identification as Domari. First, \textit{ḥāla} is realized without raising (also stressed on the last syllable [ħaːˈla]), unlike Levantine Arabic \textit{ḥāle,} and second \textit{taʕbān} does not agree in gender with \textit{ḥāla} and surfaces in its masculine form, instead of feminine \textit{taʕbāne}, as it would normally occur in Arabic.

\protectedex{
\ea \label{hala}
{Beirut/Damascus Domari}\\
\gll ʔabəl ḥāla taʕbān a\\
     Before situation tired \textsc{cop.pst}\\
\glt ‘Before, the situation was bad’
\z
}

Arabic verbs are easily integrated into Domari, because Domari has a light verb strategy. Roughly, transitive verbs tend to be integrated with the light verb \textit{kar-} ‘do’: \textit{rabbī} \textit{kar-} ‘raise’ from Arabic \textit{rabba}, \textit{yrabbi} ‘raise’. Intransitive verbs are integrated with \textit{h-} ‘become’: \textit{ʕīš} \textit{h-} ‘live’ from Arabic \textit{ʕāš}, \textit{yʕīš} ‘live’. While all the verbs that are integrated with \textit{kar-} are transitive, some verbs integrated with \textit{h-} are not intransitive: \textit{lməs} \textit{(h)rōs-s-e} ‘he has touched it’ (touch become.\textsc{pfv.3sg-3sg-prs}) from Arabic \textit{lamas}, \textit{yilmis} ‘touch’. This seems to happen with transitive verbs that are lower on the transitivity scale, or at least perceived to be so. In the case of \textit{lamas}, \textit{yilmis}, its integration into Domari by way of the light verb \textit{h-} suggests that speakers perceive it as less transitive. Formally, speakers isolate the imperfect stem of the verb, and apply a vocalism in /i/: \textit{nsī} \textit{kar-} ‘forget’ and \textit{stannī} \textit{kar-} ‘wait’, from the Arabic imperfect stems of \textit{nsa} ‘forget’ and \textit{stanna} ‘wait’.\footnote{These verbs are only available in Beirut/Damascus, other dialects use respectively \textit{ziwra} \textit{kar-} and \textit{akī} \textit{kar-}.} An exception to this tendency occurs with the so-called hollow roots in Arabic whose imperfect stem is CūC. In this case, speakers simply extract the imperfect stem and leave it unchanged: \textit{zūr} \textit{h-} ‘visit’, \textit{dūr} \textit{h-} ‘turn’, \textit{ʕūz} \textit{h-} ‘need’, from the Arabic imperfect stems \textit{zūr}, \textit{dūr} and \textit{ʕūz}.

Some English-derived items were also recorded in the Beirut/Damascus dialect, such as \textit{mōmari} ‘memory card’, \textit{hambarga} ‘hamburger’ and, more surprisingly, \textit{tōman\v{g}īre} ‘Tom and Jerry’ [toːmanʤiːˈre], expectedly stressed on the last syllable.


 \subsubsection{Speech sample}

Probably the best way to capture how Arabic integrates into Domari is to consider a piece of spontaneous speech, reproduced below in (\ref{speech}). It is part of a recorded discussion I had with a consultant in her mid-thirties in Beirut. It illustrates the level of endangerment of Beirut Domari. The consultant belongs to the last generation of fluent speakers. Her children did not acquire the language. According to what she reports, she was unable to speak to her children in their early childhood because her husband, who is a semi-speaker of Domari, prevented her from transmitting the language. Her daughter-in-law, aged twenty-one at that time, is also a fluent speaker of Domari because she grew up in Damascus, where language transmission was more solid than in Lebanon. Both of them use Domari in the household. Her son reacts negatively when he hears it, and even labels it \textit{a\v{g}nabi} ‘foreign, non-Arabic’. Linguistically, the text illustrates some of the features discussed above. Arabic-derived items are marked in boldface. 

\ea
{Beirut Domari}\\ \label{speech}
\gll nā n-ǧib karre pānǧī gāl karre gāl karre dōm wāšōm mā gāl kame wāšī \textbf{ʕādi} \textbf{bass} əʔr-ōm ʔzīn karre wat ftyare ma-gāl ka \textbf{aǧnabí} nə-\textbf{fəmm} (h)ōme watōr, gāl karse \textbf{ʕarabiy}{}-a-ma \textbf{yaʕni} ma-gāl k(a) ēhānī \textbf{laʔanno} n-\textbf{fəmm} (h)ōre watī \textbf{bass} mā l pānǧī ǧib kane \textbf{ṭūl} \textbf{il-waʔət} kəry-a-ma \textbf{yaʕni} \textbf{iza} mā l pānǧi štēn kəry-a-ma \textbf{ṭūl} \textbf{in-nhār} gāl kane dōm-a-ma \textbf{yaʕni} ʔr-ōm wāri \textbf{ʕəmr}-ōs \textbf{wāḥad} \textbf{u} \textbf{ʕišrīn} \textbf{sane} \textbf{akbar} ʔr-ōm-ki \textbf{b}{}-trən wars  \textbf{mū} \textbf{ʕādi} \textbf{ʕādi} nye amīn \textbf{lāzim} lpāran \textbf{azɣar} wēšōma \textbf{bass} bxēz e \textbf{u} \textbf{ādami} e \textbf{u} \textbf{maḥšūm} e mā ēhāny-a xr-a kē pārdōm-əs ʔr-ōm kē \textbf{u} \textbf{ǧamāʕt}-ēm kē skīr(a) ēta \textbf{baʕdēn} skīra \textbf{mahná} \textbf{baʕdēn} kām əkra wars-ā wars-ā nīm \textbf{makanīk} \textbf{baʕdēn} wəndrārda \textbf{u} īsa nə-kām kištar \textbf{wala} kkyā wēsre kəry-a-ma\\
no \textsc{neg}{}-tongue do\textsc{.ipfv.3sg} \textsc{3sg} word do.\textsc{ipfv.3sg} word do\textsc{.ipfv.3sg} Dom \textsc{1sg.com} \textsc{1sg} word do.\textsc{ipfv.1sg} \textsc{3sg.com} normal but son-\textsc{1sg} shout do\textsc{.ipfv.3sg} \textsc{3sg.supr} say.\textsc{ipfv.3sg} \textsc{neg}{}-word do.\textsc{sbjv.2sg} foreign \textsc{neg}{}-understand become.\textsc{ipfv.3sg} \textsc{2sg.supr} word do.\textsc{ipfv.2pl} Arabic-\textsc{obl-in} I.mean \textsc{neg}{}-word do.\textsc{sbjv.2sg} so because \textsc{neg}{}-understand become.\textsc{ipfv.3sg} \textsc{3sg.supr} but \textsc{1sg} and \textsc{3sg} tongue do.\textsc{ipfv.1pl} length \textsc{def}{}-time house-\textsc{obl-in} I.mean if \textsc{1sg} and \textsc{3sg} \textsc{cop.1pl} house-\textsc{obl-in} length \textsc{def}{}-day word do\textsc{.ipfv.1pl} Dom-\textsc{obl-in} I.mean son-\textsc{1sg} bride age-\textsc{3sg.f} one and twenty year bigger son-\textsc{1sg-abl} with-three year \textsc{neg} normal normal \textsc{cop.neg} \textsc{1pl} must take\textsc{.sbjv.1pl} smaller \textsc{1pl.abl} but good \textsc{cop} and humane \textsc{cop} and respectful \textsc{cop} so so-\textsc{obl} heart-\textsc{obl} \textsc{ben} take.\textsc{pfv}.\textsc{1sg-obj.3sg} son-\textsc{1sg} \textsc{ben} and folks-\textsc{1sg} \textsc{ben} learn.\textsc{pfv.3sg} here then learn.\textsc{pfv.3sg} profession then work do.\textsc{pfv.3sg} year-\textsc{indf} year-\textsc{indf} half mechanic then fire.\textsc{pfv.3sg} and now \textsc{neg}-work do.\textsc{prog.3sg} nor thing sit.\textsc{pfv.3sg} house-\textsc{obl-in}\\
\glt ‘No, [my son] doesn’t speak [Domari], [my daughter-in-law] does, she speaks with me, I speak with her normally but my son shouts at her and tells her: “Don’t speak foreign, I don’t understand you, you all speak in Arabic, don’t speak like this”, because he doesn’t understand her. But me and her we speak all the time in Domari, that is, if both of us are in the house, all day long we speak in Domari. The bride of my son, she is twenty-one years old, three years older than my son, it’s not usual, we [women] have to take someone older, but she is a good person, humane and respectful. That’s why I took her for my son and my family. [My son] studied here [in the school]. After that he went for vocational training and worked for a year a year and a half as a mechanic -- then he renounced. And now he doesn’t do anything, he stays home.’ 
\z

\section{Conclusion}

Multilingualism seems to have been a normal state of affairs amongst the Doms for a very long time, probably since the genesis of the community. The reason for this is mostly because the sociolinguistics of Domari has in likelihood remained unchanged throughout the centuries: Domari is a community language whose use is restricted to in-group communication. Out-group interactions imply the use of the majority language. Due to the very nature of their occupational profile, peripatetic groups are forced to have frequent interactions with outsiders. This involves \textit{de} \textit{facto} high levels of bilingualism. Although it is hard to assess whether the dominant language is the insider code or the outsider code, it makes sense to suspect that balanced bilingualism was the norm, as much in the past as in the present.

Van Coetsem (\citeyear{VanCoetsem1988,VanCoetsem2000}) uses the term “transfer” generically for any kind of contact-induced phenomenon. If the transfer is triggered by speakers who are dominant in the source language, he uses the term “imposition”. If it originates from recipient-language dominance, it is called “borrowing”. Lucas (\citeyear[525]{Lucas2015}) further introduces two categories, the first of which he calls ``restructuring'', defined as a “type of change […] brought about by speakers for whom the changing language is an L2, but it does not involve transfer”.  He notes that for individuals who acquired two languages simultaneously (in early childhood), “the distinction between borrowing and imposition breaks down”. In this case, both languages typically undergo ``convergence'', that is the fourth category of contact-induced change. Because I posit balanced Arabic--Domari bilingualism as the norm, the question that needs to be answered is whether all the contact-induced changes happening in Domari are the product of convergence, or whether there are changes that can be attributed to Arabic dominance (source-language agentivity or imposition). Another problem concerns the sociolinguistic limits of the model. Speakers with two first languages are expected to initiate changes that target both languages. When languages exhibit unbalanced sociolinguistic statuses (minority versus majority), one wonders how changes originating from minority language agentivity can diffuse to the majority. Although it cannot be ruled out, it remains very unlikely. Consequently, convergence will always happen in the direction of the minority language. And this is indeed what is happening between Arabic and Domari: they become more and more similar at all levels, but only Domari is moving towards Arabic.

In the realm of phonology, it was shown that Domari has kept a distinct inventory from Arabic, although convergence with Arabic is almost complete for short vowels. A possible consonantal imposition is found in Beirut/Damascus Domari where etymological /q/ is realized as /ʔ/, as in neighbouring Arabic dialects. As far as morphology is concerned, eligible candidates for imposition are the Kurdish diminutive \textit{{}-ək}, the Turkish conditional clitic \textit{sa} and superlative \textit{ān}. An evident case of imposition is the phenomenon that seems the most sensitive to dominance: so-called ``bilingual suppletion" \citep{Matras2012}. Bilingual suppletion in Northern Domari can be observed only in the dialect of Beirut/Damascus in the case of comparatives, and incipiently in the case of numerals. As far as syntax is concerned, cases of imposition are probably the transfer of Arabic auxiliaries and the negator \textit{mū}. The transfer of utterance modifiers such as fillers, adverbs, conjunctions and virtually all discourse structuring devices are so prone to replication in contact situations \citep{Matras1998} that it is difficult to assess the source of agentivity. Other features discussed in this paper, such as constituent order, the internal object and the impersonal construction are clear instances of convergence.

As noted above, the main direction of change in Domari is towards convergence with Arabic, as expected in case of absence of dominance. The dialect of Beirut/Damascus is the most convergent of all the Northern dialects, which in itself suggests that Arabic--Domari bilingualism is older in that variety. The Arabic component in Domari is largely uneven cross-dialectally and no overall statement about its nature can be made. The general picture that arises is that the impact of Arabic gradually increases from north to south, with the dialects of northern Syria and southern Turkey being the least Arabized, the Southern dialects spoken in Palestine and Jordan being the most influenced by Arabic, and the dialect of Beirut/Damascus exhibiting an intermediary stage. It was also shown that the main difference between Northern and Southern Domari as far as Arabic is concerned is the reluctance in Northern Domari to transfer Arabic inflections and the general tendency to favour the transfer of structures without substance.

\section*{Further reading}

For a general account of the Arabic component in all the varieties of Domari documented so far, see \citet{Herin2018}. The paper discusses the Arabic component in Southern and Northern dialects. This is the only paper that tackles extensively the issue of contact-induced change in Domari from a global perspective.

For a description of the Domari dialect of Aleppo, readers can refer to \citet{Herin2012}.

\citet{Herin2014Domari} identifies the grammatical features that make Northern Domari a coherent dialectal group.

\citet{Herin2016} investigates the full extent of variation in Domari as a whole, drawing on data from both Northern and Southern Domari.

Readers can refer to Matras (this volume) for a number of references relating to Jerusalem Domari.\ia{Matras, Yaron@Matras, Yaron}

\section*{Abbreviations}

\begin{tabularx}{.5\textwidth}{@{}lQ@{}}
\textsc{1, 2, 3} & 1st, 2nd, 3rd person \\
\textsc{abl} & ablative \\
\textsc{acc} & accusative \\
\textsc{ad} & adessive \\
\textsc{ben} & benefactive \\
\textsc{cmpr} & comparative \\
\textsc{com} & comitative \\
\textsc{cop} & copula \\
\textsc{def} & definite article \\
\textsc{dem} & demonstrative \\
\textsc{exs} & existential \\
\textsc{f} & feminine \\
\textsc{impf} & imperfect (prefix conjugation) \\
IPA & International Phonetic Alphabet\\
\textsc{in} & inessive \\
\textsc{ind} & indicative \\
\end{tabularx}%
\begin{tabularx}{.5\textwidth}{@{}lQ@{}}
\textsc{indf} & indefinite \\
\textsc{ipfv} & imperfective \\
\textsc{m} & masculine \\
\textsc{neg} & negation \\
NP & noun phrase \\
\textsc{obj} & object \\
\textsc{obl} & oblique \\
\textsc{pfv} & perfective \\
\textsc{pl} & plural \\
\textsc{prs} & present \\
\textsc{prf} & perfect (suffix conjugation) \\
\textsc{prog} & progressive \\
\textsc{pst} & past \\
\textsc{rel} & relative \\
\textsc{sbjv} & subjunctive \\
\textsc{sg} & singular \\
\textsc{supr} & superessive \\
\end{tabularx}%


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]\end{document}