\documentclass[output=paper,nonflat]{langsci/langscibook} 
\author{Bettina Leitner\affiliation{University of Vienna}}
\title{Khuzestan Arabic}
\abstract{Khuzestan Arabic is an Arabic variety spoken in the southwestern Iranian province of Khuzestan. It has been in contact with (Modern) Persian since the arrival of Arab tribes in the region before the rise of Islam. Persian is the socio-politically dominant language in the modern state of Iran and has influenced the grammar of Khuzestan Arabic on different levels. The present article discusses phenomena of contact-induced change in Khuzestan Arabic and considers their limiting factors.
}
\IfFileExists{../localcommands.tex}{
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle  

\section{Current state and historical development} 
\subsection{Historical development}
Arab settlement in Iran preceded the Arab destruction of the Sasanian empire with the rise of Islam. Various tribes, such as the Banū Tamīm, had settled in Khuzestan prior to the arrival of the Arab Muslim armies \citep[211]{Daniel1986}. In the centuries after the spread of Islam in the region, large groups of nomads from the Ḥanīfa, Tamīm, ʕAbd-al-Qays, and other tribes crossed the Persian Gulf and occupied some of the territories of southwestern Iran \citep[215]{Oberling1986}. The Kaʕb, still an important tribe in the area,\footnote{Cf. Oberling (\citeyear[218]{Oberling1986}) for an overview of the Arab tribes in Khuzestan.} settled there at the end of the sixteenth century \citep[216]{Oberling1986}. During the succeeding centuries many more tribes moved from southern Iraq into Khuzestan. This has led to a considerable increase of Arabic speakers in the region, which until 1925 was called Arabistan (see \citealt{Gazsi2011}: 1020; Gazsi, this volume).\ia{Gaszi, Dénes@Gaszi, Dénes} Today Khuzestan is one of the 31 provinces of the Islamic Republic of Iran, situated in the southwest, at the border with Iraq. 

There has been considerable movement to and from Iraq, to Kuwait, Bahrain, and Syria, and from villages into towns. Many of these migrations were a consequence of the Iran–Iraq war (1980–1988), but some were due to socio-economic reasons. The settlement of Persians in the region over the past decades \citep[1020]{Gazsi2011} is another important factor in its demographic history. From the early twentieth century on, Khuzestan has attracted international, especially British, interest because of its oil resources. 

\subsection{Current situation of Arabs in Khuzestan}

Information about the exact number of Arabic-speaking people in Iran, and especially in Khuzestan, is hard to find. Estimates in the 1960s of the Arabic-speaking population in Iran ranged from 200,000 to 650,000 \citep[216]{Oberling1986}. Today it is estimated that around 2 to 3 million Arabs live in Khuzestan (\citealt{MatrasShabibi2007}: 137; \citealt{Gazsi2011}: 1020). 

Many Arabs and Persians living in Khuzestan work in the sugar cane or oil industries, but few hold white-collar or managerial positions (\citealt{DePlanhol1986}: 55–56). This is one of the reasons why many Arabs in Khuzestan feel strongly disadvantaged in society and politics in comparison to their Persian neighbours.\footnote{The most common Khuzestan Arabic terms for the Persian people and their language are \textit{ʕaǧam} ‘Persian' (people and language; lit. `non-Arab'), and \textit{əl-ǧamāʕa} ‘Persians’ (lit. `group of people'). Both are often used pejoratively.} 

\section{Language contact in Khuzestan}

Currently, the main and most influential language in contact with Khuzestan Arabic is the Western Iranian language Persian. Among the other (partly historically) influential languages in the region the most prominent are English, Turkish/Ottoman (cf. \citealt{Ingham2005}), and Aramaic (see Procházka, this volume).\ia{Procházka, Stephan@Procházka, Stephan}

Persian and different forms of Arabic share a long history of contact in the region of Khuzestan, implying a long exchange of language material in both directions. 

Khuzestan Arabic belongs to the Bedouin-type south Mesopotamian \textit{gələt}-dialects.\footnote{There is as yet no comprehensive grammar of the dialects of Khuzestan. The main source of information on these dialects is the collection of data made in the 1960s by the arabist and linguist Bruce Ingham (\citeyear{Ingham1973,Ingham1976,Ingham2011khuz}). The article by Yaron Matras and Maryam Shabibi, “Grammatical borrowing in Khuzistani Arabic” \citep{MatrasShabibi2007}, is based on Shabibi’s unpublished dissertation “Contact-induced grammatical changes in Khuzestani Arabic” \citep{Shabibi2006}.} Therefore, it shows great similarity to Iraqi dialects such as Baṣra Arabic, as well as to other dialects in the Gulf, such as Bedouin Bahraini Arabic – that is, the Arabic spoken by the Sunni Arab population descended from Najd. 

The dialects of Khuzestan can be considered “peripheral” dialects of Arabic because they are spoken in a country where Arabic is not the language of the majority population and is not used in education and administration. Therefore, there is practically no influence of Modern Standard Arabic. However, because it shares a long geographically-open border with Iraq, Khuzestan is not isolated from the Arabic-speaking world. Moreover, since around 2000 it has had access to Arabic news, soaps, etc. via satellite TV. Intra-Arabic contact is limited to the linguistically very similar (southern) Iraqi dialects\footnote{Khuzestan Arabic is often differentiated from its neighboring Iraqi dialects by the number of Persian borrowings that are employed \citep[1020]{Gazsi2011}. Although the greatest influence has occurred in lexicon, Persian influence also extends to grammar (see below).} through, for example, religious visits to Kerbala.

Persian is the only official language in Iran, and the only language used in education, and is sociolinguistically and culturally dominant, especially in the domains of business and administration. Persian consequently enjoys high prestige in society. For Persian speakers, and sometimes also for Khuzestan Arabic speakers, the Khuzestan Arabic varieties have very low prestige and are not associated with the highly prestigious Arabic of the Quran, which is taught at schools. Khuzestan Arabic speakers who acquire Khuzestan Arabic as a first language usually acquire Persian at school. Later, the opportunities for Khuzestan Arabic speakers to use Persian are restricted to certain social settings outside the family, e.g. school, work (employment in a large company would probably require communication in Persian), contact with Persian friends, or through the Persian media.

Accordingly, the command of Persian or the degree of bilingualism among Khuzestan Arabic speakers varies greatly due to such factors as level of education, affiliation, age, gender, and urban or rural environment. The older generation and women have far less access to education and jobs and consequently less contact with people outside the family, which implies less exposure to contact situations and a lower degree of bilingualism. Among some members of the younger generation we may notice a certain intentional reinforcement of Arabic words alongside a resistance to recognizable Persian lexical borrowings plus a preference for the Arabic over the Persian names for the cities in Khuzestan. This of course goes in line with nationalist ideas and the separatist movement taking place in present-day Khuzestan and also shows the impact of intentionality in language contact situations. 

In sum, one might find very different degrees of Persian influence among the speakers of Khuzestan Arabic (cf. \citealt{MatrasShabibi2007}: 147). For that reason, all statements on Persian–Khuzestan Arabic contact phenomena must be seen in relation to the above factors, which are decisive for any speaker’s command of Persian. 


\section{Contact-induced changes in Khuzestan Arabic} 
\subsection{General remarks\footnote{\textnormal{The data used for the present analysis was collected mainly in Aḥwāz, Muḥammara (Khorramshahr), Ḥamīdiyye and Ḫafaǧiyye (Susangerd) in 2016. The male and female informants were bilingual as well as monolingual Khuzestan Arabic speakers from 25 to over 70 years old.}}}
The main aim of the present chapter is to highlight the most striking phenomena and trends in Khuzestan Arabic language change due to contact with Persian.

All phenomena of contact-induced change in Khuzestan Arabic can be considered as transfer of patterns or matter\footnote{Sakel (\citeyear[15]{Sakel2007}) defines matter replication as the replication of “morphological material and its phonological shape”.} from the source language (SL) Persian to the recipient language (RL) Khuzestan Arabic under RL agentivity (i.e. borrowing rather than imposition).{} The agents of transfer are cognitively dominant in the RL Khuzestan Arabic, the agents’ L1. Even though Persian is generally acquired during childhood and today is spoken by most speakers, it usually is the speakers’ L2. Cases of convergence (cf. \citealt{Lucas2015}: 530–531) are possible in the present contact situation among speakers with a very high (L1-like) command of Persian, for example university students. But of course it is hard to draw an exact line between L1 and L2 proficiency and thus between convergence and borrowing (cf. \citealt{Lucas2015}: 531). 



\subsection{Phonology}
As in other Bedouin Arabic dialects, the presence of the phonemes /č/ and /g/ is ultimately the result of internal development from original *k and *q, rather than borrowing from Persian (see Procházka, this volume).\ia{Procházka, Stephan@Procházka, Stephan}

The phoneme /p/, e.g. \textit{perde} ‘curtain’ < Pers \textit{parde}, is also common in all Iraqi dialects and probably emerged in this region due to contact with Persian and Kurdish (see Procházka, this volume).\ia{Procházka, Stephan@Procházka, Stephan} 

An interesting phonological feature of Khuzestan Arabic is that /ɣ/ often reflects etymological *q,\footnote{This phenomenon is also documented for the Arabic dialects of Kuwait, Qatar, and the United Arabic Emirates (\citealt{Holes2016}: 54, fn. 5).} which is otherwise realized as /g/ and /ǧ/. It is most likely that the shift /ɣ/ < *q first occurred in Persian borrowings ultimately of Arabic origin, e.g. \textit{ɣisma} ‘part, section’, \textit{taṣdīɣ} ‘driving licence’, \textit{yifruɣ} ‘it differs’, \textit{taɣrīban} ‘approximately’, and \textit{ɣidra} ‘power’. This feature is most probably a transfer from Persian,\footnote{Contrast Holes (\citeyear{Holes2016}: 53–54), who explains the /ɣ/–/q/ merger among the Najd-descendent Bahraini Arabic speakers as internal development.} in which both *q and *ɣ in Arabic loanwords are always pronounced /ɣ/ (\citealt{MatrasShabibi2007}: 138).\footnote{In Modern Persian the phoneme /ɣ/ has two allophones, [ɢ] and [ɣ] (\citealt{Majidi1986}: 58–60).} Later, this phonological change further affected lexemes which have no cognate form in Persian, e.g. \textit{baɣra} `cow', a borrowing from Modern Standard Arabic (the KhA dialectal form being \textit{hāyša} `cow'). There are, however, certain lexemes which are not affected by this rule, e.g. \textit{gāl} `he said'. Other lexemes show free variation in the pronunciation of /g/, e.g. \textit{gabul/ɣabul} `formerly, before'.

Lexical borrowings are often adapted to Arabic phonology. For example, Khuzestan Arabic speakers of the older generation usually pronounce the phoneme /p/ as [b]; and /s/ in Persian loanwords is sometimes pronounced [š], e.g. \textit{berde} ‘curtain’ < Pers \textit{parde}, and \textit{šebzi} ‘herb stew’ < Pers \textit{sabzi}. 

Negative structures bear stress on the first syllable,\footnote{Ingham (\citeyear[724]{Ingham1991}) describes this phenomenon also for Khuzestan Arabic \textit{wh}-interrogatives and prepositions.} e.g. KhA \textit{mā́} \textit{arūḥ} ‘I don’t go’. This is a feature shared with some Persian and Turkish varieties and other North East Arabian dialects (\citealt{Ingham2005}: 178–179). This common phonological characteristic therefore seems to be a Sprachbund phenomenon of the Mesopotamian region, which reflects the long history of contact and migration across language boundaries due to trade, war, shared cultural practices, nomadism, etc. (\citealt{Winford2003}: 70–74). Though the directions and mechanisms of borrowing within the languages of a Sprachbund are often hard to categorize \citep[74]{Winford2003}, we can probably assume that Khuzestan Arabic, being spoken by a minority group, has borrowed and adapted this phonological stress pattern under RL agentivity.



\subsection{Morphosyntax}
\subsubsection{Replication of Persian phrasal verbs}

The replication of phrasal verbs is a contact phenomenon also found in the Arabic varieties of Turkey (\citealt{Grigore2007book}: 157–159; and Procházka, this volume).\ia{Procházka, Stephan@Procházka, Stephan} As shown in examples (\ref{muxx})--(\ref{tala}), Khuzestan Arabic replicates Persian phrasal verbs by substituting the Persian light verbs with Khuzestan Arabic equivalents and directly replicating the Persian nouns (cf. \citealt{MatrasShabibi2007}: 142). The noun in example (\ref{muxx}) is Arabic in its origins but its usage in a phrasal verb construction with a new meaning is a Persian innovation.

\ea\label{ex:key:} \label{muxx}
\ea Aḥwāz, Khuzestan, male, 26 years (own data)\\
\gll ṭəgg muḫḫ\\
     hit.\textsc{prf}.3\textsc{sg}.\textsc{m} brain\\ 
\ex{Persian}\\
\gll muḫḫ zadan\\
     brain hit.\textsc{inf} \\
\glt ‘to brainwash/convince someone’\footnote{All Persian translations are given in the modern spoken (Tehrani) variety of Persian.} 
\z
\z

\ea\label{ex:key:} \label{irad}
\ea{Aḥwāz, Khuzestan, male, 39 years (own data)}\\
\gll kað̣ð̣ īrād\\
     take.\textsc{prf}.3\textsc{sg}.\textsc{m} nagging\\ 
\ex{Persian}\\
\gll īrād gereftan\\
     nagging take.\textsc{inf}\\
\glt ‘to pick at someone’ 
\z\z

 As examples (\ref{omade}) and (\ref{tala}) show, the Persian nouns are sometimes adapted morphophonologically.

\ea\label{ex:key:} \label{omade}
\ea{Aḥwāz, Khuzestan, male, 50 years (own data)}\\
\gll sawwa ʔōmāde\\
     make.\textsc{prf}.3\textsc{sg}.\textsc{m} ready\\ 
\ex{Persian}\\
\gll āmāde kardan \\
     ready make.\textsc{inf}\\
\glt ‘to prepare sth.’
\z\z

\ea\label{ex:key:} \label{tala}
\ea{Aḥwāz, Khuzestan, male, 26 years (own data)}\\
\gll ṭalaʕ ɣabūli\footnotemark \\
     emerge.\textsc{prf}.3\textsc{sg}.\textsc{m} acceptance\\
\ex{Persian}\\
\gll ɣabūl šodan\\
     acceptance become.\textsc{inf} \\
\glt ‘to pass (an exam), be accepted’
\z\z
\footnotetext{The final \textit{-i} in \textit{ɣabūli} probably originates from the Persian indefiniteness marker \textit{-i} (see \citealt{Majidi1990}: 309–314) and has become part of this word in Khuzestan Arabic, so that \textit{ɣabūli} is monomorphemic.}
The pattern for phrasal verbs – transferred into the RL Khuzestan Arabic under RL agentivity – provides Khuzestan Arabic with an easy way to convert foreign nouns into verbs.

As illustrated in examples \REF{taggi} and \REF{tamate}, the pattern is adapted according to Arabic syntactic rules: (i) the verb is moved into the initial position; and (ii) a direct object is introduced between verb and nominal element (post-verbally). In Persian, however, the verb always remains in final position following the nominal element and a direct object would be introduced before both elements (see e.g. \citealt{Majidi1990}: 447–448).

\ea\label{ex:key:} \label{taggi}
\ea{Aḥwāz, Khuzestan, male, 26 years (own data)}\\
\gll ṭəggi ɣandart-i wāks\\
     hit.\textsc{imp.2sg.f} shoe-\textsc{obl}.1\textsc{sg} wax\\
 
\ex{Persian}\\
\gll kafš-am-o wāks be-zan\\
     shoe-\textsc{obl.}1\textsc{sg-acc} wax \textsc{imp-}hit\textsc{.prs}\\
\glt ‘Polish my shoes!’
\z\z

\ea\label{ex:key:} \label{tamate}
\ea{Aḥwāz, Khuzestan, female, 35 years (own data)}\\
\gll yṭəggūn əṭ-ṭamāṭe rande\\
     hit.\textsc{impf.3pl.m} \textsc{def}{}-tomato grater \\
 
\ex{Persian}\\
\gll gūǧe\_farang-i rō rande mī-zanan\\
     tomato-\textsc{indf} \textsc{acc} grater \textsc{prog}-hit\textsc{.prs.3pl.m} \\
\glt ‘They grate some tomato.’
%\check glosses 
\z\z

This structure has become productive in Khuzestan Arabic. For example, in the phrasal verb \textit{ṭəgg} \textit{dabbe} ‘to cheat’ (lit. ‘to hit a water canister’) both the verb and noun are taken from Khuzestan Arabic and only the construction’s syntactic pattern is taken from Persian. 

\subsection{Syntax}
\subsubsection{Definiteness marking}

Matras \& Shabibi (\citeyear{MatrasShabibi2007}: 141–142) see Khuzestan Arabic relative clauses without definite heads as evidence for the decline of overt definiteness marking in Khuzestan Arabic, based on the Persian model with generally unmarked definiteness, e.g. \textit{mara} \textit{lli} \textit{šiftū-ha} \textit{ḫābarat} ‘The woman that you saw called’ (\citeyear{MatrasShabibi2007}: 142). However, this pattern is also documented in Arabic dialects which have had no contact with Persian (\citealt{Pat-El2017}: 454–455; cf. \citealt{Procházka2018}: 269).

Matras \& Shabibi (\citeyear[140]{MatrasShabibi2007}) further postulate that the Persian \textit{ezāfe} pattern in adjectival attribution is replicated in Khuzestan Arabic.\footnote{See e.g. Ahadi (\citeyear[103–109]{Ahadi2001}) for the usage of the Persian \textit{ezāfe}.} According to their theory, the construct state marker -\textit{t} (with an indefinite head) and/or the definite article (of the attribute) are reanalysed as markers of attribution matching the Persian  \textit{ezāfe} marker \textit{-(y)e}, as in \REF{island}. 

\ea\label{island}
\ea
{KhA (\citealt{MatrasShabibi2007}: 140)}\\
\gll ǧazīra-t l-ḫað̣ra \\
     island-\textsc{con} \textsc{def}-green.\textsc{f}\\
 
\ex{Persian}\\
\gll ǧazīre-ye sabz\\
     island-\textsc{ez} green\\
\glt ‘the green island’ 
\z\z

However, this pattern is also observed in other modern Arabic dialects which have not been exposed to Persian influence as well as in older forms of Arabic.\footnote{See Pat-El (\citeyear{Pat-El2017}: 445–449) for numerous examples from different modern Arabic dialects, Middle Arabic, Pre-Islamic Arabic, Quran Arabic, and other Central Semitic languages. See also Retsö (\citeyear{Retsö2009}: especially 21–22) and Procházka (\citeyear{Procházka2018}: 267–269), who also proves that this is an old feature already found in Old Arabic and points out that it is mainly found among dialects which are spoken in regions with no or only marginal influence from Modern Standard Arabic.} 

Consequently, it is highly unlikely that this phenomenon has developed due to Persian influence, although it cannot be ruled out that contact with Persian has fostered the preservation of this apparently old feature. 


\subsubsection{Word order changes} \label{woc}

Khuzestan Arabic shows no changes due to contact in basic word order.\footnote{Ingham (\citeyear[715]{Ingham1991}) states that in Khuzestan Arabic neither VSO nor SVO word order is particularly dominant. Matras \& Shabibi (\citeyear[147]{MatrasShabibi2007}) postulate that the usage of OV order in Khuzestan Arabic is increasing as “the beginning of a shift in word order” on the basis of the Persian type, where OV prevails. In both of their examples the objects are topicalized (with pronominal resumption), which is a common phenomenon in spoken Arabic (\citealt{Brustad2000}: 330–333; 349), and as such not obviously the result of Persian influence.}  The only attested word order changes concern the position of the verbs \textit{čān} ‘to be’ and \textit{ṣār} ‘to become’, both of which can appear in final position as an unmarked construction. This sentence-final position in no case functions as the default, and is in fact less frequent than its non-final position.\footnote{In my data, \textit{čān} appears 23 of 152 times in sentence-final position, \textit{ṣār} 11 of 165 times. The additional examples are taken from my questionnaire.} \textit{čān} or \textit{ṣār} in final position are never stressed.

The sentence-final position of \textit{čān} or \textit{ṣār} (see examples \REF{port}--\REF{cold}) is likely a pattern replication of the Persian model, i.e. sentences with final \textit{būdan} ‘to be’ or \textit{šodan} ‘to become’.

\ea\label{port}
\ea
{ʕAbbādān, Khuzestan, male, 35 years (own data)}\\
\gll šuɣul-hum b-əl-bandar čān\\
     work-\textsc{3pl}.\textsc{m} in-\textsc{def}-port be.\textsc{prf.3sg.m}\\
 
\ex{Persian}\\
\gll kār-ešūn tū-ye bandar būd\\
     job-\textsc{obl}.3\textsc{pl} in-\textsc{ez} port be.\textsc{pst.3sg}\\
\glt ‘Their job was at the port.’
\z\z

\ea
\ea {Muḥammara, Khuzestan, male, 30 years
 (own data)}\\
\gll əǧdād-i mallāk-īn čānaw\\
     grandparents-\textsc{obl.1sg} owner-\textsc{pl} be.\textsc{prf.3pl.m}\\
 
\ex{Persian}\\
\gll pedarbozorgā-m mālek būdan\\
     grandparents-\textsc{obl.1sg} owner be.\textsc{pst.3pl}\\
\glt ‘My grandparents were owners [of land].’
\z\z

\ea\label{cold}
\ea {Aḥwāz, Khuzestan, female, 40 years (own data)}\\
\gll hassa šway l-māy bārəd ṣār\\
     now a\_bit \textsc{def}{}-water cold become.\textsc{prf.3sg.m}\\
 
\ex{Persian}\\
\gll alʔān yekam ʔāb sard šod\\
     now a\_bit water cold become.\textsc{pst}.\textsc{3sg}\\
\glt ‘The water has become a bit cold now.’
\z\z

The next example might show a tendency to use a present-tense copula with human subjects, expressed with the verb \textit{ṣār} ‘to become’: 
\ea
\ea {Aḥwāz, Khuzestan, female, 35 years
 (own data)}\\
\gll əhya mart uḫū-y əṣṣīr\\
     3\textsc{sg.f} wife brother-\textsc{obl.1sg} \textsc{cop.impf.3sg.f}\\
 
\ex{Persian}\\
\gll ūn zan dādāš-am-e\\
     3\textsc{sg} wife brother-\textsc{obl.1sg}-\textsc{cop.prs.3sg}\\
\glt ‘She is the wife of my brother.’
\z\z


In the Khuzestan Arabic construction for pluperfect tense, \textit{čān} can also appear in sentence-final position, after the active participle. This construction, although not very frequent, is very likely a direct transfer of the Persian structure, in which the auxiliary \textit{būdan} also follows the participle.\footnote{Matras and Shabibi (\citeyear{MatrasShabibi2007}: 142–143) describe the use of this construction as a change in the Khuzestan Arabic tense system. However, the pattern \textit{kān} + active participle is also commonly used in other Arabic dialects to express pluperfect meaning or to describe completed actions which have an impact on the present, see for example Denz (\citeyear{Denz1971}: 92–94; 115–116) for Iraqi (Kwayriš) and Grotzfeld (\citeyear[88]{Grotzfeld1965}) for Syrian Arabic.} 

\ea \ea {Aḥwāz, Khuzestan, male, 26 years (own data)}\\
\gll lamman əyēna l-əl-bīət, əhma mākl-īn čānaw\\
     when come.\textsc{prf.1pl} to-\textsc{def}-house 3\textsc{pl}.\textsc{m} eat.\textsc{ptcp-pl.m} be.\textsc{prf}.\textsc{3pl.m}\\
 
\ex{Persian}\\
\gll vaɣti-ke mā bargaštīm ḫūne, ūnhā ɣazā-ro ḫorde būdan  \\
     when-\textsc{rel} \textsc{1pl} come\_back.\textsc{pst}.\textsc{1pl} home 3\textsc{pl} food-\textsc{acc} eat.\textsc{ptcp} be.\textsc{pst}.\textsc{3pl}\\
\glt ‘When we came home, they had (already) eaten.’
\z
\z

This word order change has probably been triggered by the high frequency in speech of Persian sentences with forms of \textit{būdan} in final position. Lucas (\citeyear[295]{Lucas2012}) explains the usage of foreign patterns as the result of the human cognitive tendency to minimize the high processing efforts associated with the extensive use of two languages.\footnote{{Connections between units of a neural network associated with certain syntactic patterns can be strengthened from repeated exposure to and use of that pattern} {\citep[291]{Lucas2012}. Hence, the employment of a Persian syntactic structure in Khuzestan Arabic needs less processing effort because the same strengthened neural network is activated.}}

\textit{čān} is also used in sentence-final positions after the main verb in the imperfect in Khuzestan Arabic constructions expressing the continuous past. In spoken Persian, the continuous past is formed without a sentence-final \textit{būdan}.\footnote{The Persian continuous past is formed with the particle \textit{mī} prefixed to the simple past of the respective main verb and can (for the progressive form) be preceded by the simple past of \textit{dāštan} ‘to have’: e.g. (\textit{dāšt}) \textit{mī-raft} ‘he was going’ (\citealt{Majidi1990}: 232, 235).} This case is not a direct transfer of the Persian pattern, but perhaps a construction analogous to the pluperfect and other Persian forms with \textit{būdan} in final position. 

\ea
\ea {Aḥwāz, Khuzestan, male, 55 years (own data)}\\
\gll hāda ham mən zuɣur yəštəɣəl čān  \\
     \textsc{dem}.\textsc{sg}.\textsc{m} also from childhood work.\textsc{impf.3sg.m} be.\textsc{prf.3sg.m}\\
 
\ex{Persian}\\
\gll īn ham az kūdeki kār mī-kard \\
     \textsc{dem.sg} also from childhood work \textsc{prog}{}-do.\textsc{pst.3sg}\\
\glt ‘This one has also been working from childhood on.’
\z\z

Example \REF{mother} shows both syntactic variants in one sentence, i.e. \textit{čān} before and after the main verb.

\ea\label{mother} \ea {Muḥammara, Khuzestan, female, 40 years (own data)}\\
\gll umm-i čānat tətḥaǧǧab, eh, əb-zamān əš-šāh, bass tətbawwaš čānat\\
     mother-\textsc{obl.1sg} be\textsc{.prf.3sg.f} veil.\textsc{impf.3sg.f} yes in-time \textsc{def}{}-shah only veil.\textsc{impf.3sg.f} be\textsc{.prf.3sg.f}\\
 
\ex{Persian}\\
\gll mādar-am (dāšt) naqāb mī-zad, āre, dar zamān-e šāh, hamīše naqāb mī-zad \\
     mother-\textsc{1sg} (have.\textsc{pst.3sg}) veil \textsc{prog}-hit.\textsc{pst.3sg} yes in time-\textsc{ez} shah always veil \textsc{prog}-hit.\textsc{pst.3sg}\\
\glt  ‘My mother was veiling her face (with a \textit{būšiyye}),\footnote{\textit{būšiyye} or \textit{pūšiyye} ‘veil’ is also documented for Iraqi Arabic (\citealt{WoodheadBeene1967}: 53).} yes, during the times of the shah, she was always veiling her face.’ 
\z\z

Because all the above examples equally work with \textit{čān/ṣār} in non-final position, this process of word-order-related pattern replication in Khuzestan Arabic is still ongoing. Indeed, all informants, when asked for the correct structure in the above examples, preferred the verb \textit{čān} in non-final position.\footnote{My informants from Baghdad considered all constructions with \textit{čān} in final position to be wrong. However, this structure is used in Baṣra Arabic (Qasim Hassan, personal communication, January 2018).}

Lucas (\citeyear{Lucas2015}: 530–531) explains the basic word order changes (from VSO to SOV) in Bukhara Arabic (cf. \citealt{Ratcliffe2005}: 143–144; and \citealt{Versteegh2010}: 639) as a result of convergence with Uzbek.\footnote{Lucas (\citeyear[525]{Lucas2015}) defines convergence as changes made to a language under the agentivity of speakers who are native speakers of both the SL and the RL.} Although a clear division between convergence and borrowing is hard to make, I consider the contact-induced word order changes that occur in Khuzestan Arabic to be instances of borrowing because most speakers are clearly native speakers of, and therefore dominant in, Khuzestan Arabic only.


\subsubsection{\textit{ḫōš} preceding verbs and nouns} \label{xosh}

In Persian, \textit{ḫōš} ‘good, well’ is used as a prefixed (lexicalized) element preceding some nouns and verbs to coin compound adjectives, nouns, and verbs (\citealt{Majidi1990}: 411, 413): e.g. Pers \textit{ḫōš-andām} ‘handsome’ (< \textit{andām} ‘shape; body’), \textit{ḫōš-nevīs} ‘calligrapher’ (< present stem \textit{nevīs-} ‘to write’), and \textit{ḫōš-āmadan} ‘to please’ (< \textit{āmadan} ‘to come’).

Khuzestan Arabic has borrowed some of these Persian compound adjectives: e.g. \textit{ḫōš-bū} ‘nice-smelling’ (< Pers \textit{bū} ‘smell, scent’), \textit{ḫōš-tīp} ‘handsome’ (< Pers \textit{tīp} ‘type’), and \textit{ḫōš-aḫlāq} ‘(with) good manners’ (< Pers \textit{aḫlāq} ‘decency; ethics, morality’, pl. of \textit{ḫolq} ‘character, nature’). However, in Khuzestan Arabic the use of the Persian element \textit{ḫōš} has been further developed. It is productively used as an attributive adjective preceding nouns, but not agreeing in gender or number with them, e.g. \textit{ḫōš} \textit{walad} ‘a good boy’, \textit{ḫōš} \textit{əbnayya} ‘a good girl’, \textit{ḫōš} \textit{banāt} ‘good girls’, \textit{ḫōš} \textit{əwlād} ‘good kids’.\footnote{This construction is also found in Iraqi Arabic (cf. \citealt{Erwin1963}: 256), which might prove that the element \textit{ḫōš} is an older borrowing.}



\subsection{Lexicon}
\subsubsection{Lexical transfer}

The greatest influence from Persian on Khuzestan Arabic has occurred in lexicon. Many Persian lexemes were borrowed generations ago. The most frequently borrowed elements are nouns denoting cultural or technological innovations which have filled lexical gaps in the RL Khuzestan Arabic. Verbs, adverbs, adjectives, and many discourse particles have also been borrowed from the SL Persian.

The majority of the examples below are cases of transfer of morpho-phonological material (matter) and semantic meaning (pattern) under RL agentivity. 

Many of the Persian borrowings have been phonologically and morphologically integrated into the RL. For instance, for many borrowed Persian nouns Arabic internal plural forms are created, e.g. \textit{ḫətākīr} ‘ball-point pens’ (sg. \textit{ḫətkār} < Pers \textit{ḫod-kār} ‘ball-point pen’), or \textit{banādər} ‘ports’ (sg. \textit{bandar} < Pers \textit{bandar} ‘port’).

Again, the borrowing of foreign (L2) elements into the speakers’ L1 might be explained by the human cognitive tendency to minimize the processing effort in lexical selection between two languages (\citealt{Lucas2012}: 291; see §\ref{woc}). So if a certain Persian word is frequently used and often heard (for example at school), the connections of a neural network associated with this word are strengthened \citep[291]{Lucas2012}, which makes it easier to employ the word in one’s L1. 


\subsubsection{Semantic fields}

The following illustrative list of Persian loans in Khuzestan Arabic shows the most important semantic fields of lexical borrowing.

\begin{itemize} 
\item Administration/military: \textit{farmāndāri} ‘governorship’ < Pers \textit{farmāndāri}; \textit{čārra} ‘crossroad’ < Pers \textit{čahār-rāh}; \textit{sarbāz}/\textit{šarbāz} ‘soldier’ < Pers \textit{sarbāz}.
\item Agriculture: \textit{ʕalafkoš} ‘pesticide' (lit. weed-killer) < Pers \textit{ʕalaf-koš}; \textit{kūd} ‘dung’ < Pers \textit{kūd}. 
\item Dress/textiles: \textit{dāmen} ‘skirt’ < Pers \textit{dāman}; \textit{šīəla} ‘head covering’ < Pers \textit{šāl} ‘Kashmir shawl’ \citep[174]{Ingham2005}.
\item Education:  \textit{klāṣ} ‘class, grade’ < Pers \textit{kelās}; \textit{ḫətkār} ‘biro’ < Pers \textit{ḫod-kār}; \textit{dānišga} ‘university’\textit{<} Pers \textit{dānišgāh}.
\item Food: \textit{ǧaʕfari} ‘parsley’ < Pers \textit{ǧaʔfari}; \textit{češmeš} ‘raisins’ < Pers \textit{kešmeš}; \textit{šalɣam} ‘turnip’ < Pers \textit{šalɣam}; \textit{serke} ‘vinegar’ < Pers \textit{serke}.
\item Material culture: \textit{šīše} ‘bottle’ < Pers \textit{šīše}; \textit{ǧām} ‘(window) glass’ < Pers \textit{ǧām}; \textit{tīɣe} ‘blade’ < Pers \textit{tīɣe}; \textit{yəḫčāle} ‘refrigerator’ < Pers \textit{yahčāl}; \textit{sīm} \textit{buksel} ‘towrope’ < Pers \textit{sīm-e} \textit{boksol}; \textit{perde}/\textit{berde} ‘curtain’ < \textit{parde}; \textit{gīre} ‘hair barrette’ < Pers \textit{gīre-ye} \textit{sar/mūy}; \textit{mīz} ‘table’ < Pers \textit{mīz}; \textit{darīše} ‘window’ < Pers \textit{darīče}; \textit{pənǧara} ‘window’ < Pers \textit{panǧare}.
\item Other: \textit{ɣīme} ‘price’ < Pers \textit{ɣīmat}; \textit{bandar} ‘port’ < Pers \textit{bandar}; \textit{nāmard} ‘brute’ < Pers \textit{nāmard}.
\end{itemize}

Some items ultimately of Arabic origin have been re-borrowed into Khuzestan Arabic from Persian, preserving the Persian meaning, e.g. KhA \textit{bərɣi} ‘electronic’ < Pers \textit{barɣ} ‘electricity; lightning’ < Arabic \textit{barq} ‘lightning’. 


\subsubsection{Verbs and adverbs} 

Khuzestan Arabic verbs and adverbs resulting from language contact are always morphologically integrated. These are either directly borrowed Persian verbs, e.g. \textit{bannad} ‘to close (e.g. the tap)’ < Pers imperfect and present stem \textit{band-} ‘close’;\footnote{Also common in the Gulf region and in Yemen (\citealt{BehnstedtWoidich2014}: 290).} \textit{gayyər} ‘to get stuck’ <  Pers \textit{gīr} \textit{kardan} ‘to get stuck’; \textit{ʕamm{әr}} ‘to repair’ < Pers \textit{taʕmīr} \textit{kardan} ‘to repair’; \textit{čass{әb}} ‘to glue’ < Pers \textit{časb} \textit{zadan} ‘to glue’; \textit{gəzar} ‘to pass (time)’ < Pers present stem \textit{gozar-} ‘to pass (time)’ (see example \REF{ca} below);\footnote{The verb \textit{gəzar} is used only in phrases that refer to the “passing by” of life.} or Persian nouns turned into Khuzestan Arabic (ad)verbs, e.g. \textit{əb-zūr} ‘by force’ < Pers \textit{zūr} ‘power; violence; force’; and \textit{zaḥəm} ‘to bother’ (transitive) < Pers \textit{zahmat} ‘trouble, bother’ (see examples \REF{zah} and \REF{mumkin} below).\footnote{Khuzestan Arabic \textit{zaḥme} is also used as a noun for a rebuke, e.g. \textit{zaḥme} \textit{ʕalīək!} ‘Shame on you!’, which would be expressed in a different way in Persian: \textit{ḫeǧālat} \textit{nemi} \textit{keši?} ‘Shame on you!' (lit. ‘Are you not ashamed?’).}

\ea
{Aḥwāz, Khuzestan, male, 26 years (own data)}\\ \label{ca}
\gll čā hāy əl-ḥayāt lō la? təgzar baʕad, təmši\\
     \textsc{dm} \textsc{dem}.\textsc{f} \textsc{def}{}-life or no pass.\textsc{impf}.\textsc{3sg.f} after\_all go.\textsc{impf.3sg.f}\\
\glt ‘See, that is how life is, right? It passes by (quickly), it goes.’
\z

\ea \label{zah}
{Aḥwāz, Khuzestan, male, 26 years (own data)}\\
\gll zaḥmīət-kum, ʕafwan\\
     bother.\textsc{prf}.\textsc{1sg}{}-\textsc{2pl.m} sorry\\
\glt ‘Sorry, I must have bothered you/caused you trouble.’\footnote{A phrase often used when leaving, for example after an invitation for dinner.}
\z

\ea \label{mumkin}
{KhA, Aḥwāz (own data)}\\
\gll mumkin azaḥm-ək əb-šuɣla\\
     possible bother.\textsc{impf}.\textsc{1sg}{}-\textsc{2sg.m} with-issue\\
\glt ‘May I bother you with something/ask you a favour?’
\z


\subsubsection{Discourse elements}

A range of Persian discourse elements have been borrowed by Khuzestan Arabic (cf. \citealt{MatrasShabibi2007}: 143–145),\footnote{Matras \& Shabibi (\citeyear[144]{MatrasShabibi2007}) claim that the Persian conjunctions \textit{agarče} and \textit{bāīnke}, both meaning ‘although/even though’, and the Persian factual complementizer \textit{ke} ‘that’ have also been borrowed by Khuzestan Arabic. However, I have found no evidence for their usage in my data.} e.g. KhA \textit{ham}/\textit{hamme} ‘also, as well’ < Pers \textit{ham} and KhA \textit{ham}…\textit{ham} ‘(both)…and’ < Pers \textit{ham}…\textit{ham};\footnote{This discourse element is also known for Iraq \citep[36]{Malaika1963} and, like KhA \textit{hast}/\textit{hassət} ‘there is’ < Pers \textit{hast} (\citealt{Ingham1973}: 25, fn.27), is probably an older borrowing.} and KhA \textit{hīč} ‘nothing; no(t)… at all’ < Pers \textit{hīč}.\footnote{Shabibi (\citeyear{Shabibi2006}: 176–177) derives KhA \textit{balkət} ‘maybe, hopefully’ from Pers \textit{balke} \textit{ham}, which can mean ‘maybe’ . A Turkish origin of this word seems more likely: cf. Aksoy (\citeyear[620]{Aksoy1963}) for the existence of \textit{belke}/\textit{belkit} in Eastern Turkish dialects. Malaika (\citeyear[35]{Malaika1963}) also derives the Baghdadi Arabic \textit{belki} ‘rather, maybe’ from Turkish, as does Seeger (\citeyear[28]{Seeger2009}) for \textit{balki,} \textit{balkīš,} \textit{balkin} ‘maybe; possibly; probably’ in Ramallah Arabic.} 

The Khuzestan Arabic discourse elements \textit{ḫō}/\textit{ḫōš} ‘well; okay’ < Pers \textit{ḫō(b)}/\textit{ḫōš} are often used phrase-initially, (\ref{okay}).\footnote{According to my informants and data, the form \textit{ḫōb} is not used in Khuzestan Arabic (contrast \citealt{MatrasShabibi2007}: 143).} They are of Persian origin, but have partly adopted a different form and function in Khuzestan Arabic.\footnote{In Persian, \textit{ḫō(b)} is a discourse particle and simple adjective, and \textit{ḫōš} just an adjective (see §\ref{xosh}; \citealt{Shabibi2006}: 160). Thus the Persian adjective \textit{ḫōš} has been desemanticized in Khuzestan Arabic to function as a discourse particle with the meaning ‘well, okay’ \citep[160]{Shabibi2006}.}

\ea\label{okay}
{Aḥwāz, Khuzestan, male, 55 years (own data)}\\
\gll ḫōš, š-ʕəd-na, taʕay əhna baba\\
     \textsc{dm} what-at-\textsc{1pl} come.\textsc{imp}.\textsc{sg.}\textsc{f} here father\\
\glt ‘Okay, what (else) do we have, come here, dear!’
\z

Both \textit{ḫō} and \textit{ḫōš} are also often used in stories following the verb \textit{gāl} ‘to say’.

\ea\label{ex:key:}
{Aḥwāz/Fəllāḥiyya, Khuzestan, female, 50 years (own data)}\\
\gll lamman ɣada mən ʕəd-hum, gāl-la ḫō, hāy ər-rummānāt š-asawwi bī-hən\\
     when leave.\textsc{prf}.\textsc{3sg.m} from at-\textsc{3pl.m} say.\textsc{prf}.\textsc{3sg.m}-\textsc{dat.3sg.m} \textsc{dm} \textsc{dem}.\textsc{sg}.\textsc{f} \textsc{def}-pomegranate.\textsc{pl} what-make.\textsc{impf}.\textsc{1sg} with-\textsc{3pl.f}\\
\glt ‘When he left them, he said to him, ``Well, what shall I do with these pomegranates?”’
\z

\section{Conclusion}

Because of the dominance of Persian in the Iranian educational system and work environment, the lack of influence from Modern Standard Arabic, and the long period of geographical proximity, the Persian-speaking society of southwest Iran has left many linguistic traces in the language of the Arabic-speaking community of Khuzestan. 

Van Coetsem (\citeyear{VanCoetsem2000}: 59; cf. \citealt{Lucas2015}: 532) suggests that lexical, but not syntactic and phonological transfer is to be expected under RL agentivity. However, Khuzestan Arabic phonology and syntax have been influenced by the SL Persian under RL agentivity, albeit to a much lesser extent than the lexicon.

Khuzestan Arabic does not show transfer of patterns from Persian in either inflectional or derivational morphology. However, we do find an adapted pattern replication of Persian phrasal verbs (with preservation of the Arabic word order). 

As for syntax and contact-induced word order changes, the alternative sentence construction with \textit{čān} in sentence-final position can be explained as a result of Persian influence on Khuzestan Arabic. This change might have been triggered by the similar and very frequent Persian constructions with sentence-final \textit{būdan}. Thus, we do have some syntactic change due to transfer under RL agentivity, which van Coetsem considered to be unexpected (see above).

Persian lexical items have often been borrowed in Khuzestan Arabic for novel concepts (lexical gaps), which is why semantic fields relating to technical or cultural innovations, education, and administration show the greatest amount of Persian borrowing. This also explains why nouns are generally more often transferred than verbs (cf. \citealt{Lucas2015}: 532). Persian words are regularly integrated into Khuzestan Arabic phonology and morphology, for example the Arabic internal plural is formed for Persian nouns. Also, many discourse particles have been transferred from Persian into Khuzestan Arabic. Some of them, e.g. \textit{ham} ‘also’, had been in use generations ago among Arabic speakers in Khuzestan and beyond (Iraq, Gulf).

Of course, contact between Khuzestan Arabic and Persian has always been limited to certain social contexts (outside the family), especially for women, who had and still have much less access to education and employment and thus to the Persian-speaking world. This fact, and some structural differences between the languages, explain the limits of contact-induced language change in Khuzestan Arabic, especially in morphology and syntax.

Hopefully, future research on the dialects of Khuzestan will provide more empirical data on instances of contact-induced change. An enlarged database should especially provide further evidence concerning the development and extent of word order changes.

\section*{Further reading}


\citet{Ingham2011khuz} provides a sketch grammar of Khuzestan Arabic.\\
\citet{Ingham2005} discusses Turkish and Persian borrowings in Khuzestan Arabic and northeastern Arabian dialects.\\
\citet{MatrasShabibi2007} is an article on contact-induced changes in Khuzestan Arabic based on \citet{Shabibi2006}.\\
\citet{Shabibi2006} is an unpublished doctoral dissertation on contact-induced changes in Khuzestan Arabic.



\section*{Acknowledgements}

I would like to thank Stephan Procházka and Dina El-Zarka for their critical remarks and bibliographical suggestions on this chapter. I would additionally like to thank my informant and good friend Majed Naseri for all his help on the transcription and translation of my recordings.

\section*{Abbreviations}

\begin{multicols}{2}

\textsc{acc}  accusative

\textsc{cop}  copula

\textsc{dat}  dative

\textsc{def}  definite

\textsc{dem}  demonstrative

\textsc{dm}  discourse marker

\textsc{ez} Persian \textit{ezāfe}

\textsc{f}  feminine

\textsc{imp}   imperative

\textsc{impf}  imperfect

\textsc{inf}  infinitive

\textsc{indf}  indefinite

KhA  Khuzestan Arabic

\textsc{m}  masculine

\textsc{obl}  oblique

Pers  Persian

\textsc{pl}/pl.  plural

\textsc{ptcp}  participle

\textsc{prf}  perfect

\textsc{prog}  progressive

\textsc{prs}  present

\textsc{pst}  past

\textsc{rel}  relative particle

\textsc{sg}/sg.  singular

\end{multicols}



\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}