\documentclass[output=paper]{langsci/langscibook} 
\author{Ergin Öpengin\affiliation{University of Kurdistan-Hewlêr}}
\title{Kurdish}
% \keywords{} 
\abstract{This chapter provides an overview of the influence of Arabic on Kurdish, especially on its Northern and Central varieties spoken mainly in Turkey–Syria–Iraq and Iraq–Iran, respectively. It summarizes and critically assesses the limited research on the contact-induced changes in the phonology and syntax of Kurdish, and proposes several new dimensions in the morphology and syntax in addition to providing also a first treatment of lexical convergence in Kurdish through borrowings from Arabic.}

\begin{document}
\maketitle

\section{Kurdish and its speech community} 

Kurdish is a Northwestern Iranian language spoken by 25 to 30 million speakers in a contiguous area of western Iran, northern Iraq, eastern Turkey and northeastern Syria. There are also scattered enclaves of Kurdish speakers in central Anatolia, the Caucasus, northeastern Iran (Khorasan province) and Central Asia, with a large European diaspora population. The three major varieties of Kurdish are: (i) Southern Kurdish, spoken under various names near the city of Kermanshah in Iran and across the border in Iraq; (ii) Central Kurdish (also known as Sorani), one of the official languages of the autonomous Kurdish region in Iraq, also spoken by a large population in western Iran along the Iraqi border; (iii) Northern Kurdish (also known as Kurmanji), spoken by the Kurds of Turkey, Syria and the northwestern perimeter of Iraq, in the province of West Azerbaijan in northwestern Iran and in pockets in the west of Armenia (cf. \citealt{HaigÖpengin2014} for a discussion on defining “Kurdish”). Of these three, the largest group in terms of speaker numbers is Northern Kurdish. The Kurdish population in respective states is difficult to reliably determine since none of the sovereign countries make the relevant census information available. \tabref{tab:opengin:1} provides some cautious estimates based on various sources (especially \citealt{Sirkeci2005,ZeyneloğluEtAl2016}; and Ethnologue).\footnote{See \url{https://www.ethnologue.com/language/kur}.}\textsuperscript{,}\footnote{The population figures should not be taken as equivalent to “number of speakers”, since especially in Turkey a significant portion of the Kurdish population grow up with no or very limited knowledge of Kurdish (cf. \citealt{Öpengin2012,ZeyneloğluEtAl2016}).} 

\begin{table}
\begin{tabular}{lr}
\lsptoprule
Country & Population size\\\midrule
{Turkey} & c. 15,000,000\\
{Iraq}   & c. 6,000,000\\
{Iran}   & c. 8,000,000\\
{Syria}  & c. 2,000,000\\
\lspbottomrule
\end{tabular}
\caption{Estimates of Kurdish population numbers}
\label{tab:opengin:1}
\end{table}


\section{The history of Kurdish–Arabic contact} 

Information about the pre-Islamic history of the Kurds and their language is scarce. According to early Islamic sources, at the time of the Islamic conquest of the Near East (Upper Mesopotamia, Iran, and Armenia) in the seventh century (\citealt{BoisEtAl2012}: 451), the communities designated with the term \textit{Kurd} were already living in most of the present-day Kurdish-inhabited areas, namely from Mosul to the north of Lake Van, and from Hamadan to the Jazira region situated around the intersection of present-day Syria, Iraq and Turkey \citep[111]{James2007}. The Kurds have thus been living in contact with various Aramaic-speaking Christian and Jewish communities as well as Arabic-speaking communities since at least the early Islamic period, though the contact of Iranian-speaking populations with Aramaic dates back to the fifth century BCE (cf. \citealt[69]{Utas2005}, citing also \citealt{Folmer1995} and \citealt{Kent1953}). Kurdish differs from other Iranian languages such as Persian in sharing the same or close geographical spaces with Arabic-speaking populations, especially in Upper Mesopotamia. The historical socio-cultural contact between Kurdish and Arabic-speaking communities requires a more refined treatment than is currently possible, but there are a number of medieval Arabic sources which attest to the interaction and mobility of Kurdish and Arabic communities in some regions (e.g. Erbil, Mosul), as well as language shift of some Kurdish communities to Arabic and vice versa (cf. \citealt{BoisEtAl2012}: 449, 452, 456; \citealt{James2007}: 115–120). 

Given the unquestionably prestigious status of Arabic in administration and sciences in the islamicized Near East, consolidated especially under Abbasid rule (which included most of the Kurdish-inhabited areas), Kurdish was heavily dominated by Arabic. Even in several of the important medieval Kurdish dynasties such as that of the Marwānids (10th–11th centuries), Arabic enjoyed the high status of being the administrative and literary language (cf. \citealt{James2007}: 112), since the coins bore Arabic script, while \textit{qaṣīda} reading ceremonies or contests would feature primarily Arabic, but to a limited extent also Persian pieces (\citealt{Ripper2012}: 507–528). With the conquest of the Kurdish-inhabited regions by Turkic peoples and Mongols from tenth century onwards, which led also to the final overthrowing of the Abbasid state in 1248 by the Mongols, the Arabic-speaking populations may have started to diminish and retreat. Although at this stage Persian attained a firm status as the literary language in the Islamic East (\citealt{Perry2012}: 73), Arabic preserved its higher status in administration and, later on, especially in education, well into the end of the nineteenth century. Thus, Kurdish developed a literary tradition only starting from the sixteenth century, but its limited usage was largely restricted to writing verse throughout the following several centuries. The literature in this period is heavily dominated by the vocabulary and literary formulas and metaphors of the two dominant languages, Arabic and Persian (cf. \citealt{Öpenginforthcoming}). 

In the early twentieth century, with the dissolution of the Ottoman Empire, Kurdish in Iraq and Syria again came into primary direct contact with Arabic. In Iraq, up until 1991, with the establishment of a Kurdish autonomous region, the language configuration was one in which Arabic was the prestigious language of higher domains. Not being in possession of any official status, the Kurds in Syria have been in a highly asymmetric language-contact situation with Arabic. In Turkey, especially in Mardin and Siirt provinces, Kurds have been in contact with Arabic-speaking communities, but as the lingua franca of the communities of cultural–historical Kurdistan (cf. \citealt{Edwards1851}: 121), Kurdish must have been the dominant language of interaction between these communities, and it is indeed possible to observe important influences from Kurdish on the local Arabic dialects (cf. \citealt{Jastrow2011Kurdish} and §\ref{bkm:Ref14688184} below.). 

As a result of these differing degrees and modalities of contact with Arabic, the influence of Arabic should be viewed as consisting of at least two layers, and viewed separately for different country contexts where Kurdish is spoken. Of the two layers, there should be assumed a deeper contact influence, shared in larger portions of Kurdish-speaking areas, dating to before the twentieth century; and a more shallow layer that is the result of the more recent societal bilingualism in Iraq and Syria. Likewise, while in Syria and Iraq the Arabic influence on Kurdish continues, this influence is largely replaced by influence from the dominant state languages in Turkey and Iran.~Naturally, the intensity of Arabic influence on Kurdish shows a great deal of variation across Kurdish varieties and dialects within varieties. Accordingly, the historically deeper-layer Arabic influence on Kurdish is characterized by its being restricted mostly to lexicon and being shared in the majority of Kurdish dialects. This has been the result of borrowing under RL-agentivity in the sense of Van Coetsem (\citeyear{VanCoetsem1988,VanCoetsem2000}). On the other hand, the relatively advanced Arabic influence on the Kurdish spoken in the historical Jazira region (including Mosul, northeast Syria, and Mardin province in southeast Turkey), as well as the more recent Arabic influence on the Kurdish spoken in Syria, but also – albeit more restrictedly – in Iraq, concerns also grammatical constructions and at least some of that contact influence could be due to imposition under SL-agentivity.    

\section{Contact-induced changes in Kurdish} 

\subsection{\label{bkm:Ref14688184}Phonology}

The consonant inventory of Kurmanji is given in \tabref{tab:opengin:2}.\footnote{Kurdish data are transcribed in the standard Kurdish Latin alphabet with some additions for emphatics and pharyngeals, mostly consonant with the Library of Congress approach for the romanization of Kurdish: \url{https://www.loc.gov/catdir/cpso/romanization/kurdish.pdf}.} 

\begin{table}
\begin{tabularx}{\textwidth}{lXXXXXXXX} 
\lsptoprule
& \rotatebox{66}{bilabial} & \rotatebox{66}{labio-dental} & \rotatebox{66}{alveolar} & \rotatebox{66}{palatal} & \rotatebox{66}{velar} & \rotatebox{66}{uvular} & \rotatebox{66}{pharyngeal} & \rotatebox{66}{glotttal}\\\midrule
{Plosive} & {p’ p b} & & {t’ t d} & & {k’ k  g} & {q} &  & {’}\\
{Fricative} &  & {f v} & {s z} & {ş j} & {x ẍ} &  & {ḥ  ʿ} & {h}\\
{Affricate} &  & & & {ç’  ç  c} &  &  &  & \\
{Nasal} & {m} & & {n} &  &  &  &  & \\
{Trill} &  & & {\={r}} &  &  &  &  & \\
{Flap} &  & & {r} &  &  &  &  & \\
{Approximant} & {w} &  &  & {y} &  &  &  \\
{Lateral} &  & & {l} &  &  &  &  & \\
\lspbottomrule
\end{tabularx}
\caption{\label{bkm:Ref14688948}\label{tab:opengin:2}Consonant phonemes in Kurmanji}
\end{table}

In cells of doublets/triplets, the voiceless phonemes come first. The apostrophe on plosive and fricative phonemes indicates aspiration, which marks a phonemic distinction in Kurmanji. In addition to these consonants with indisputable phonemic status, there are the so-called emphatic or pharyngealized variants of the obstruents /p, b, t, d, s, z/. These variants are transcribed in the text with a dot beneath the characters. 

The consonant inventory of Sorani is basically identical with \tabref{tab:opengin:2}, except: (i) it does not have unaspirated stop phonemes; and (ii) it has velar nasal and velarized lateral phonemes \citep[27]{Öpengin2016}. 

Arabic (or more generally Semitic) influence on the phonology of Kurdish is most clearly observed in the presence of the two pharyngeal phonemes \textit{ḥ} [ħ] and \textit{ʿ} [ʕ] (cf. \citealt{Kahn1976,Haig2007,Anonbyforthcoming,Barryforthcoming}), as well as the series of emphatic obstruants \textit{ṭ,} \textit{ḍ,} \textit{ṣ,} and \textit{ẓ} (Haig \& Öpengin 2018), respectively. The precise Semitic source language for these sounds cannot be determined, since Kurdish (or rather its ancestor languages) must have been in close contact with various Semitic languages for more than two millennia \citep[69]{Utas2005}. However, these phonemes set the consonant inventory of Kurdish clearly apart from other West Iranian languages such as Persian, with the only other West Iranian languages possessing both pharyngeals and emphatic consonants being Zazaki, and the Kumzari language spoken mainly in Oman \citep{Anonbyforthcoming}. In what follows, I illustrate the presence and interactions of the pharyngeal and emphatic consonants in Kurdish, and provide a brief discussion of their paths of development.\footnote{The Kurmanji lexical items presented in this section are based on my native-speaker knowledge of the Şemdînan (Şemdinli) dialect, and my knowledge of Kurmanji-internal dialectal variation, drawing also on \citep{Chyet2003}, \citep{ÖpenginHaig2014}, and the  Manchester Database of Kurdish Dialects presented in \citep{MatrasKoontz-Garboden2016}. The Sorani lexical items are from \citet{Öpengin2016} and the popular press.}  

The pharyngeal phonemes are found in varying degrees in both Central Kurdish and Northern Kurdish. They are retained in most of the Arabic loanwords originally bearing them, a list of which is given in \tabref{tab:opengin:3}.\footnote{Note that all through the article, unless stated otherwise, the Arabic data represents Classical Arabic, giving an approximation of the ultimate Arabic etyma of the items without necessarily implying that these are the immediate source of the Kurdish items (as they may have been borrowed from local Arabic dialects as well as through the intermediary languages such as Persian or Ottoman). Furthermore, the glosses in tables are for Kurdish items, as sometimes the meanings of the Arabic etyma are not completely identical.} 

\begin{table}
\begin{tabular}{llll}
\lsptoprule
{Arabic} & {Northern Kurdish} & {Central Kurdish} & {Gloss}\\\midrule
\textit{ʕarab}     & \textit{ʿereb}        & \textit{ʿereb}                    & ‘Arab’\\
\textit{maʕlūm}    & \textit{meʿlûm}       & \textit{meʿlûm}    & ‘evident’\\
\textit{ʕadāla(t)} & \textit{ʿedalet}      & \textit{ʿedaḷet}                  & ‘justice’\\
\textit{ṭābiʕ}     & \textit{tabiʿ/ṭabiʿ}  & \textit{ṭabiʿ}                    & ‘dependent’\\
\textit{maḥall}    & \textit{miḥele}       & \textit{meḥel}                    & ‘neighborhood’\\
\textit{maḥšar}    & \textit{meḥşer}       & \textit{meḥşer}                   & ‘resurrection (day)’\\
\textit{ḥākim}     & \textit{ḥakim}        & \textit{ḥakim}                    & ‘judge, governor’\\
\textit{ḥammām}    & \textit{ḥemam}        & \textit{ḥemam}                    & ‘bath’\\
\textit{baḥr}      & \textit{beḥr}         & \textit{beḥr}                     & ‘sea’\\
\lspbottomrule
\end{tabular}
\caption{\label{bkm:Ref14691447}Loanwords with retained pharyngeals in Kurdish}
\label{tab:opengin:3}
\end{table}

Some loanwords with original pharyngeals are reanalysed as containing their non-pharyngeal counterparts. Such is the word \textit{haq} from Arabic \textit{ḥaqq} ‘right’, or the Arabic word \textit{ṭaʕm} ‘taste’ that is seen in eastern dialects of Northern Kurdish and in Central Kurdish without the voiced pharyngeal as \textit{ṭam} and \textit{tam}, respectively. 

Furthermore, an original pharyngeal in a loanword may be substituted with the alternative pharyngeal sound, so, for example, the voiced pharyngeal of the Arabic \textit{ṭamaʕ} ‘greed’ may be realized as either of the pharyngeals in different Kurdish dialects. Such indeterminate or alternative use of pharyngeals may exist within a single dialect (cf. \citealt{Kahn1976}: 25). For instance, in the Mukri dialect of Central Kurdish, (\citealt{Öpengin2016}: 41–42) the following Arabic-origin words can be found in both of the form pairs: \textit{saʿib} {\textasciitilde} \textit{saḥib} ‘owner’, \textit{ʿerz} {\textasciitilde} \textit{ḥerz} ‘honour’, \textit{cemaʿet} {\textasciitilde} \textit{cemaḥet} ‘community’. 

Finally, a pharyngeal may develop in loanwords that have no pharyngeal in the source language. Thus, in most of Northern Kurdish the Arabic word \textit{ʔarḍ} ‘earth’ appears with a non-etymological pharyngeal as \textit{ʿerd}, while the Arabic word \textit{ǧāhil} ‘naïve, young’ is seen with a pharyngeal as \textit{caḥêl} (but also \textit{cahil}).    

Although the pharyngeals in Kurdish occur mostly in Arabic loanwords, they have expanded also into inherited native Iranian lexicon, especially in Northern Kurdish. However, unlike in Arabic loanwords, fluctuation between pharyngeal and non-pharyngeal uses of such words among the dialects (sometimes in immediate geographic proximity) is readily apparent. \tabref{tab:opengin:4} presents some native Iranian words of this kind. Where relevant, the non-pharyngeal forms are also noted, while Persian cognates are included for comparison.  

\begin{table}
\begin{tabular}{llll}
\lsptoprule
{Persian} & {Northern Kurdish} & {Central Kurdish} & {Gloss}\\\midrule
\textit{abr} & \textit{ʿewr} & \textit{hewr} & ‘cloud’\\
\textit{zabān} & \textit{ʿezman {\textasciitilde} ziman} & \textit{ziman} & ‘language’\\
\textit{āsemān} & \textit{ʿesman} & \textit{asman {\textasciitilde} hasman} & ‘sky’\\
\textit{ḫošk} & \textit{ḥişk {\textasciitilde} hişk}  & \textit{wişk} & ‘dry, hard’\\
\textit{haft} & \textit{ḥeft {\textasciitilde} heft} & \textit{ḥewt}  & ‘seven’\\
\textit{hašt} & \textit{ḥeşt {\textasciitilde} heşt} & \textit{heşt} & ‘eight’\\
\textit{bahašt} & \textit{biḥeşt {\textasciitilde} bihişt} & \textit{beheşt} & ‘paradise’\\
\lspbottomrule
\end{tabular}
\caption{\label{bkm:Ref14705119}Pharyngeal sounds in native Iranian lexical items}
\label{tab:opengin:4}
\end{table}

More striking, however, is the emergence of a voiced pharyngeal in a subset of words with similar structure in the northern dialects of Northern Kurdish that are geographically farthest from direct Arabic/Semitic contact but close to Caucasian languages which also possess pharyngeals. Thus, the native words such as \textit{masî} ‘fish’, \textit{çav} ‘eye’, \textit{mar} ‘snake’ (in Central Kurdish and in central and southern dialects of Northern Kurdish) appear in the northern dialects of Northern Kurdish with a pharyngeal, as \textit{meʿsî,} \textit{çeʿv,} \textit{meʿr}. These are obviously the result of language-internal processes, though nested in an initial introduction of the phonemes into the language via contact with either Arabic or Caucasian languages, or both.

As for their distribution, the pharyngeal phonemes are most robustly present in the central areas of the Northern and Central Kurdish speech zones. Their presence in Arabic loanwords is weakened towards the extreme northern and southern peripheries in heavy contact with Turkish and Persian (cf. Map 1.27 in the Manchester Kurdish Database, which illustrates such weakening of pharyngeals at the peripheries through the distribution of the Arabic loanword \textit{ḥeywan} ‘animal’).\footnote{See \url{http://kurdish.humanities.manchester.ac.uk/pharyngeal-retentionloss-animal/}.}

We turn now to the series of emphatic (pharyngealized) obstruents \textit{ṭ, ḍ} and \textit{ṣ, ẓ}. \tabref{tab:opengin:5} gives a list of Arabic loanwords in which the original emphatic consonant is retained in Kurdish. 

\begin{table}
\begin{tabular}{lll}
\lsptoprule
Arabic & Northern/Central Kurdish & Gloss\\\midrule
\textit{ṭaʕm}   & \textit{ṭam \textasciitilde ṭeʿm\slash tam} & ‘taste’\\ 
\textit{ṭāʔir}  & \textit{ṭeyr}                               & ‘bird’\\ 
\textit{baṭṭāl} & \textit{beṭal}                              & ‘empty, cancel’\\ 
\textit{ð̣ulm}  & \textit{ẓulm}                               & ‘oppression’\\ 
\textit{ḍābit}  & \textit{ẓabit}                              & ‘clerk’\\ 
\textit{ṣūfī}   & \textit{ṣofî}                               & ‘devotee, Sufi’\\ 
\textit{ṣāfī}   & \textit{ṣaf/ṣafî}                           & ‘clear’\\
\lspbottomrule
\end{tabular}
\caption{Arabic loanwords with emphatic consonants in Kurdish\label{tab:opengin:5}}
\end{table}

In the deeper-layer loanwords, the Arabic interdental and voiced alveolar emphatics are merged into the voiced emphatic alveolar phoneme \textit{ẓ} in Kurdish. But in present-day Iraqi and Syrian Kurdish speech, especially those speakers with formal education may also pronounce the interdental phoneme, especially in the case of nonce borrowings and code-mixing.  

On the other hand, quite a number of Arabic loanwords are pronounced without their original emphatic consonants, reanalysed thus as the corresponding plain phonemes (similarly to Persian), as in the items in \tabref{tab:opengin:6}. 

\begin{table}
\begin{tabular}{llll}
\lsptoprule
{Arabic} & {Northern Kurdish} & {Central Kurdish} & {Gloss}\\\midrule
\textit{ḫāṭir}  & \textit{xatir}                           & \textit{xatir}  & ‘mind’\\
\textit{ṭaraf}  & \textit{teref}                           & \textit{teref}  & ‘side, direction’\\
\textit{šayṭān} & \textit{şeytan}                          & \textit{şeytan} & ‘devil’\\
\textit{ḍaʕīf}  & \textit{zeʿîf}                           & \textit{zeʿîf}  & ‘weak’\\
\textit{ḥāḍir}  & \textit{ḥazir}                           & \textit{ḥazir}  & ‘ready’  \\
\textit{qaṣṣāb} & \textit{qesab}                           & \textit{qesab}  & ‘butcher’\\
\textit{fasīḥ}  & \textit{fesîḥ}                           & \textit{fesîḥ}  & ‘clear’\\
\textit{ṣabr}   & \textit{sebr{\textasciitilde}ṣebr}       & \textit{sebr}   & ‘patience’\\
\lspbottomrule
\end{tabular}
\caption{\label{bkm:Ref14707500}Arabic loanwords with lost emphatics in Kurdish\label{tab:opengin:6}}
\end{table}

On the reverse side, some Arabic loanwords with no original emphatic consonants are pronounced with emphatic consonants in Kurdish, such as \textit{ẓełał} ({\textasciitilde} \textit{ẓelal} and \textit{zelal}) from Arabic \textit{zulāl} ‘clear’ (dialectal \textit{zalāl}), or \textit{ẓelam} ‘man’ from Syrian Arabic \textit{zalame}.    

Finally, just as with the pharyngeal consonants, emphatic sounds also appear in inherited native Iranian words, as illustrated in Table \ref{tab:opengin:7}. 

\begin{table}
\begin{tabular}{lll}
\lsptoprule
Northern Kurdish & Central Kurdish & Gloss\\\midrule
\textit{meẓin} & - & ‘big’\\
\textit{ẓiman} & \textit{ẓiman} & ‘language’\\
\textit{ẓik} \textit{{\textasciitilde} zik} & \textit{zik} & ‘stomach’\\
\textit{aẓad}  & \textit{aẓad} & ‘free’\\
\textit{ẓava} & \textit{ẓawa} & ‘groom’\\
\textit{beẓîn} & \textit{beẓîn}  & ‘to run’\\
\textit{peẓ} & \textit{pez} & ‘sheep’\\
\textit{ṣal} & \textit{ṣał} & ‘year’\\
\textit{ṣed} \textit{{\textasciitilde} sed} & \textit{ṣed} & ‘hundred’\\
\textit{ṣe} & \textit{ṣeg} & ‘dog’\\
\textit{beṣ} \textit{{\textasciitilde} bes} & \textit{beṣ} \textit{{\textasciitilde} bes} & ‘enough’\\
\textit{ṣawa} & \textit{sawa} & ‘very young, newborn’\\
\textit{ṣotin} & \textit{sûtan} & ‘to burn’\\
\textit{ṣiṣt} & \textit{ṣiṣt} & ‘loose’\\
\textit{ṭarî} & \textit{tarîk} & ‘dark’\\
\textit{ṭezî} & \textit{tezî} & ‘cold’\\
\textit{ṭeng} & \textit{teng} & ‘narrow’\\
\textit{ṭerm} & \textit{term} & ‘dead body’\\
\textit{ṭirş} & \textit{tirş} & ‘sour’\\
\textit{ḍaṣ(ik)} & \textit{das} & ‘sickle’\\
\textit{ḍiṙî} & - & ‘blackberry bush’\\
\lspbottomrule
\end{tabular}
\caption{Emphatic consonants in native Iranian lexical items}
\label{tab:opengin:7}
\end{table}

Of the emphatic obstruents, the fricative pair (\textit{ṣ,} \textit{ẓ}) are found both in Northern and Central Kurdish, while the stops (\textit{ṭ,} \textit{ḍ}) are found only in Northern Kurdish, with the voiced counterpart being extremely rare. The fact that the voiceless emphatic stop is widespread only in Northern Kurdish most probably has to do with the presence of two series of aspirated and unaspirated voiceless stops in the language (cf. \tabref{tab:opengin:2}). The unaspirated stops are probably intermediary in the development of emphatics. This is further reinforced by the fact that in Northern Kurdish the bilabial voiceless stop \textit{p} also has an emphatic version, as in the native words \textit{ṗeẓ} ‘sheep’ and \textit{ṗenîr} ‘cheese’ (in some dialects; cf. \citealt{Kahn1976}: 27). Within Northern Kurdish, they are found in more southerly dialects, and are noted to be particularly frequent in both the Kurdish and Neo-Aramaic of Duhok and Hakkari provinces  \citep[329]{Blau1989}. They tend to be less present moving northwards (Erzurum–Kars) while MacKenzie (\citeyear[43]{MacKenzie1961}) notes that they are altogether absent in the Yerevan dialect. This distribution is of course consistent with a language-contact scenario, in the sense that in the northern dialects away from Semitic influence the language either did not develop emphatics or lost them as a result of contact with and bilingualism in Armenian, Turkic and Caucasian languages that do not possess such emphatics.   

Given the shallow history of written Kurdish, it is not possible to determine the historical period of the introduction of the emphatics and pharyngeals into the language. However, they are found abundantly even in the earliest Kurdish texts, especially in the Arabic component, but also in inherited lexical items, such as \textit{ṣal} ‘year’, \textit{ṣar} ‘cold’, \textit{ṣed} ‘hundred’, \textit{meẓin} ‘big’, \textit{ḥemyan} ‘all of them’ (items taken from \textit{Şêxê} \textit{Senʿaniyan} by the early seventeenth-century poet Feqiyê Teyran, cf. \citealt{Teyran2011}). 

Three studies have treated the pharyngeals and emphatics in Kurdish, namely \citet{Kahn1976}, \citet{Anonbyforthcoming} and \citet{Barryforthcoming}. \citet{Barryforthcoming} suggests that the pharyngeal sounds (including emphatics) in Kurdish are the result of contact influence from Arabic with a phonetic basis. The phonetic basis consists in the recategorization of vowels and the \textit{h} sound within syllables with “flat” consonants (including pharyngeals, rhotics, grooved fricatives, and labials). Thus, initially, through extensive language contact with and bilingualism in Arabic, the speakers attained an active category of pharyngeals. Then the (inherited or loan) vocabulary with sounds that have pharyngeal-like effects on neighbouring vowels led to the reanalysis of the given vocabulary items as pharyngeal. In this account, the whole syllable is pharyngeal rather than individual sound segments. This account is particularly appropriate since, while it acknowledges the role of language contact with Arabic in the initial stage, it posits a phonetic mechanism of language-internal development of pharyngealization that captures an expansion of pharyngeals into historically non-pharyngeal lexical items that would be impossible to explain on purely language-contact grounds. It is, for instance, consistent with the fact that, in the above-presented data, the emphatics, but not pharyngeals in loan words, are restricted to the environment of more open vowels: \textit{e}, \textit{a}, \textit{o}, and \textit{i} [ɪ]. Furthermore, although not stated in the source study, the assumed subsequent development of a phonetic basis for the propagation of the pharyngeals into items originally without pharyngeal sounds is consonant with the facts of different stages or layers of borrowing. For instance, from the Arabic root \textit{√ǧmʕ} we have three forms in Kurmanji: \textit{civat} ‘community, company’, \textit{cimat} ‘the assembly of prayers in a funeral’, and \textit{cemaʕet} ‘community’. The first form is probably the result of an early borrowing right after the  initial islamicization of the Kurds, as the fricativization of the bilabial nasal was active then (as seen also in \textit{silav} ‘greeting’ from Arabic \textit{salām}; \citealt{Paul2008}). The second form with a slightly specialized semantic difference may have originated in a dialect where the mentioned fricativization did not occur. In any case, the first two forms, which are clearly early borrowings, did not retain the original pharyngeal, whereas in a later borrowing from the same root, when one can assume that the pharyngeals were better tolerated in the language (and that the fricativization of bilabial nasal was not active), the pharyngeal sound did survive.   

However, this account fails to explain why, in the great majority of the vocabulary with the relevant phonetic environment (syllables with “flat” consonants and low and back vowels), pharyngealization has not occurred. If the phonetic mechanism is integrated into the phonological system of the language, then pharyngealization would be expected in all relevant contexts. In this sense, although there is a phonetic ground to the propagation of the pharyngeals and emphatics in Kurdish, it may be safer not to postulate it as integrated into the phonological system of the language. Rather, the pharyngeals and emphatics should still be considered as peripheral to the phonological system (cf. \citealt{Haig2007,Anonbyforthcoming}), since, as noted by Haig (\citeyear[167]{Haig2007}), they are restricted to individual lexical items, their functional load is very limited, and there is considerable cross-speaker and cross-dialectal variability in the extent of their presence. 

Although it is not the main focus of this chapter, a note on the reverse direction of contact influence is in order at this point. The Arabic dialects of Anatolia or Upper Mesopotamia (Mardin, Siirt, Kozluk, Sason, and the plain of Muş) have adopted some consonant and vowel phonemes via loanwords from Kurdish and Turkish, which do not exist in mainstream Arabic dialects (\citealt[84]{Jastrow2011Kurdish}; Akkuş, this volume: §3.1.1).\ia{Akkuş, Faruk@Akkuş, Faruk} The phonemes and example words with their sources are given in Table \ref{tab:opengin:8}.

\begin{table}
\begin{tabular}{ll}
\lsptoprule 
Phoneme & Example\\\midrule
bilabial stop /p/   &   \textit{parčāye} ‘piece’ < Tr. \textit{parça}\\
voiced labio-dental fricative /v/  & \textit{davare} ‘ramp’  < Kr. \textit{dever} (\textsc{f)} ‘place’\\
voiceless affricate /č/     &  \textit{č\kern 0.75ptǝ\kern -1ptqmāq} ‘lighter’ < Tr. \textit{çakmak}\footnote{It is more probable that this word (and others attributed to Turkish) is borrowed via Kurdish, since the uvularization (/k/ > /q/) in loanwords and the change in the vowel of the first syllable (cf. also \textit{qeymaẍ} ‘cream’, from Turkish \textit{kaymak}) are typical of Kurmanji spoken in the region.}\\
voiced palatal fricative /ž/\footnote{Note that the reflex of Arabic 〈\kern .75pt{\arabscript{ج}}〉 in this variety is /ǧ/, not /ž/.}    & \textit{ṭāžī} ‘greyhound’ < Kr. \textit{ṭajî}\\
voiced velar stop /g/      & \textit{gōmlak} ‘shirt (modern)’ < Tr. \textit{gömlek}\\
mid long front vowel /ē/    & \textit{tēl} ‘wire’ < Tr. \textit{tel} (via Kr. \textit{têl})\\
mid long back vowel /ō/\footnote{Note also that the original Arabic diphthongs *ay and *aw are preserved in this variety, not monophthongized to /ē/ and /ō/.}  &  \textit{ḫōrt} ‘young man’ < Kr. \textit{xort}\\
\lspbottomrule
\end{tabular}
\caption{Borrowed phonemes in Arabic dialects of Anatolia}
\label{tab:opengin:8}
\end{table}

These additions into the phoneme inventory of the Anatolian Arabic are evidently the result of contact with Kurdish and Turkish. The introduction of these new phonemes has, as noted by Jastrow (\citeyear[84]{Jastrow2011Kurdish}), on the one hand re-established the lacking symmetry caused by historical sound changes in Old Arabic, while on the other hand causing further sound shifts in the inherited Arabic vocabulary. 

\subsection{Morphology}

It is usually assumed that Arabic influence on Kurdish is absent in the grammar (e.g. \citealt{Edwards1851}), being largely restricted to phonology and lexicon. This is indeed to a large extent true. There are, however, several potential grammatical features that may be related to such contact influence. 

Matras (\citeyear[75]{Matras2010}) suggests that the presence of aspect--mood prefixes in the languages of the Eastern Anatolian linguistic zone, namely Persian, Kurdish, Neo-Aramaic, Arabic and Western Armenian, is an outcome of language contact. Accordingly, all of these languages have a progressive–indicative aspectual prefix (in turn: \textit{mī-,} \textit{di-,} \textit{gǝ-}, \textit{ko-,} \textit{ba-/-a-}), while subjunctive is marked either by the absence of the indicative prefix (Armenian and Neo-Aramaic) or by a specialized subjunctive prefix (Persian, Kurdish, Arabic). Since such aspect--mood prefixes are considered typical of Iranian languages of the region, they would have diffused from Kurdish and Persian into the other languages of the zone, including Arabic (which in its standard grammar does not have such forms; \citealt{Ryding2014}: 46–47). However, assessing the validity of Eastern Anatolia’s being a linguistic area, Haig (\citeyear{HaigÖpengin2014}: 20–25) casts doubt on this claimed contact scenario, primarily since: (i) the feature exists in Arabic dialects outside the region; and (ii) it is absent in the two major languages of Anatolia, namely Turkish and Zazaki. Jastrow (\citeyear[92]{Jastrow2011Kurdish}), on the other hand, although acknowledging the source of such verbal prefixes grammaticalizing from Old Arabic verb forms, hypothesizes – though without providing supporting arguments – that they may have developed under Turkish and Kurdish influence. Assessing also the grammaticalization of such formatives in various languages and rejecting a contact scenario behind their frequent occurrence in the languages of Anatolia, Haig (\citeyear[26]{Haig2014}) concludes that the present indicative prefixes found in Kurmanji, and in certain varieties of Aramaic and Arabic in Anatolia, could be interpreted as reflexes of an inherited morphological template, which is well-attested in the related Northwest Iranian and Semitic languages outside Anatolia.

Another (not previously discussed) candidate for Arabic influence on Kurmanji Kurdish relates to gender assignment in more recent loanwords from European languages. In Kurmanji, like Arabic, nouns are assigned to feminine and masculine genders. The gender of inanimate nouns is largely arbitrary, with limited morpho-phonological basis in both languages. In Arabic, words carrying the \textit{{}-a} ending are feminine, while in Kurmanji abstract nouns ending in \textit{-î} are feminine, while the rest may be of either gender. Now, when Arabic borrows modern vocabulary items from European languages, items ending in \textit{{}-a} are assigned to feminine gender, while the rest are assigned to masculine gender \citep[5]{Ibrahim2015}. The default gender assigned to new lexical borrowings is masculine in Arabic. There is as yet no research on the gender assignment of borrowings in Kurmanji. However, it is easily observed that Kurmanji spoken in Turkey mostly favors feminine, while the Kurmanji of Iraq uses masculine gender for integrating modern vocabulary items into the language. The modern lexical borrowings (boldface) in \REF{bkm:Ref14712415} are all assigned to masculine gender in Badini Kurmanji of Iraq. Note that the gender of the nouns is visible in the \textit{ezāfe} (see §\ref{bkm:Ref520275931}) and oblique case suffixes. 

%%1st subexample: change \ea\label{...} to \ea\label{...}\ea; remove \z  
%%further subexamples: change \ea to \ex; remove \z  
%%last subexample: change \z to \z\z 
\ea{Badini dialect of Kurmanji in Iraq (from media outlets)}\label{bkm:Ref14712415}\\
\ea 
\gll \textbf{sîstem}-ê endroyd-ê\\
     system-\textsc{ez.m} android-\textsc{obl.f}\\
\glt ‘Android system’
\ex\label{parl}
\gll serok-ê \textbf{parleman}î\\
     president-\textsc{ez.m} parliament-\textsc{obl.m}\\
\glt ‘the president of the parliament’
\ex \gll \textbf{form}ê têgehiştin-ê\\
     form-\textsc{ez.m} understanding\textsc{{}-obl.f}\\
\glt ‘the form of understanding’
\ex \gll \textbf{moral}-ê diyalog\\
     moral-\textsc{ez.m} dialogue\\
\glt ‘the moral of dialogue’
\ex \gll \textbf{proj(e)}-ê av-ê\\
     project-\textsc{ez.m} water-\textsc{obl.f}\\
\glt ‘the water project’
\ex
\gll \textbf{prensîp}-ê hevwelatîbûn-ê\\
     principle-\textsc{ez.m} citizenship-\textsc{obl.m}\\
\glt ‘the principle of citizenship’
\z
\z

Now, all of these lexical borrowings exist also in Kurmanji as spoken in Turkey, but they are systematically used with feminine gender. For instance the phrase in (\ref{parl}) would be realized as \textit{serok-ê} \textit{parleman-ê} (president-\textsc{ez.m} parliament-\textsc{obl.f}), with the feminine form of the oblique case suffix. 

As was stated above, the majority of such modern lexical borrowings in Arabic are assigned to masculine gender. The masculine gender assignment in Kurmanji in Iraq is thus most probably motivated by the Arabic gender assignment pattern. This is all the more plausible when we consider that Arabic, as the dominant state language for the Iraqi Kurds for almost a century, serves also as the intermediary language via which such lexical items are normally borrowed into Kurmanji in Iraq. However, this contact influence must have been established relatively recently, since earlier technical borrowings in Kurmanji in Iraq such as \textit{têlevizyon} and \textit{radyo} are treated as feminine nouns, despite being masculine in Arabic.  

\subsection{\label{bkm:Ref520275931} Syntax}

Although several studies have dealt with the outcomes of language contact between Kurdish and (Neo-)Aramaic in the grammar of these languages – especially on such topics as alignment \citep{Coghill2016}, word order \citep{Haig2014}, and noun phrase morphology \citep{Noorlander2014} – as far as I am aware, the only study on Arabic–Kurdish contact in grammar is the short note of \citet{Tsabolov1994} about the distinctive position of the possessor in a multiple-modifier noun phrase in Northern Kurdish. 

As is well known, a number of West Iranian languages (Middle and contemporary Persian, Kurdish, Zazaki, etc.) employ a vocalic bound morpheme for linking post-head modifiers in a noun phrase, called \textit{ezāfe} or \textit{izāfe} (from Arabic \textit{ʔiḍāfa} ‘joining, addition’), as in (\ref{bkm:Ref14712641}) and (\ref{bkm:Ref14712647}). 

\ea\label{bkm:Ref14712641}Persian (personal knowledge)\\
\gll ḫāna-e bozorg\\
     house-\textsc{ez} big\\
\glt ‘(the) big house’
\z

\protectedex{
\ea\label{bkm:Ref14712647}Northern Kurdish (personal knowledge)\\
\gll xanî-yê mezin\\
     house-\textsc{ez.m} big\\
\glt ‘the big house’
\z
}

The \textit{ezāfe} in Northern Kurdish differs from its cognates in, for instance, Central Kurdish and Persian, as it inflects for gender (masculine \textit{{}-ê} vs. feminine \textit{{}-a}) and number (singular \textit{{}-ê/-a} and plural \textit{{}-ên/-êd}), in addition to having secondary or pronominal forms used in chain \textit{ezāfe} constructions with multiple modifiers (and some other predicative functions; cf. \citealt{Haig2011,HaigÖpengin2018}). In most West Iranian languages, noun phrases with multiple modifiers have their head noun first, followed by qualitative then possessive modifiers, as in (\ref{bkm:Ref14712778}) and (\ref{bkm:Ref14712785}). This is also the order in Middle Persian, as in (\ref{bkm:Ref14712798}), where Tsabolov (\citeyear[122]{Tsabolov1994}) considers such constructions may be regarded as prototypes of the \textit{ezāfe} constructions of modern West Iranian languages. 

\ea\label{bkm:Ref14712778}Persian (personal knowledge)\\
\gll ḫāna-e bozorg-e Malek\\
     house-\textsc{ez} big-\textsc{ez} \textsc{pn}\\
\glt ‘Malek’s big house’
\z

\ea\label{bkm:Ref14712785}\label{bkm:Ref14771780}Central Kurdish  (personal knowledge)\\
\gll kurr-î gewre-y Karwan\\
     son-\textsc{ez} big-\textsc{ez} \textsc{pn}\\
\glt ‘my friend’s beautiful daughter’
\z

\ea\label{bkm:Ref14712798}Middle Persian \citep[122]{Tsabolov1994}\\
\gll pus ī mas ī Artavān\\
     son \textsc{ez} big \textsc{ez} \textsc{pn}\\
\glt ‘Artavan’s elder son’
\z

However, in Northern Kurdish the order of modifiers is reversed, such that a possessor of the head noun in the noun phrase comes before attributive modifiers, as in (\ref{bkm:Ref14712924}), where the secondary linking element is glossed as \textsc{sec}.  

\protectedex{
\ea\label{bkm:Ref14712924}Northern Kurdish (personal knowledge)\\
\gll xanî-yê Malik-î (y)ê mezin\\
     house-\textsc{ez.m} \textsc{pn-obl.m} \textsc{ez.m.sec} big\\
\glt ‘Malik’s big house’
\z
}

Tsabolov observes that these syntactic particularities of Northern Kurdish have no parallels in other Kurdish varieties and Iranian languages as a whole, but that they correspond to the word order in noun phrases in Arabic, as can be seen in the comparison of (\ref{bkm:Ref14712996}) and (\ref{bkm:Ref14713006}). 

\ea\label{bkm:Ref14712996}\label{bkm:Ref14771925}Arabic \citep[123]{Tsabolov1994}\\
\gll miḥfað̣atu ṭ-ṭālibi l-ǧadīdatu\\
     bag\textsc{.nom} \textsc{def}{}-student.\textsc{gen} \textsc{def}{}-new.\textsc{f.nom} \\
\glt ‘the student’s new bag’
\z

\ea\label{bkm:Ref14713006}Northern Kurdish \citep[123]{Tsabolov1994}\\
\gll çent-ê şagirt-î taze\\
     bag-\textsc{ez.m} student-\textsc{ez.m.sec} new\\
\glt ‘the student’s new bag’
\z

Note that although in standard Kurmanji (Northern Kurdish) the forms of the primary and secondary \textit{ezāfes} are identical, with the difference being in the latters’ status either as enclitics or independent particles, in the northern dialect of Northern Kurdish considered by Tsabolov, the singular forms of the secondary \textit{ezāfe} are different (with masculine \textit{{}-î} and feminine \textit{{}-e}). In Tsabolov’s view, the centuries-old close contacts between Kurdish and Semitic dialects, especially Arabic, have not only resulted in the above-described change of noun-phrase-internal word order (syntactic) but also in the development of secondary forms of \textit{ezāfe} through the “weakening” of the primary ones (morphological), because, he argues, such distinct forms “were necessary for correlating each attribute in an [\textit{ezāfe}] chain with the ruling noun they refer to” (\citeyear{Tsabolov1994}: 123). 

On closer scrutiny, however, the motivation Tsabolov puts forward for the morphological change may not be entirely correct, since, on the one hand, \textit{ezāfe} forms in Northern Kurdish distinguish gender and number, which already correlate the modifiers with their head nouns, and on the other hand, in the majority of Northern Kurdish dialects the primary and secondary \textit{ezāfes} are formally identical. The change in form is an instance of vowel raising (\textit{a} > \textit{e}, \textit{ê} > \textit{î}) that is also observed elsewhere in the morphology of noun phrase (cf. \citealt{HaigÖpengin2018}). 

As for Tsabolov’s main claim regarding word-order change leading to the initial positioning of a possessor modifier in the noun phrase, here too the role of language contact might require revision, since it might have more to do with language-internal organization of morphological material: Zazaki (geographically contiguous with Kurmanji but from a separate historical source to Kurdish), which, like Kurmanji, has gender/number-marking \textit{ezāfe} forms and a case distinction in its nominal system, follows precisely the same word order pattern as Kurmanji in the noun phrase (cf. \citealt{Todd2002}: 95), while Sorani, which has lost gender/number-marking in \textit{ezāfes} and case distinctions in its nominal system, differs from them following the Persian and Middle Persian pattern (cf. \citealt{Öpengin2016}: 61–64). That is, the determining factor seems to be the presence or absence of gender/number-marking \textit{ezāfe} forms, which enable reference tracking between heads and dependents in a noun phrase independently of word order.   

Despite the scepticism one may have towards Tsabolov’s hyopthesis, there is a rather parallel more recent syntactic change in progress stemming from the Arabic influence on the Kurdish of Iraq. This change concerns especially the naming of institutions, such as schools and airports. Recall that in Central Kurdish the possessor in a chain \textit{ezāfe} construction is positioned at the end of the noun phrase, as illustrated in (\ref{bkm:Ref14771780}). However, in the case of these examples, the proper name occurs right after the head noun and before the qualitative modifier, as in (\ref{bkm:Ref14771794}) and (\ref{bkm:Ref14771799}). 

\ea\label{bkm:Ref14771794}Central Kurdish (official signage)\\
\gll qutabxane-y Qemeryan-î seretayî\\
     school-\textsc{ez} \textsc{pn-ez} primary\\
\glt ‘Qamaryan primary school’
\z

\ea\label{bkm:Ref14771799}Central Kurdish (official signage)\\
\gll firokexane-y Hewlêr-î nêwdewletî\\
     airport-\textsc{ez} \textsc{pn-ez} international\\
\glt ‘Hawler international airport’
\z

If the proper name is understood as having the function of possessor here, this is an order that is rather different from the typical Central Kurdish syntax of chain \textit{ezāfe} constructions. But this is precisely the order described for multiple modifier noun phrases of Arabic, as in (\ref{bkm:Ref14771925}). Thus the order in (\ref{bkm:Ref14771799}) is the exact replication of the Arabic version of the same name illustrated in (\ref{bkm:Ref14772143}).

\protectedex{
\ea\label{bkm:Ref14772143}Arabic (official signage)\\
\gll maṭār arbīl ad-dawlī\\
     airport \textsc{pn} \textsc{def}{}-international\\
\glt ‘Erbil international airport’
\z
}

This is clearly a recent imposition from Arabic which does not seem to have gone much beyond naming institutions, especially official signage: the Arabic-like ordering of the name of the airport appears only half as frequently as the inherited order in a Google search. Furthermore, there is no trace of such a word order pattern in the use of Central Kurdish in Iran.  

\subsection{Lexicon}

Arabic influence on Kurdish and all other Near Eastern languages is observed most clearly and abundantly in the vast number of loanwords. According to Perry (\citeyear[97]{Perry2005}), the process of lexical convergence initially took place in Persian between the ninth and thirteenth centuries, when a large number of learned terms were borrowed into literary Persian, and thence transmitted to the other languages of the region. This scenario explains some of the similarities of loanword integration in the two languages (e.g. the borrowing of \textit{tāʔ} \textit{marbūṭa} as \textit{{}-at/-et} (rather than \textit{a}) in a number of words, such as \textit{hukūmat} ‘government’, Persian \textit{hokūmat}, and \textit{quwet} ‘strength’, Persian \textit{qovvat}). However, being spoken in a region that is closer geographically to Arabic-speaking communities, and having had its own educational and religious institutions where Arabic served as the high literary language, Kurdish must have also followed its own course of contact with Arabic. Despite this, there are no studies of lexical borrowing from Arabic into Kurdish. Given the vastness of the topic, with its layers of time-depth and subsantial extra-linguistic aspects, I can only propose here to sketch the major lexical domains of borrowing, and note some observations on the word class and morpho-phonological integration of the borrowings. 

The three major varieties differ in their proportions of borrowing from Arabic. Impressionistically, Northern Kurdish seems to have borrowed most extensively. There is, however, a deeper layer of lexical borrowings shared throughout Kurdish (some of which are common to all or most of the Near Eastern languages), such as the following (cited in their Northern Kurdish forms):\footnote{The main source for the lexical items in this section, together with the information regarding their Arabic origin, is \citet{Chyet2003}. However, I have supplied the interpretation and the discussion of the material and as such only I am responsible for any shortcomings.} 

\ea
\begin{tabbing}
\textit{xiyal} ‘thought, grief’ \hspace{1em} \= Ar. \textit{ḫayāl} ‘imagination’\kill
\textit{xerab} ‘bad’     \> < Ar. \textit{ḫarāb} ‘ruins’\\
\textit{xelk/xelq} ‘people’  \> < Ar. \textit{ḫalq} (\textit{√xlq} ‘to create’)\\
\textit{xiyanet} ‘betrayal’  \> < Ar. \textit{ḫiyāna} \\
\textit{xizêm} ‘nose-ring’  \> < Ar. \textit{ḫizām}\\
\textit{xizmet} ‘service’  \> < Ar. \textit{ḫidma} \\
\textit{ʿeql}/\textit{aqil} ‘reason’  \> < Ar. \textit{ʕaql} (\textit{qəltu} Ar. \textit{ʕaqəl})\\
\textit{qelem} ‘pen’    \> < Ar. \textit{qalam}\\
\textit{quwet} ‘strength’  \> < Ar. \textit{quwwa}\\
\textit{kitêb} ‘book’    \> < Ar. \textit{kitāb}\\
\textit{xiyal} ‘thought, grief’  \> < Ar. \textit{ḫayāl} ‘imagination’\\
\textit{hevîr} ‘dough’    \> < Ar. \textit{ḫamīr} \\
\textit{fikr} ‘thought, idea’  \> < Ar. \textit{fikr}\\
\textit{fêkî/fêqî} ‘fruit’  \> < Ar. \textit{fākiḥa}\\
\textit{ḥal} ‘condition’  \> < Ar. \textit{ḥāl} \\
\textit{ḥazir} ‘ready’    \> < Ar. \textit{ḥāḍir}\\
\textit{şol/şuẍul} ‘work’  \> < Ar. \textit{šuɣl} \\
\textit{terk} ‘abandonment’  \> < Ar. \textit{√trk} ‘to abandon’
\end{tabbing}
\z

Within varieties too, the dialect zones where the communities have had historically closer contact with Arabic-speaking areas show greater Arabic influence in vocabulary. Thus, the dialect of Northern Kurdish named as Southern Kurmanji by  \citet{ÖpenginHaig2014}, spoken around Mardin and Diyarbekir provinces in Turkey, the Jazira province of northeast Syria, and the Sinjar region of Iraq, is the dialect with most extensive Arabic lexical borrowings. Thus, the following items are restricted to this dialect of Northern Kurdish: \textit{tefa-ndin} ‘extinguish-\textsc{tr.inf}’ (from dialectal Ar. \textit{ṭafa} or standard \textit{ṭafiʔa}), \textit{şiteẍl-în} ‘speak-\textsc{intr.inf}’ (from dialectal Ar. \textit{ištaɣal} ‘to work’), \textit{hersim} ‘unripe and sour grapes’ (from Ar. \textit{ḥiṣrim}), \textit{siʿûd} ‘good luck’ (Ar. \textit{suʕūd}, pl. of \textit{saʕd}), \textit{şîret} and \textit{şêwr} ‘advice, counsel’ (Ar. \textit{√šwr}).     

Arabic loanwords in Kurdish belong to various semantic fields, such as kinship, body parts, animals, agriculture, basic tools, temporal concepts and religion. Regarding kinship terms, while the terms for the members of the nuclear family are all inherited, the four second-degree kin terms are all borrowed from Arabic: \textit{met} ‘paternal aunt’ (cf. Ar. \textit{ʕamma(t)}; this item does not exist in Sorani), \textit{xalet/xaltî} ‘maternal aunt’ (Ar. \textit{ḫāla}), \textit{mam} \textit{{\textasciitilde} am} ‘paternal uncle’ (Ar. \textit{ʕamm}), \textit{xal} ‘maternal uncle’ (Ar. \textit{ḫāl}). Considering that the language had its own kin terms before its contact with Arabic, the borrowing of such kin terms constitutes a case of prestige borrowing, probably motivated by the use of such kin words as address forms in the cultures of the region (cf. \citealt{HaigÖpengin2015}).

Similarly, while words for basic animals are inherited, the animals not indigenous to the mountainous region of core Kurdistan are borrowed from Arabic, such as \textit{tîmseḥ} ‘crocodile’ (Ar. \textit{timsāḥ}), \textit{fîl} ‘elephant’ (Ar. \textit{fīl}), \textit{xezal} ‘gazelle, deer’ (Ar. \textit{ɣazāl}). Likewise, the generic term for ‘bird’ or ‘large birds’ is the Arabic loanword \textit{ṭeyr} (Ar. \textit{ṭayr}), while the category word \textit{ferx} ‘young bird/chicken’ is also from Arabic \textit{farḫ}. Several agricultural terms are also borrowed from Arabic, such as \textit{ẓad} ‘grain, food’ (Ar. \textit{zād} 'provisions'), \textit{simbil} ‘spike (of corn or wheat)’ (Ar. \textit{sunbul}), \textit{xox} ‘peach’ (Ar. \textit{ḫawḫ}), \textit{dims} ‘grape molasses’ (Ar. \textit{dibs}). Various terms for spaces and tools of daily life are also borrowed from Arabic, such as \textit{saʿet} ‘hour’ (Ar. \textit{sāʕa}), \textit{sifre} ‘tablecloth’ (Ar. \textit{sufra} ‘dining table’), \textit{qefes} ‘cage’ (Ar. \textit{qafaṣ}), \textit{ḥubr} ‘ink’ (Ar. \textit{ḥibr}), \textit{ḥemam} ‘bath’ (Ar. \textit{ḥammām}), \textit{ḥewş} ‘yard’ (Ar. \textit{ḥawš)}, \textit{meẍmer} ‘velvet’ (Ar. \textit{muḫmal}). Some occupational terms from Arabic are \textit{neqş} ‘embroidery’ (Ar. \textit{naqš} ‘painting, drawing’), \textit{ḥedad} ‘blacksmith’ (Ar. \textit{ḥaddād}), \textit{ʿesker} ‘soldier’ (Ar. \textit{ʕaskar} ‘army’), \textit{tucar} and its older form \textit{têcirvan} (Ar. \textit{tuǧǧār} ‘traders’, sg. \textit{tāǧir}).  

The older layer of administrative and legal terms are predominantly derived from Arabic – though they may have mostly entered via Persian and Ottoman Turkish – such as \textit{sultan} ‘monarch’ (Ar. \textit{sulṭān}), \textit{walî} ‘provincial governor’ (Ar. \textit{wālī}), \textit{muxtar} ‘village chief’ (Ar. \textit{muḫtār}), \textit{ḥukûmet} ‘government’ (Ar. \textit{ḥukūma}), \textit{meḥkeme} ‘court’ (Ar. \textit{maḥkama}), \textit{deʿwā} ‘request, court case’ (Ar. \textit{daʕwa} ‘request, invitation’ and \textit{daʕwā} ‘court case’), \textit{qanûn} ‘law’ (Ar. \textit{qānūn}), \textit{mekteb} ‘school’ (Ar. \textit{maktab} ‘office, desk’).  

As for religious terms, similar to the Persian case (cf. \citealt{Perry2012}: 72), a good number of basic Islamic concepts are inherited, such as the words for god, prophet, angel, devil, heaven, purgatory, prayer, fasting, and sin. In some instances, the Arabic equivalents of these terms exist alongside the inherited ones, restricting the use of the latter, as in the cases of \textit{şeytan} ‘devil’ and \textit{cehnem} ‘hell’, from Arabic \textit{šayṭān} and \textit{ǧahannam}, replacing the Iranian \textit{dêw} and \textit{dojeh}. Many other basic and more peripheral concepts are borrowed from Arabic, such as the following: \textit{xêr} ‘good’ (Ar. \textit{ḫayr}), \textit{xezeb} ‘wrath’ (Ar. \textit{ɣaḍab}), \textit{civat/cemaʿet} ‘society, gathering’ (Ar. \textit{ǧamāʿa}), \textit{ḥec} ‘pilgrimage’ (Ar. \textit{ḥaǧǧ}), \textit{şeytan} ‘devil’ (Ar. \textit{šayṭān}), \textit{weʿz} ‘(Islamic) sermon’ (Ar. \textit{waʿð̣}), \textit{ḥelal} ‘permitted’ (Ar. \textit{ḥalāl}), \textit{ḥeram} ‘forbidden’ (Ar. \textit{ḥarām}), \textit{ruḥ} ‘soul, spirit’ (Ar. \textit{rūh}), \textit{tizbî} (Sorani \textit{tezbêḥ}) ‘prayer beads’ (Ar. \textit{tasbīḥ}).  

Finally, there are also a large number of concepts (temporal, moral, cosmological) that originate from Arabic roots, such as \textit{sibe(h)} ‘morning, tomorrow’ (Ar. \textit{ṣabāḥ}), \textit{heyam} ‘period’ (Ar. \textit{ayyām} ‘days’), \textit{hêsîr} ‘prisoner’ (Ar. \textit{ʔasīr}), \textit{dinya} ‘world’ (Ar. \textit{dunyā}), \textit{ḥesab} ‘count, calculation’ (Ar. \textit{ḥisāb}), \textit{ḥîle} ‘trick, ruse’ (Ar. \textit{ḥīla}), \textit{ḥel}  ‘solution’ (Ar. \textit{ḥall)}, \textit{eşq} ‘love’ (Ar. \textit{ʕišq}), \textit{ʿerz} ‘honor, esteem’ (Ar. \textit{ʕirḍ}). Note also that the word \textit{dinya} is used corresponding to the English expletive subject \textit{it} in time and weather expressions, as in \textit{dinya} \textit{esr} \textit{e} ‘it is late afternoon’ or \textit{dinya} \textit{ewr} \textit{e} ‘it is cloudy’. This usage is noted to exist also in colloquial Arabic \citep[155]{Chyet2003}.

Some other interesting developments with Arabic material in Kurdish lexicon may be noted here. The Arabic \textit{daʕwa} ‘invitation’ has resulted in two related but different concepts: \textit{dawet} ‘wedding ceremony’ and \textit{deʿwet} ‘invitation’. While the latter meaning is shared in Ottoman/Turkish and Persian, the former is a Kurdish-internal semantic expansion of the source meaning. The Kurdish (in all three varieties) word for ‘home’ \textit{mal}, in the sense of family and familial belongings, rather than the house as a structure, is probably derived from the Arabic word \textit{māl} ‘goods, property’. The generic term in Kurdish that designates Christians regardless of their ethnicity and confession is \textit{fileh/file} which derives from Arabic \textit{fallāḥ} ‘peasant, farmer’. Finally, there is the word \textit{mixaletî} ‘the son of the maternal uncle or aunt’ in the southern Kurmanji dialect of Northern Kurdish that can probably be analysed as \textit{mi} (< \textit{ben} ‘son’) + \textit{xalet} ‘aunt’ (< Ar. \textit{ḫāla}) + \textit{î} ‘my’.

Turning now to the word class categories of the loanwords, as has been seen from the presentation of semantic domains above, most Arabic loanwords in Kurdish are nouns. However, many Arabic noun loans are incorporated into Kurdish verb forms. This takes place either through morphological integration or syntactic composition. In morphological integration, the Arabic root (whether nominal or verbal) is taken as the stem onto which the Kurdish verbal suffixes \textit{-în/-îyan} for instransitives and \textit{{}-andin} for transitives are added. Thus the Arabic noun \textit{ʕilm} ‘knowledge’, apart from being used in its nominal sense, serves as the stem for the derivation of the intransitive \textit{ʿelimîn} (\textit{ʿelim-în}) ‘to learn’ and transitive \textit{ʿelimandin} ‘to teach, educate’. The following verbs are further examples of using Arabic roots (whether the original borrowings are nouns or verbs is not always clear) in the derivation of verbs in Kurdish: \textit{tefandin} ‘to extinguish’ (Ar. \textit{ṭafa/ṭafiʔa}), \textit{fetisandin} ‘to suffocate’ (Ar. \textit{faṭṭas}), \textit{fetilîn} ‘to turn around’ (?Ar. \textit{fatala} ‘to twist together’), \textit{qulibîn} ‘to be overturned’ (Ar. \textit{qalaba} ‘to overturn’), \textit{sekinîn} ‘to stand, stop’ (Ar. \textit{√skn} ‘calm, rest’), \textit{fikirîn} ‘to think; to look at’ (Ar. \textit{fikr} ‘thought’).\footnote{Kurdish possesses a number of preverbs such as \textit{ve-} and \textit{ra-}. When inflected with tense--aspect--mood prefixes, these preverbs are detached from the verb stem, as with the verb \textit{ve-kirin} ‘to open’ in \textit{ve-di-ki-m} (\textsc{pvb-ind}{}-do.\textsc{prs-1sg}) ‘I open (it)’. Now, the initial syllable of the verbs \textit{sekinîn} and \textit{fekirîn}, which are based on Arabic loan roots, resemble such Kurdish preverbs. As a result, in some dialects, they are treated as preverbal elements detaching from the verb stem, as with \textit{fe-di-ki-m-ê} ‘I look at it’ (own data, Şirnak area) or \textit{se-di-kin-e} ‘s/he stands’ (own data, from Gevaş), where the initial syllables of originally Arabic roots are reanalysed as preverbs.}  The verb \textit{qelandin} ‘to roast; to uproot’ has two sources as Ar. \textit{qalā} and \textit{qalaʕa,} respectively, which explains its polysemy in Kurdish. 

In syntactic composition, on the other hand, a compound verb\footnote{Here the term \textit{compound} \textit{verb} is employed in a pre-theoretical sense, regardless of whether or not the given complex verb is considered to form a compound. See \citet{Haig2002} for a discussion of complex verbs in Kurdish.}  is obtained by combining an Arabic root with an inherited auxiliary light verb, such as \textit{kirin} ‘do’ or \textit{dan} ‘give’ for transitives, and \textit{bûn} ‘to be’ for instansitives. Thus, the combination of Arabic adjective loanword \textit{xerab} ‘bad’ (< Ar. \textit{ḫarāb} ‘ruin’) with \textit{kirin} yields the verbal meaning ‘to destroy’ while its combination with \textit{bûn} means ‘to go bad, be spoiled’. Some example compound verbs with Arabic roots are given in \REF{compound}.

\ea\label{compound}
\textit{qedr} ‘respect’              (Ar. \textit{qadr})                +   \textit{girtin} ‘to hold’    =  ‘to respect’     \\
\textit{silav} ‘greeting’            (Ar. \textit{salām} ‘peace’)       +   \textit{dan} ‘to give’       =   ‘to greet’      \\
\textit{teʿn\slash ṭan} ‘scolding’   (Ar. \textit{ṭaʕn} ‘piercing’)     + \textit{dan}                   =   ‘to criticize’  \\
\textit{qedeẍe} ‘forbidden’          (Ar. \textit{qadaḥa} ‘to rebuke’)  +   \textit{kirin} ‘to do’       =   ‘to forbid’     \\
\textit{qesd} ‘intention’            (Ar. \textit{qaṣd})                +   \textit{kirin}               =   ‘to head for’   \\
\textit{zeʿîf} ‘weak’                (Ar. \textit{ḍaʕīf})               +   \textit{bûn} ‘to be’         =   ‘to become slim’
\z

What motivates the choice between the morphological versus syntactic technique in the integration of Arabic loan roots in forming verbs in Kurdish is not yet clear. While a few such verbs are found to be used in both synthetic and analytic forms, such as \textit{ceribandin} and \textit{cerebe} \textit{kirin} ‘to try’ (Ar. \textit{< ǧarraba}), most verbs are used in just one of the two forms. However, there is a great deal of dialectal differentiation as to whether a verb is analytically or synthetically integrated. Thus, the morphologically integrated verbs of most Northern Kurmanji dialects such as \textit{emilandin} ‘to use’ (dialectal Ar. \textit{ʕimil} ‘to do’), \textit{şuẍulîn} (Ar. \textit{šuɣl} ‘work’), \textit{fikirîn} (Ar. \textit{fikr} ‘thought’) are seen in the southeastern Badini dialect in synthetic form, with a nominal base combining with a light verb, as \textit{emel} \textit{kirin}, \textit{şol} \textit{kirin}, \textit{fikr} \textit{kirin}.   

There are also various function words (discourse markers, conjunctions, adverbs) which are either borrowed from Arabic or developed in Kurdish based on material borrowed from Arabic. Thus, the conjunction \textit{xeyrî} (also seen as \textit{xeyr} \textit{ji} and \textit{xêncî}) ‘apart from, besides’ is based on Arabic \textit{ɣayr} ‘other than’, while the adversative \textit{emā} ‘but’ is dervied from Arabic \textit{ʔammā} ‘however’. The similative \textit{şibî} (also \textit{şubhetî} and \textit{şitî}) is derived from the Arabic root \textit{√šbh} ‘resemblance’. The classifiers \textit{ḥeb} (and the adverbial \textit{hebekî} ‘a little’) and \textit{lib} are derived from Arabic \textit{ḥabb} ‘grain(s)’ and \textit{lubb} ‘kernels’, respectively. Finally, some discourse and verbal adverbs resulting from Arabic sources are as follows: \textit{meselen} ‘for instance’ and \textit{helbet} ‘of course’ are from Arabic \textit{maθalan} and \textit{al-batta}; in the eastern section of the Badini dialect of Kurmanji, there is the use of the discourse marker \textit{seḥî} ‘apparently, that means’, which is derived from the Arabic \textit{aṣaḥḥ} ‘more correct’ – which separately exists in wider Kurdish as \textit{esseḥ} ‘certainly’; while, finally, the Arabic adjective \textit{qawī} ‘strong’ has evolved into an adverb \textit{qewî} ‘very; very much’ (though this is more literary than spoken).      

All of these lexical borrowings illustrate matter transfer (in the sense of \citealt{MatrasSakel2007}). In the following we have two instances of pattern transfer. First, there is a particular adverbial form \textit{nema} ‘no longer’, found only in the Southeastern dialect of Kurmanji, spoken in the Mardin region of Turkey and Jazira region of northeast Syria. This can be analysed as \textit{ne-ma}, consisting of the negative prefix \textit{ne-} and the past tense 3\textsc{sg} conjugation of the verb \textit{man} ‘to stay’, as in (\ref{bkm:Ref14778636}).

\ea
Southern dialect of Northern Kurdish (Media)\label{bkm:Ref14778636}\footnote{From a poem by an author from Syria, available online at: \url{http://avestakurd.net/blog/2016/10/26/romanivs-kurd-jan-dost-lal-b-ye-vdyo/}} \\
\gll nema di-kar-im veger-im welêt\\
     no.longer \textsc{ind}{}-be.able.\textsc{prs-1sg} return.\textsc{prs.sbjv-1sg} country.\textsc{obl}\\
\glt `I can no longer return to the homeland'\z

There is an immediately corresponding adverbial form \textit{mā} \textit{ʕād} ‘no longer’ in Arabic, which is based on the negative form of the semantically similar verb \textit{ʕād} ‘to return, keep doing’. This is obviously not a very recent development as it is shared in the whole dialect area across country borders, but seemingly not so deep either as to be shared by all Kurdish varieties, not even by all Northern Kurdish dialects, further strengthening the particular status of the Jazira region in Arabic--Kurdish language contact. 

Second, there is a particular lexical construction \textit{bi} \textit{X} \textit{rabûn} ‘to do; to complete; to achieve’ in Northern Kurdish and \textit{hellsan} \textit{be} \textit{X} in Central Kurdish, where \textit{X} stands for any activity or task (usually in the form of an infinitive verb). The construction is based on the verb for ‘to rise, stand’ and a preposition in both varieties, as illustrated in (\ref{bkm:Ref14779023}) and \REF{17}.

\ea Central Kurdish (Media)\label{bkm:Ref14779023}\footnote{URL of article: \url{http://www.kurdistan24.net/so/news/5ca67132-7a7f-4840-bfb4-dea5bf25ea2e}}\\
\gll polîs hellsa be kokirdinewe-y zanyarî\\
     police rise.\textsc{pst.3sg} with collecting-\textsc{ez} information\\
\glt ‘The police undertook (the task of) collecting information.’\z

\ea\label{17} Northern Kurdish (Media)\footnote{URL of article: \url{http://portal.netewe.com/mir-celadet-bedirxan-bi-tene-sere-xwe-bi-kare-dewleteke-rabu/}}\\
\gll Mîr {Celadet (…)} bi kar-ê dewlet-ek-ê rabû\\
     Emir Celadet with work-\textsc{ez.m} state-\textsc{indf-obl.f} rise.\textsc{pst.3sg}       \\
\glt ‘Emir Celadet undertook the work of a state.’\z

This lexical construction also has a parallel in Modern Standard Arabic, based on the verb \textit{qāma} ‘to stand (up)’ and the preposition \textit{bi} ‘with’, with the collocation \textit{qāma} \textit{bi} meaning ‘to undertake’. This is obviously a recent influence on Kurdish, as it is seen only in Iraq and Syria, and in a manner cross-cutting the broad variety borders between Sorani and Kurmanji. 

\section{Conclusion}
Contact with Arabic, which started in the early medieval period (approx. 7th–8th centuries) with the arrival of Islam in the Near East, has had a profound impact on Kurdish, particularly on its lexicon and phonology. Given the total absence of any substantial previous study on the matter, the present chapter provides a first assessment of the influence of Arabic on Kurdish, primarily as represented in Kurdish phonology and lexicon but also, albeit more restrictedly, in morphology and syntax. Kurmanji Kurdish seems to be the variety that is most affected by contact with Arabic, which is understandable considering the geographical continuity of the Kurdish and Arabic communities, especially in the historical Jazira region and more widely in Upper Mesopotamia (in Mardin–Diyarbekir, Mosul–Sinjar, and Haseke province). There are thus areas which show more intensive Arabic influence within the speech zones of major Kurdish varieties, while the outcomes of the contact reflect different layers in terms of time depth. Accordingly, the deeper-layer influence comes in the form of lexical convergence with Arabic, sometimes through the intermediary of Persian and/or Ottoman Turkish. This contact has repercussions in the expansion of the phonological inventory of the language, and is shared across most Kurdish varieties. There are no unquestionably demonstrated changes in the morphosyntax resulting from contact with Arabic at this layer. At the relatively shallower layer, the influence is mainly seen in Syria and Iraq, and in the form of further expansion of the phonological inventory and a vocabulary heavily lexified by Arabic roots incorporated also into the verbal domain. There are also several cases illustrating morphosyntactic and lexicosyntactic change, such as the default gender assignment and word order in complex noun phrases, as well as certain phrasal and adverbial lexical items. 

In terms of “cognitive dominance”, in the sense of \citet{VanCoetsem1988,VanCoetsem2000} and \citet{Lucas2015}, in these instances of contact influence, the deeper-layer influence, which is restricted to, or related to, lexical borrowing, takes place with the speakers being cognitively dominant in the recipient language, Kurdish. The more recent instances of heavy lexification, and morphosyntactic and lexicosyntactic changes may, however, be the result of imposition, where the speakers are dominant in the source language. 

These outcomes may also be related to bilingualism and language configuration in historical perspective. That is, the absence of imposition (in the form of morphosyntactic changes) in the deeper historical layer, and the restriction of the influence to lexicon, point to the absence of widespread Arabic–Kurdish bilingualism among the speakers of Kurdish at those historical stages. Some imposition of this kind is observed in the Kurmanji of the Jazira region, which is known to have had greatest speaker contact between Kurdish and Arabic speech communities. By contrast, the widespread bilingualism and Arabic-dominant linguistic configuration in Syria and Iraq for at least a century has led to instances of imposition where the morphosyntactic and lexical patterns of Arabic are replicated in Kurdish. These outcomes are also mostly consonant with the predictions of Van Coetsem’s (\citeyear{VanCoetsem1988,VanCoetsem2000}) “stability gradient”, which argues that lexicon is less stable than syntax and phonology, which require dominance in the source language in order to be affected by contact-induced change.    

Given the limitations of a first attempt, much is yet to be explored regarding Kurdish–Arabic language contact. In particular, the precise paths of development of pharyngeals and emphatics in Kurdish should be analysed through fieldwork-based comparative dialect data, while, in the domain of lexicon, it is important to analyse the morphophonological integration of borrowings into Kurdish. It is also of interest to be able to develop diagnostics to disentangle direct Arabic influence on Kurdish from influence via other major languages such as Persian and Ottoman Turkish. Finally, a detailed account of the history of Kurdish–Arabic socio-political and cultural contact is of utmost importance in order to complement the linguistic data and enable a more fine-grained analysis of the agentivity of contact-induced change in Kurdish.      

\section*{Further reading}
\citet{Barryforthcoming} is a comprehensive and theoretically grounded treatment of the introduction and further propagation of pharyngeal sounds in Kurdish.\\ 
\citet{Chyet2003} is the most comprehensive Kurdish–English dictionary,  providing information on the source language of most loanwords in Kurdish, including those from Arabic.\\
\citet{Tsabolov1994} is the only work published so far on Arabic influence on the grammar of Kurdish. 

\section*{Acknowledgements}
I would like to thank the editors of the volume and an anonymous reviewer for their helpful and detailed feedback. Thanks also to Adam Benkato for his help with Arabic data. Only I am responsible for any remaining shortcomings and errors. 

\section*{Abbreviations}
\begin{multicols}{2}
\begin{tabbing}
\textsc{ipfv} \hspace{1em} \= before common era\kill
Ar.              \> Arabic\\
BCE             \> before Common Era\\
ca.             \> circa\\
\textsc{def}    \> definite \\
\textsc{drct}   \> directional\\
dial.           \> dialectal \\
\textsc{ez}     \> \textit{ezāfe} \\
\textsc{f}      \> feminine \\
\textsc{gen}    \> genitive\\
Kr.             \> Kurdish\\
\textsc{ind}    \> indicative \\
\textsc{indf}   \> indefinite \\
\textsc{inf}    \> infinitive\\
\textsc{intr}   \> intransitive\\
\textsc{ipfv}   \> imperfective \\
\textsc{m}      \> masculine \\
\textsc{neg}    \> negative\\
\textsc{nom}    \> nominative\\
\textsc{obl}    \> oblique\\
\textsc{pl/}pl. \> plural \\
\textsc{pn}     \> proper noun\\
\textsc{poss}   \> possessive \\
\textsc{prs}    \> present \\
\textsc{pst}    \> past \\
\textsc{pvb}    \> preverb\\
RL              \> recipient language\\
\textsc{sbjv}   \> subjunctive \\
\textsc{sec}    \> secondary or pronominal \\ \> \textit{ezāfe}/linking element\\
\textsc{sg/}sg. \> singular\\
SL              \> source language\\
\textsc{tr}     \> transitive\\
Tr.             \> Turkish
\end{tabbing}
\end{multicols}


{\sloppy\printbibliography[heading=subbibliography,notkeyword=this]}
\end{document}
