\documentclass[output=paper]{langsci/langscibook} 
\author{Catherine Taine-Cheikh\affiliation{CNRS, LACITO}}
\title{Ḥassāniyya Arabic}
\abstract{The area where Ḥassāniyya is spoken, located on the outskirts of the Arab world, is contiguous with those of several languages that do not belong to the Afro-Asiatic phylum. However, the greatest influence on the evolution of Ḥassāniyya has been its contact with Berber and Classical Arabic. Loanwords from those languages are distinguished by specific features that have enriched and developed the phonological and morphological system of Ḥassāniyya. In other respects, Ḥassāniyya and Zenaga are currently in a state of either parallel evolution or reciprocal exchanges. }
\IfFileExists{../localcommands.tex}{
 \input{../localpackages}
 \input{../localcommands} 
 \input{../localhyphenation} 
 \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
 

\section{Current state and historical development}

\subsection{Historical development of Ḥassāniyya} %1.1. /

The arrival in Morocco of the Banī Maʕqil, travelling companions of the Banī Hilāl and Banī Sulaym, is dated to the thirteenth century. However, the gradual shift to the territories further south of one of their branches – that of the Banī Ḥassān, the origin of the name given to the dialect described here – began closer to the start of the subsequent century. 

At that time, the Sahel region of West Africa was inhabited by different communities: on the one hand there were the “white” nomadic Berber-speaking tribes, on the other hand, the sedentary “black” communities.

Over the course of the following centuries, particularly during the seventeenth and eighteenth centuries, the sphere of Zenaga Berber gradually diminished, until it ceased to exist in the 1950s, other than in a few tribes in the southwest of Mauritania. At the same time, Ḥassāniyya Arabic became the language of the nomads of the west Saharan group, maintaining a remarkable unity (\citealt{Taine-Cheikh2016,Taine-Cheikh2018historical}). There is virtually no direct documentation of the region’s linguistic history during these centuries. This absence of information itself suggests a very gradual transformation and an extended period of bilingualism.

Despite the lack of documentation of the transfer phenomenon, it seems highly likely that bilinguals played a very important role in the changes described in this chapter. 

\subsection{Current situation of Ḥassāniyya} %1.2. /

The presence of significant Ḥassāniyya-speaking communities is recognized in six countries. With the exception of Senegal and especially of Niger, the regions occupied by these communities, more or less adjacent, are situated primarily in Mauritania, in the north, northeast and east of the country. 

The greatest number of Ḥassāniyya-speakers (approximately 2.8 out of a total of four million) are found in Mauritania, where they constitute the majority of the population (approximately 75\%). The Ḥassāniyya language tends to fulfil the role of the lingua franca without, however, having genuine official recognition beyond, or even equal to, that which it has acquired (often recently) in neighbouring countries.

\section{Contact languages} %2. /

\subsection{Contact with other Arabic varieties} %2.1. /

The Islamization of the Ḥassāniyya-speaking population took place at an early date, and Ḥassāniyya has therefore had lengthy exposure to Classical Arabic. For many centuries this contact remained superficial, however, except for among the Marabout tribes, where proficiency in literary Arabic was quite widespread and in some cases almost total. The teaching of Islamic sciences in other places reached quite exceptional levels in certain \textit{mḥāð̣ə{\R}} (a type of traditional desert university).\footnote{These may be referred to as universities both in terms of the standard of teaching and the length of students’ studies. They were, however, small-scale, local affairs, located either in nomadic encampments or in ancient caravan cities.} In the post-colonial era, the choice of Arabic as official language, and the widespread Arabization of education, media and services, greatly increased the Ḥassāniyya-speaking population’s contact with literary Arabic (including in its Modern Standard form), though perfect fluency was not achieved, even among the young and educated populations. 

Excluding the limited influence of the Egyptian and Lebanese–Syrian dialects used by the media, the Arabic dialects with which Ḥassāniyya comes in to contact most often today are those of the neighbouring countries (southern Moroccan and southern Algerian). Most recently Moroccan koiné Arabic has established a presence in the Western Sahara, since the region came under the administration of Morocco.

\subsection{Contact with Berber languages} %2.2. /

Ḥassāniyya has always been in contact with Berber languages. Currently, speakers of Ḥassāniyya are primarily in contact with Tašlḥiyt (south Morocco), Tuareg (Malian Sahara and the Timbuktu region) and Zenaga (southwest Mauritania). In these areas, some speakers are bilingual in Ḥassāniyya and Berber. 

In Mauritania, where Zenaga previously occupied a much larger area, Berber clearly appears as a substrate.

\subsection{Contact with languages of the Sahel} %2.3. /

Contacts between Ḥassāniyya speakers and the languages spoken in the Sahel have varied across regions and over time, but have left few clearly discernible traces on Ḥassāniyya.

The contact with Soninke is ancient (cf. the toponym \textit{Chinguetti} < Soninke \textit{sí-n-gèdé} ‘horse well’), but the effects are hardly noticeable outside of the old cities of Mauritania. The contact with Songhay is both very old and still ongoing, but is limited to the eastern part of the region in which Ḥassāniyya is spoken (especially the region of Timbuktu).

The influence of Wolof, albeit marginal, has always been more substantial in southwestern Mauritania, especially among the Awlād Banʸūg of the Rosso region. It peaked in the years 1950–70, in connection with the immigration to Senegal of many Moors (e.g. \textit{gordʸigen} ‘homosexual’, lit. ‘man-woman’). In Mauritania, the influence of Wolof can still be heard in some areas of urban crafts (e.g. mechanics, electricity), but it is primarily a vehicle for borrowing from French.

Although Pulaar speakers constitute the second-largest linguistic community of Mauritania, there is very limited contact between Ḥassāniyya and Pulaar, with the exception of a few bilingual groups (especially among the Harratins) in the Senegal River valley.

Certain communities (particularly among the Fulani) were traditionally known for their perfect mastery of Ḥassāniyya. As a result of migration into major cities and the aggressive Arabization policy led by the authorities, Ḥassāniyya has gained ground among all the non-Arabic speakers of Mauritania (especially in the big cities and among younger people), but this has come at the cost of a sometimes very negative attitude towards the language. 

\subsection{Contact with Indo-European Languages} %2.4. /

Exposure to French has prevailed in all the countries of the region, the only exception being the Western Sahara, which, from the end of the nineteenth century until 1975, was under Spanish occupation.

In Mauritania the French occupation came relatively late and was relatively insignificant. However, the influence of the colonizers’ language continued well after the country proclaimed its independence in 1960. That said, it has tended to regress since the end of the twentieth century (especially with the rise of Standard Arabic, e.g. \textit{minəstr} has been replaced by \textit{wazīr} ‘minister’), whilst exposure to English has become somewhat more significant, at least in the better educated sections of the population.

\section{Contact-induced changes in  Ḥassāniyya} %3. /

\subsection{Phonology} %3.1. /

\subsubsection{Consonants} %3.1.1. /
\subsubsubsection{The consonant /ḍ/} %3.1.1.1. / 
As in other Bedouin dialects, /ð̣/ is the normal equivalent of the 〈\kern 0.75pt{\arabscript{ض}}〉 of Classical Arabic (e.g. \textit{ð̣mər} ‘to have an empty stomach’ (CA \textit{ḍamira}) and \textit{ð̣ḥak} ‘to laugh (CA \textit{ḍaḥika}). Nonetheless, /ḍ/ is found in a number of lexemes in Ḥassāniyya. 

The form [ḍ] sometimes occurs as a phonetic realization of /\kern -0.5ptd/ simply due to contact with an emphatic consonant (compare \textit{ṣḍam} ‘to upset’ and \textit{ṣadma} ‘annoyance’, CA \textit{√ṣdm}). However, /ḍ/ generally appears in the lexemes borrowed from Standard Arabic, either in all words of a root, or in a subset of them, for example: \textit{staḥḍa{\R}} ‘to be in agony’ and \textit{ḥaḍari} ‘urbanite’ but \textit{ḥð̣a{\R}} ‘to be present’ and \textit{maḥ\textsuperscript{ə}\kern -0.75ptð̣{\R}a} ‘Quranic school’. The opposition /ḍ/ vs. /ð̣/ can therefore distinguish a classical meaning from a dialectal meaning: compare \textit{staḥḍa{\R}} to \textit{staḥð̣a{\R}} ‘to remember’.

/ḍ/ is common in the vocabulary of the literate. The less educated speakers sometimes replace /ḍ/ with /ð̣/ (as in \textit{qāð̣i} for \textit{qāḍi} ‘judge’), but the stop realization is stable in many lexemes, including in loanwords not related to religion, such as \textit{ḍʕīv} ‘weak’.

The presence of the same phoneme /ḍ/ in Berber might have facilitated the preservation of its counterpart in Standard Arabic loans, even though in Zenaga /ḍ/ is often fricative (intervocalically). Moreover, the /ḍ/ of Berber is normally devoiced in word-final position in Ḥassāniyya, just as in other Maghrebi dialects, for example: \textit{ṣayvaṭ} ‘to say goodbye’, from Berber \textit{√fḍ} ‘to send’.

\subsubsubsection{The consonant /ẓ/} %3.1.1.2. / 
/ẓ/ is one of the two emphatic phonemes of proto-Berber. This emphatic sibilant sound regularly passes from the source language to the recipient language when Berber words are used in Ḥassāniyya. For example: \textit{aẓẓ} \~{} \textit{āẓẓ} ‘wild pearl millet’ (Zenaga \textit{īẓi}).

However, /ẓ/ is also present in lexemes of a different origin. Among Ḥassāniyya roots also attested in Classical Arabic, *z often becomes /ẓ/ in the environment of /{\R}/ (e.g. \textit{{\R}āẓ} ‘to try’, CA \textit{rāza}; \textit{{\R}aẓẓa} ‘lightning’, CA \textit{rizz}; \textit{ẓəb{\R}a} ‘anvil’, CA \textit{zubra}). Sometimes /ẓ/ appears in lexemes with a pejorative connotation, e.g. \textit{ẓ{\R}aṭ} ‘fart; lie’ (CA \textit{ḍaraṭa}), \textit{ẓagg} ‘make droppings (birds)’ (CA \textit{zaqq}).

\subsubsubsection{The consonant /q/} %3.1.1.3. / 
The normal equivalent of the 〈{\arabscript{ق}}〉 of Classical Arabic is the velar stop /g/, as in other Bedouin dialects (e.g. \textit{bag{\R}a} ‘cow’, CA \textit{baqara}). However, /q/ is in no way rare.

First of all, /q/ appears, like /ḍ/, in a number of words borrowed from Classical Arabic by the literate: \textit{ʕaq\kern 0.5pt\textsuperscript{ə}\kern -1.5ptd} ‘religious marriage contract’; \textit{vassaq} ‘to pervert’. The opposition /g/ vs. /q/ can therefore produce two families of words, such as \textit{qibla} ‘Qibla, direction of Mecca’ and \textit{gəbla} ‘one of the cardinal directions (south, southwest or west, depending on the region)’. It can also create a distinction between the concrete meaning (with /g/) and the abstract meaning (with /q/): \textit{θgāl} ‘become heavy’, \textit{θqāl} ‘become painful’.

Next, /q/ is present in several lexemes of non-Arabic origin, such as \textit{bsaq} ‘silo’, \textit{mzaw{\R}aq} ‘very diluted (of tea)’, (in southwest Mauritania) \textit{sə{\R}qəlla} ‘Soninke people’, (in Néma) \textit{sasundaqa} ‘circumcision ceremony’, (in Walata) \textit{raqansak} ‘decorative pattern’, \textit{asanqās} ‘pipe plunger’, \textit{sayqad} ‘shouting in public’, and (in the southeast) \textit{šayqa} ‘to move sideways’. These lexemes, often rare and very local in use, seem to be borrowed mostly from the languages of the Sahel region.\footnote{I am currently unable to specify the origin of these terms except that \textit{bsaq} (attested in Zenaga) could be of Wolof origin.}

Finally, /q/ is the outcome of *ɣ in cases of gemination, (/ɣɣ/ > [qq]): compare \textit{raqqad} ‘to make porridge’ to \textit{rɣīda} ‘a variety of porridge’ (CA \textit{raɣīda}). This correlation, attested in Zenaga and more generally in Berber, can be attributed to the substrate.

Insofar as the contrast between /ɣ/ and /q/ is poorly established in Berber, the substrate could also explain the tendency, sometimes observed in the southwest, to velarize non-classical instances of /q/ (or at least instances not identified as classical): hence \textit{ɣandīr} ‘candle’ for \textit{qandīr} < CA \textit{qandīl} – this is despite the fact that the shift /ɣ/ > /ʔ/ is very common in Zenaga. However, the influence of Berber does not explain the systematic shift of /ɣ/ to /q/ throughout the eastern part of the Ḥassāniyya region (including Mali): thus eastern \textit{qlab} ‘defeat’ for southwestern \textit{ɣlab} (CA \textit{ɣalaba}).\footnote{The regular passage from /ɣ/ to /q/ is a typical Bedouin trait, related to the voiced realization (/g/) of *q. It occurs especially in southern Algeria, in various dialects of the Chad–Sudanese area, and in some Eastern dialects \citep[72]{Cantineau1960book}.}

\subsubsubsection{Glottal stop} %3.1.1.4. / 
The glottal stop is one of the phonemes of Zenaga (its presence in the language is in fact a feature that is unique among Berber varieties), however it is not found in Ḥassāniyya, with the exception of words borrowed from Standard Arabic, e.g. \textit{tʔabbad} ‘to live religiously’, \textit{danāʔa} ‘baseness’ and \textit{taʔḫīr} ‘postponement’. Very rarely the glottal stop is also maintained when it occurs at the end of a word as in \textit{ba{\R}{\R}aʔ} ‘to declare innocent’.

\subsubsubsection{Palatalized consonants} %3.1.1.5. / 
There are three palatalized consonants: two dental (/tʸ/ and /dʸ/) and a nasal /nʸ/. Unlike the phonemes discussed above, these are very rare in Ḥassāniyya, especially /nʸ/.

The palatalized consonants are also attested in certain neighbouring languages of the Sahel, as well as in Zenaga (but these are not phonemes of Common Berber). They are rather infrequent in the Zenaga lexicon, occurring especially in syntagmatic contexts (\textit{{}-d}+\textit{y-}, \textit{{}-n}+\textit{y-}) and in morphological derivation (formation of the passive by affixation of a geminate /tʸ/).

In Ḥassāniyya, the palatalized consonants mostly appear in words borrowed from Zenaga or languages of the Sahel. Interestingly, certain loanwords from Zenaga are ultimately of Arabic origin and constitute examples of phonological integration, as in \textit{tʸfāɣa}, a given name and, in the plural, the name of a tribe < Zenaga \textit{atʸfāɣa} ‘marabout’ < CA \textit{al-faqīh}, and \textit{ḫurūdʸ} ‘leave (from Quranic school)’ < Zenaga \textit{ḫurūdʸ} < CA \textit{ḫuruǧ} ‘exit’.

One should also note the palatization of /t/ in certain lexemes from particular semantic domains (such as the two verbs related to fighting \textit{tʸbəl} ‘to hit hard’ and \textit{kawtʸam} ‘boxer’). This may suggest the choice of a palatalized consonant for its expressive value (and would then be a marginal case of phonosymbolism).

\subsubsubsection{Labial and labiovelar consonants} %3.1.1.6. / 
The labiovelar consonants /mʷ, bʷ, f\kern 1ptʷ, vʷ/ or /ṃ, ḅ, f̣, ṿ/) are common in Ḥassāniyya, as they are in Zenaga. In both cases, they often come in tandem with a realization [u] of the phoneme /ə/. 

This phenomenon may have originally arisen in Zenaga, since the Ḥassāniyya of Mali (where it was most likely in contact with other languages) exhibits greater preservation of a [u] vowel sound and, at the same time, less pronounced labiovelarization of consonants. 

The Ḥassāniyya of Mali also has a voiceless use of the phoneme /f/, where the Ḥassāniyya of Mauritania is characterized by the use of /v/ in its place (\citealt{Heath2004}; an observation that my own studies have confirmed). This phonetic trait does not come directly from Zenaga (in which /v/ exists but is very rare). However, it could be connected with the preference for voiced phonemes in Berber generally and in Zenaga in particular.

\subsubsection{Syllabic structures} %3.1.2. /

In Ḥassāniyya, Arabic-derived syllabic structures do not contain short vowels in word-internal open syllables, with the exception of particular cases such as passive participles in \textit{mu-} (\textit{mudagdag} ‘broken’) and certain nouns of action (\textit{ḥašy} > \textit{ḥaši} ‘filling’). However, loanwords from literary Arabic and other languages (notably Berber and French) display short vowels quite systematically in this context: \textit{abadan} ‘never’ and \textit{ḥazīn} ‘sad’ (from Standard Arabic); \textit{tamāt} ‘gum’ (from Zenaga \textit{taʔmað}); \textit{taṃāta} ‘tomato’. In fact, it may be noted that, unlike the majority of Berber varieties (particularly in the north), Zenaga has a relatively substantial number of lexical items with short vowels (including \textit{ə}) in open syllables: \textit{ka{\R}að̣} ‘three’, \textit{tuðu}\textit{ṃaʔn} ‘a few drops of rain’ \textit{awayan} ‘languages’, \textit{əgəðih} ‘necklace made from plants'.\footnote{It is precisely for this reason that, regarding the loss of the short vowels in open syllables, I deem the hypothesis of a parallel evolution of syllabic structures in Maghrebi Arabic and Berber to be more convincing than the frequently held alternative hypothesis of a one-way influence of the Berber substrate on the Arabic adstrate.} 

Furthermore, a long vowel \textit{ā} occurs word-finally in loaned nouns which in Standard Arabic end with \textit{{}-āʔ}: \textit{vidā/vidāy} ‘ransom’. In other cases, underlyingly long word-final vowels are only pronounced long when non-final in a genitive construct.

\subsection{Morphology} %3.2. /

\subsubsection{Nominal morphology} %3.2.1. /

\subsubsubsection{Standard forms} %3.2.1.1. /

Nouns and adjectives borrowed from Standard Arabic may often be identified by the presence of: a) open syllables with short vowels, e.g. \textit{vaḍa}\textit{lāt} ‘rest of a meal’, \textit{ɣaḍab} ‘anger’, \textit{vasād} ‘alteration’, \textit{ḥtimāl} ‘possibility’, b) short vowels /i/ (less frequently /u/) in a closed syllable: \textit{miḥ{\R}āb} ‘mihrab’, \textit{muḥarrir} ‘inspector; editor’.

Some syllables are only attested in loanwords, such as the nominal pattern CVCC, where the pronunciation of the double coda necessitates the insertion of a supporting vowel, in which case the dialect takes on the form CCVC: compare \textit{ʕaq\kern 0.5pt\textsuperscript{ə}\kern -1.5ptd} ‘religious marriage’ with \textit{ʕqal} ‘wisdom’.

The most characteristic loanword pattern, however, is that of \textit{taḥrīr} ‘liberation; verification (of an account)’. In Ḥassāniyya the equivalent of the pattern taCCīC is təCCāC. For the root \textit{√ḥrr}, this provides a verbal noun for other meanings of the verb \textit{ḥa{\R}{\R}a{\R}:} \textit{təḥ{\R}ā{\R}} ‘whipping of wool (to untangle it); adding flour to make dumplings’. As for the form taCaCCuC, the /u/ is sometimes lengthened: \textit{taḥammul} ‘obligation’, but \textit{tavakkūr} ‘contemplation’.

\subsubsubsection{Berber affixes} %3.2.1.2. /

Nouns borrowed from Berber are characterized by the frequent presence of the vowels /a, ā, i, ī, u, ū/. These are of varying lengths, except that in a word-final closed syllable they are always long and stressed. Since these vowels appear in all types of syllables – open and closed – this results in much more varied syllabic patterns than in nouns of Arabic origin. 

These loans are also characterized by the presence of affixes which, in the source language, are markers of gender and/or number: the prefix \textit{a/ā-} or \textit{i/ī-} for the masculine, to which the prefix \textit{t-} is also added for the feminine or, more frequently (especially in the singular), a circumfix \textit{t-...-t}. Compare \textit{iggīw} \~{} \textit{īggīw} ‘griot’ with the feminine form \textit{tiggiwīt} \~{} \textit{tīggīwīt}. A suffix in \textit{-(ə)n} characterizes the plurals of these loanwords which, moreover, differ from the singulars in terms of their vocalic form: \textit{iggāwən} \~{} \textit{īggāwən} ‘griots’, feminine \textit{tiggawātən} \~{} \textit{tīggawātən}. The presence of these affixes generally precludes the presence of the definite article.

Though these affixes pass from the source language to the target language along with the stems, the syllabic and vocalic patterns of such loans are often particular to Ḥassāniyya: compare Ḥassāniyya \textit{āršān}, plural \textit{īršyūn} \~{} \textit{īršīwən} ‘shallow pit’ with Zenaga \textit{aʔraš}, plural \textit{aʔraššan} (see \citealt{Taine-Cheikh1997Zenaga}).

Ḥassāniyya speakers whose mother tongue is Zenaga have most likely played a role in the transfer of these affixes and their affixation to nouns of all origins (including those of Arabic origin: a possible example being \textit{tasūvra} ‘large decorated leather bag for travelling’, cf. \textit{sāvər} ‘to travel’). The forms that these speakers use can also be different from those used by other Ḥassāniyya speakers – especially if the latter have not been in contact with Berber speakers for a long time. 

It is not proven that Berber speakers are the only ones to have created and imposed these forms which are more Berberized than authentically Berber. However, it may be noted that the gender of nouns borrowed from Berber is generally well preserved in Ḥassāniyya, even for the feminine nouns losing their final \textit{{}-t}, other than in special cases such as the collective \textit{tayšəṭ} ‘thorny tree (\textit{Balanites} \textit{aegyptiaca})’ with a final \textit{{}-ṭ} (< Zenaga \textit{tayšaḌ} for \textit{tayšaḍt}).\footnote{In Zenaga, non-intervocalic geminates are distinguished not by length, but rather by tension, and it is this that is indicated by the use of uppercase for the final \textit{Ḍ.} } In fact, this indicates a deep penetration of the meaning of these affixes and of Berber morphology in general (up to and including the incompatibility of these affixes with the definite article). 

The borrowing of the formants \textit{ən-} ‘he of’ and \textit{tən-} ‘she of’ (quasi-equivalents of the Arabic-derived \textit{bū-} and \textit{ūṃ(ṃ)-}) is fairly widespread, in particular in the formation of proper nouns. It is also mostly in toponyms and anthroponyms that the diminutive form with prefix \textit{aɣ-} and suffix \textit{{}-t} is found, e.g. the toponym \textit{Agjoujt} (< \textit{aɣ-žoʔž-t} ‘small ditch’).

\subsubsection{Verbal Morphology} %3.2.2. /

\subsubsubsection{The derivation of \textit{sa-}}\label{sa}

The existence of verb forms with the prefix \textit{sa-} is one of the unique characteristics of Ḥassāniyya (\citealt{Cohen1963,Taine-Cheikh2003}). There is nothing, however, to indicate that the prefix is an ancient Semitic feature that Ḥassāniyya has preserved since its earliest days. Instead, the regular correspondences between the three series of derived verb forms (causative--factitive vs. reflexive vs. passive) and the specialization of the morpheme \textit{t} as a specific marker of reflexivity underlie the creation of causative--factitives with \textit{sa-}. Neologisms with \textit{sa-} generally appear when forms with the prefix \textit{sta-} have a particular meaning: \textit{staslaʕ} ‘to get worse (an injury)’ – \textit{saslaʕ} ‘to worsen (injury)’; \textit{stab{\R}ak} ‘to seek blessings’ – \textit{sab{\R}ak} ‘to give a blessing’; \textit{stagwa} ‘to behave as a griot’ – \textit{sagwa} ‘to make someone a griot’; \textit{staqbal} ‘to head towards the Qibla’ – \textit{saqbal} ‘to turn an animal for slaughter in the direction of the Qibla’. 

Furthermore, the influence of Berber has certainly played a role, since the prefix \textit{s(a)-} (or one of its variants) very regularly forms the causative--factitive structure in this branch of the Afro-Asiatic language family. 

In Zenaga, the most frequent realization of this prefix is with a palato-alveolar shibilant, but a sibilant realization also occurs, particularly with roots of Arabic origin. For example: Hass. \textit{sādəb} (variant of \textit{ddəb}) – Zen. \textit{yassiʔðab} ‘to train an animal (with a saddle)’ < CA \textit{√ʔdb} (cf. \textit{ʔaddaba} ‘educate, carefully bring up’); Hass. \textit{sasla} \textit{–} Zen. \textit{yassaslah} ‘to let a hide soak to give it a consistency similar to a placenta’ and Hass. \textit{stasla} – Zen. \textit{staslah} ‘start to lose fur (of hides left to soak)’ < CA \textit{√sly} (cf. \textit{salā} ‘placenta’).

Parallel to these examples where the Berber forms (at least those with the prefix \textit{st(a)}{}-) are most likely themselves borrowed, we also find patterns with \textit{sa-/ša-} which are incontestably of Berber origin: compare Ḥassāniyya \textit{niyyər} ‘to have a good sense of direction’, \textit{sanyar} ‘to show the way’, \textit{stanyar} ‘to know well how to orient oneself’ and Tuareg \textit{ener} ‘to guide’, \textit{sener} ‘to make guide’. Typically, however, when Ḥassāniyya borrows causative forms from Berber, it usually integrates the Berber prefix as part of the Ḥassāniyya root, making it the first radical of a quadriliteral root, e.g. Hass. \textit{sadba} \textit{–} Tuareg \textit{sidou} ‘to make s.o. leave in the afternoon’ and Hass. \textit{ssadba} \textit{(<} \textit{tsadba)} – Tuareg \textit{adou} ‘to leave in the afternoon’. 

The parallelism between Arabic and Berber is not necessarily respected in all cases, but the forms with initial \textit{s-/š-} are usually causative or factitive in both cases. The only exception concerns certain Zenaga verbal forms which have become irregular upon contact with Ḥassāniyya: thus \textit{yassəðbah} ‘to leave in the afternoon’ or \textit{yišnar} ‘to orient oneself’ (a variant of \textit{yinar}), of which the original causative value is now carried by a form with a double prefix (\textit{ž}+\textit{š}): \textit{yažəšnar} ‘to guide’.

\subsubsubsection{The Derivation of \textit{u-}} %3.2.2.2. /

The existence of a passive verbal prefix \textit{u-} for quadrilateral verbs and derived forms constitutes another unique feature of Ḥassāniyya. For example: \textit{udagdag}, passive of \textit{dagdag} ‘to break’; \textit{uṭabbab}, passive of \textit{ṭabbab} ‘to train (an animal)’; \textit{udāɣa}, passive of \textit{dāɣa} ‘to cheat (in a game)’.

The development of passives with \textit{u-} was most likely influenced by Classical Arabic, since here the passives of all verbal measures feature /u/ in the first syllable in both the perfect and the imperfect, e.g. \textit{fuʕila}, \textit{yu}\textit{fʕalu}; \textit{fuʕʕila}, \textit{yu}\textit{faʕʕalu}; and \textit{fūʕila}, \textit{yu}\textit{fāʕalu}, the respective passives of \textit{faʕala}, \textit{faʕʕala} and \textit{fāʕala}.

However, influence from Berber cannot be excluded here since, in Zenaga, the formation of passives with the prefix \textit{Tʸ} is directly parallel to those of the passives with \textit{u-} in Ḥassāniyya. Moreover, this prefix is \textit{t(t)u-} or \textit{t(t)w-} in other Berber varieties (especially those of Morocco) and this could also have had an influence on the emergence of the prefix \textit{u-}. 

\subsection{Syntax} %3.3. /

\subsubsection{Ḥassāniyya–Zenaga parallelisms} %3.3.1. /

Ḥassāniyya and Zenaga have numerous common features, and this is especially true in the realm of syntax. In general, the reason for these common traits is that they both belong to the Afro-Asiatic family and remain conservative in various respects; for example, in their lack of a discontinuous negative construction.

There are, however, also features of several varieties of both languages documented in Mauritania that represent parallel innovations. Thus, corresponding to the diminutive forms particular to Zenaga, we have in Ḥassāniyya \textit{mutatis} \textit{mutandis} a remarkably similar extension to verbs of the diminutive pattern with infix \textit{-ay-}, e.g. \textit{mayllas}, diminutive of \textit{mallas} ‘to smooth over’ (\citealt{Taine-Cheikh2008chapter}: 123–124).

In the case of aspectual--temporal forms, there are frequent parallels, such as Ḥassāniyya \textit{mā} \textit{tla} and Zenaga \textit{war} \textit{yiššiy} ‘no longer’, Ḥassāniyya \textit{ma-zāl} and Zenaga \textit{yaššiy} ‘still’, Ḥassāniyya \textit{tamm} and Zenaga \textit{yuktay} ‘to continue to’, Ḥassāniyya \textit{ʕgab} and Zenaga \textit{yaggara} ‘to end up doing’. One of the most notable parallel innovations, however, concerns the future morpheme: Ḥassāniyya \textit{lāhi} (invariable participle of an otherwise obsolete verb, but compare \textit{ltha} ‘to pass one’s time’) and Zenaga \textit{yanhāya} (a conjugated verb also meaning ‘to busy oneself with something’, in addition to its future function). In both cases we have forms related to Classical Arabic \textit{lahā} ‘to amuse oneself’, with the Zenaga form apparently being a borrowing. It seems, therefore, that this borrowing preceded the \textit{lāhi} of Ḥassāniyya and likely then influenced its adoption as a future tense marker. Note also that in the Arabic dialect of the Jews of Algiers, \textit{lāti} is a durative present tense marker (\citealt{Cohen1924}: 221; \citealt{Taine-Cheikh2004}: 224; \citealt{Taine-Cheikh2008chapter}: 126–127; \citealt{Taine-Cheikh2009}: 99).

Ḥassāniyya and Zenaga also display common features with regard to complex phrases. For example, concerning completives, Zenaga differs from other Berber languages in its highly developed usage of \textit{ad} \~{} \textit{að}, and in particular in the grammaticalized usage of this demonstrative as a quotative particle after verbs of speaking and thinking \citep{Taine-Cheikh2010Zenaga}. This may have had an influence on the usage of the conjunctions \textit{an(n)-} and \textit{ʕan-} (the two forms tend to be confused) in the same function in Ḥassāniyya.

Finally, regarding the variable appearance of a resumptive pronoun in Ḥassāniyya object relative clauses, if influence from Berber (where a resumptive pronoun is always absent) has played any role here, it has simply been to reinforce a construction already attested in the earliest Arabic, whereby the resumptive pronoun is absent if the antecedent is definite, as in ‎(1).

\ea
\gll n{\R}ədd ʕlī-kum ə{\R}-{\R}wāye lli {\R}add-Ø ʕlī-ya muḥammad\\
 tell.\textsc{impf.}1\textsc{sg} on-2\textsc{pl} \textsc{def-}story \textsc{rel} tell.\textsc{prf.3sg.m-}Ø on-\textsc{obl.1sg} Mohammed\\
\glt ‘I am going to tell you the story that Mohammed told me.’ 
\z

\subsubsection{Regional influence of Maghrebi Arabic} %3.3.2. /

The Ḥassāniyya spoken in the south of Morocco is rather heavily influenced by other Arabic varieties spoken in the region. Even among those who conserve virtually all the characteristic features of Ḥassāniyya (preservation of interdentals, synthetic genitive construction, absence of the pre-verbal particle \textit{kā-} or \textit{tā-}, absence of discontinuous negation, absence of the indefinite article), particular features of the Moroccan Arabic koiné appear either occasionally or regularly among certain speakers. The most common such features are perhaps the genitive particle \textit{dyal} (\citealt{Taine-Cheikh1997socio}: 98) and the preverbal particle \textit{kā} (\citealt{Aguadé1998}: 211, §37; 213, §42).

In the Ḥassāniyya of Mali, usage of a genitive particle remains marginal, although Heath (\citeyear[162]{Heath2004}) highlights a few uses of genitive \textit{(n)tāʕ} in his texts.

\subsection{Lexicon} %3.4. /

\subsubsection{Confirmed loanwords} %3.4.1. /

\subsubsubsection{Loanwords from Standard Arabic} %3.4.1.1. /

Verbs loaned from Standard Arabic are as common as nominal and adjectival loans. Whatever their category, loans are often distinctive in some way (whether because of their syllabic structure, the presence of particular phonemes or their morphological template), since the lexeme usually (though not always) has the same form in both the recipient language and the source language. Examples of loans without any distinctive features are \textit{ba{\R}{\R}a{\R}} ‘to justify’, and \textit{ðahbi} ‘golden’. 

A certain number of Standard Arabic verbs with the infix \textit{{}-}\textit{t-} or the prefix \textit{sta-} are borrowed, but these verbal patterns can be found elsewhere in Ḥassāniyya. 

Certain lexical fields exhibit a particularly high degree of loans from Standard Arabic: anything connected with Islamic studies or abstract concepts (religion, rights, morality, feelings, etc.) and, more recently, politics, media and modern material culture. These regularly retain the meaning (or one of the meanings) of the source-language item. 

\subsubsubsection{Loanwords from Berber} %3.4.1.2. /

There are many lexical items that are probable loans from Berber, with a number of certain cases among them. 

Here we may point to several non-Arabic-origin verbs with cognates across a wide range of Berber languages, such as \textit{k{\R}aṭ} ‘to scrape off’ (Zenaga \textit{yug{\R}að̣}); \textit{šayð̣að̣} ‘to make a lactating camel adopt an orphaned calf from another mother’ (Zenaga \textit{yaṣṣuð̣að̣} ‘to breastfeed’, \textit{yuḍḍað̣} ‘to suckle’); \textit{santa} ‘to begin’ (Zenaga \textit{yassanta} ‘to begin’, Tuareg \textit{ent} ‘to be started, to begin’); \textit{gaymar} ‘to hunt from a distance’ (Berber \textit{gmər} ‘to hunt'). 

Other verbs are derived from nouns loaned from Berber. Hence, \textit{ɣawba} ‘to restrain a camel, put it in an \textit{aɣāba}’ (Tuareg \textit{aɣaba} ‘jaws’). Sometimes there is both a verb and an adjective stemming from a loaned root, as in \textit{gaylal} ‘to have the tail cut’ and \textit{agīlāl} ‘having a cut tail’ (Tuareg \textit{gilel} and \textit{agilal}). 

Some loaned Ḥassāniyya nouns are found with the same root (or an equivalent root) in Berber languages other than Zenaga. For example: \textit{agayš} ‘male bustard’ (Tuareg \textit{gayəs}); \textit{āškə{\R}} ‘partridge’ (Kabyle \textit{tasekkurt} in the feminine form); \textit{tayffārət} ‘fetlock (camel)’ (Zenaga \textit{tiʔffart}, Tuareg \textit{téffart}); \textit{azāɣər} ‘wooden mat ceiling between beams’ (Zenaga \textit{azaɣri} ‘lintel, beam (of a well)’, Tuareg \textit{ǝ\kern -0.5ptzgər} ‘to cross’, \textit{ăzəgər} ‘crossbeam’); \textit{talawmāyət} ‘dew’ (Zenaga \textit{tayaṃut}, Tuareg \textit{tălămut}); \textit{(n)tūrža} ‘\textit{Calotropis} \textit{procera}’ (Zenaga \textit{turžah}, Tuareg \textit{tərza}).

Most of the loanwords cited above are attested in Zenaga (sometimes in a more innovative form than is found in other Berber varieties, such as \textit{yaggīyyay} ‘to have a cut tail’ where /y/ < *l). However, there are numerous cases where a corresponding Berber item is attested only in Zenaga. In such cases it is difficult to precisely identify the source language, even if the phonology and/or morphology seems to indicate a non-Arabic origin. 

Loanwords from Berber seem to be particularly common in the lexicon of fauna, flora, and diseases, as well as in the field of traditional material culture (objects, culinary traditions, farming practices, etc.; \citealt{Taine-Cheikh2010lexiques,Taine-Cheikh2014}). Unlike the form of the loans, which is often quite divergent from that of the source items, their semantics tends to remain largely unchanged. However, there are some exceptions, notably when the verbs have a general meaning in Berber (cf. above `to breastfeed' vs. `to make a lactating camel adopt an orphaned calf from another mother'). 

\subsubsubsection{Loanwords from Sahel languages} %3.4.1.3. /

Rather few Ḥassāniyya lexical items seem to be borrowed directly from African languages, and the origin of those that are is rarely known precisely. We may note, however, in addition to \textit{gadʸ} ‘dried fish’ (< Wolof) and \textit{dʸəngra} ‘warehouse’ (< Soninke), a few terms which appear to be borrowed from Pulaar: \textit{ḅamḅa} ‘to carry a child on one’s back’, \textit{tʸəhli} ‘roof on pillars’ and \textit{kīri} ‘boundary between two fields’.

In some regions we find a concentration of loans in particular domains in relation to specific contact languages. For example, in the ancient town of Tichitt, we find borrowings from Azer and Soninke (\citealt{Jacques-Meunié1961}; \citealt{Monteil1939}; \citealt{Diagana2013}): \textit{kā} ‘house’ (Azer \textit{ka(ny)}, Soninke \textit{ká}) in \textit{kā} \textit{n} \textit{laqqe} ‘entrance of the house’; \textit{killen} ‘path’ (Azer \textit{kille}, Soninke \textit{kìllé}); \textit{kunyu} \~{} \textit{kenyen} ‘cooking’ (Azer \textit{knu} \~{} \textit{kenyu}, Soninke \textit{kìnŋú}).

A significant list of loanwords from Songhay has been compiled by \citet{Heath2004} in Mali, including e.g.: \textit{ṣawṣab} (< \textit{sosom} \~{} \textit{sosob}) ‘pound (millet) in mortar to remove bran from grains’; \textit{daydi} \~{} \textit{dayday} (< \textit{deydey}) ‘daily grocery purchase’; \textit{ākā{\R}āy} (< \textit{kaarey}) ‘crocodile’; \textit{sari} (< \textit{seri}) ‘millet porridge’. Only \textit{sari} has been recorded elsewhere in Mauritania (in the eastern town of Walata). On the other hand, all authors who have done field work on the Ḥassāniyya of Mali (particularly in the region of Timbuktu and the Azawad), have noted loanwords from Songhay. This is true also of \citet{Clauzel1960} who, as well as a number of Berber loanwords, gives a small list of Songhay-derived items used in the salt mine of Tāwdenni, such as \textit{titi} ‘cylinder of saliferous clay used as a seat by the miners’ (< \textit{tita}) and \textit{tʸar} ‘adze’ (< \textit{tʸara}).

\subsubsubsection{Loanwords from Indo-European languages} %3.4.1.4. /

The use of loanwords from European languages tends to vary over time. Thus, a large proportion of the French loanwords borrowed during the colonial period have more recently gone out of use, such as \textit{bə{\R}ṭmāla} or \textit{qo{\R}ṭmāl} ‘wallet’ (< \textit{porte-monnaie}), \textit{dabbīš} ‘telegram’ (< \textit{dépêche} ‘dispatch’) or \textit{ṣa{\R}waṣ} ‘to be very close to the colonizers’ (\textit{<} \textit{service} ‘service’). This is true not only of items referring to obsolete concepts (such as the currency terms \textit{sūvāya} ‘sou’ or \textit{ftən/vəvtən} ‘cent’, likely < \textit{fifteen}), but also of those referring to still-current concepts which are, however, now referred to with a term drawn from Standard Arabic (e.g. \textit{minəstr} ‘minister’, replaced by \textit{wazīr}). This does not, however, eliminate the permanence of some old loanwords such as \textit{wata} ‘car’ (< \textit{voiture}) or \textit{ma{\R}ṣa} ‘market’ (< \textit{marché}).\footnote{\citet{OuldMohamedBaba2003} gives an extensive list of loanwords from French and offers a classification by semantic field.}

Although not unique to Ḥassāniyya, the frequency of the emphatic phonemes (especially /ṣ/ and /ṭ/) in loans from European languages is notable. Consider, in addition to the treatment of \textit{service}, \textit{porte-monnaie} and \textit{marché} as noted above, that of \textit{baṭ{\R}ūn} ‘boss’ (< \textit{patron}), which gives rise to \textit{tbaṭ{\R}an} ‘to be(come) a boss’ \textit{ṭawn} ‘ton’ (< \textit{tonne}).

\subsubsection{More complex cases} %3.4.2. /

\subsubsubsection{\textit{Wanderwörter}} \label{wander}

Various Arabic lexical items derive from Latin, Armenian, Turkish, Persian, and so on. In the case of, for example, the names of calendar months, or of items such as trousers (\textit{sərwāl)}, these terms are not borrowed directly from the source language by Ḥassāniyya and are found elsewhere (e.g. \textit{balbūẓa} ‘eyeball’ < Latin \textit{bulbus}, attested throughout the Maghreb). The history of such items will not be dealt with here. We can, however, mention the case of some well-attested terms in Ḥassāniyya that appear to have been borrowed from sub-Saharan Africa.

One such is \textit{mā{\R}u} ‘rice’, which seems to come from Soninke (\textit{máarò}), although it is also attested in Wolof (\textit{maalo}) and Zenaga (\textit{mārih}). Another term, which is just as emblematic, is \textit{mbū{\R}u} ‘bread’, whose origin has variously been attributed to Wolof, Azer, Mandigo, and even English \textit{bread}.

To these very everyday terms, we may also add \textit{ṃutri} ‘pearl millet’ and \textit{makka} ‘maize’, which have the same form both in Ḥassāniyya and in Zenaga. The first is a loanword from Pulaar (\textit{muutiri}). The second is attested in many languages and seems to have come from the placename Mecca. 

As for \textit{garta} ‘peanut’, \textit{ḷāḷo} \~{} \textit{ḷaḷu} ‘pounded baobab leaves that serve as a condiment’ (synonym of \textit{taqya} in the southwest of Mauritania) and \textit{kəddu} ‘spoon’, these appear to be used just as frequently in Pulaar as they are in Wolof. 

\subsubsubsection{Berberized items} %3.4.2.2. /

Despite the absence of any Berber affixes in the loanwords listed in §\ref{wander}, only \textit{kəddu} ‘spoon’ is regularly used with the definite article. In this regard, these loanwords act like words borrowed from Berber, or more generally, those with Berber affixes. 

It is, in fact, difficult to prove that a noun with this kind of affix is definitely of Berber origin, since we find nouns of various origins with Berber affixes. Some of them are loanwords from the languages of the sedentary people of the valley, such as \textit{adabāy} ‘village of former sedentary slaves (\textit{ḥ{\R}āṭīn})’ (< Soninke \textit{dèbé} ‘village’); \textit{iggīw} \~{} \textit{īggīw} ‘griot’ (Zenaga \textit{iggiwi}, borrowed from Wolof \textit{geewel} or from Pulaar \textit{gawlo}). Others are borrowed from French: \textit{agā{\R}āž} ‘garage’; \textit{təmbīskit} ‘biscuit’. Even terms of Arabic origin are Berberized, as is likely the case with with \textit{tasūvra} ‘large decorated leather bag for travelling’ (cf. \textit{sāvər} ‘to travel’) or \textit{tāẓəẓmīt} ‘asthma’ (cf. CA \textit{zaǧma} ‘shortness of breath when giving birth’).

\subsubsubsection{Reborrowings} %3.4.2.3. /

Instances of back and forth between two languages – primarily Ḥassāniyya and Zenaga – seem to be the reason for another type of mixed form, illustrated previously in §\ref{sa} by the Zenaga verbs \textit{yassəðbah} ‘to leave in the afternoon’ and \textit{yišnar} ‘to orient oneself’.

Ḥassāniyya \textit{saɣnan} ‘to mix gum with water to make ink’ provides another example, where this time the points of departure and arrival seem to be from the Arabic side. In fact, this loanword is a borrowing of Zenaga \textit{yassuɣnan} ‘to thicken (ink) by adding gum’, a verb formed from \textit{əssaɣan} ‘gum’. This noun in turn appears to be an adaptation of the Arabic \textit{samɣa} ‘ink’. 

In the case of \textit{sla} ‘placenta’, there is a double round-trip between the two languages, this time without metathesis: after a passage from Arabic to Zenaga (> \textit{əs(s)la}), there is return to Ḥassāniyya with the causative verb \textit{sasla} ‘to soak a hide’, and a second loan into Zenaga with the reflexive form \textit{(yə)stasla} ‘to start to lose fur (of soaked hides)’.

\subsubsubsection{Calques} %3.4.2.4. /

Calques are undoubtedly common, but they are particularly frequent in locutions such as \textit{rəggət} \textit{əž-žəll} ‘susceptibility’ and \textit{bū-damʕa} ‘rinderpest’ (literally ‘thinness of skin’ and ‘the one with a tear’). These are exact calques of their Zenaga equivalents \textit{taššəddi-n} \textit{əyim} and \textit{ən-anḍi} (\citealt{Taine-Cheikh2008chapter}).

\subsubsubsection{Individual variation} %3.4.2.5. /

Receptivity to loanwords differs from one individual to another. This is natural when we are dealing with bilingual speakers and this probably explains the special features of the Ḥassāniyya of the Awlād Banʸūg (often bilingual speakers of Ḥassāniyya and Wolof) or the Ḥassāniyya of Mali (where Arabic speakers often speak Songhay and sometimes Tamasheq). However, it also depends on the individuals in question in terms of what we might call their “loyalty” to the language, whether the language is under pressure from Moroccan Arabic koiné in Morocco (\citealt{Taine-Cheikh1997socio}; \citealt{Heath2002}; \citealt{Paciotti2017}), or whether it is imposed as a lingua franca in Mauritania \citep{Dia2007}.

\section{Conclusion} %4. /

The principal domain affected by contact in Ḥassāniyya is that of the lexicon (though an assessment in percentage terms is not at present possible). However, the integration of loanwords – in particular those from Standard Arabic and Berber – has resulted in a significant enrichment of the phonological system and of the inventory of nominal patterns. The effects of contact on the verbal morphology and syntax of the dialect are more indirect. The major developments in Ḥassāniyya seem most likely to instead be a product of internal evolution. In certain cases, Zenaga has probably had an influence; in others, we rather witness instances of parallel evolution. 

In future, by studying the vehicular Ḥassāniyya of Mauritania and of the border regions (southern Morocco, southern Algeria, Senegal, Niger, and so on) we will perhaps discover new developments as a result of contacts triggered by the political and societal changes of the twenty-first century. 

\section*{Further reading}

Links between Ḥassāniyya and other languages are particularly complex at the level of semantics and lexicon. On these topics, beyond the available Ḥassāniyya and Zenaga dictionaries (\citealt{Heath2004}; \citealt{Taine-Cheikh1988dictionary,Taine-Cheikh2008dictionary}), readers may consult the available studies of specific fields (\citealt{Monteil1952}; \citealt{Taine-Cheikh2013}) or particular templates (\citealt{Taine-Cheikh2018quadri}).

\section*{Abbreviations}
\begin{tabularx}{.5\textwidth}{@{}lQ@{}}
\textsc{1, 2, 3} & 1st, 2nd, 3rd person \\
CA           & Classical Arabic \\
\textsc{def}     & definite article \\
Hass.           & Ḥassāniyya Arabic \\
\textsc{impf}   & imperfect \\
\end{tabularx}%
\begin{tabularx}{.5\textwidth}{@{}lQ@{}}
\textsc{m}   & masculine\\
\textsc{pl}  & plural\\
\textsc{rel}     & relativizer\\
\textsc{sg}  & singular \\
Zen. & Zenaga
\end{tabularx}%
 

\sloppy
\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
